---
title: "Metabolite panel associations"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
if (!requireNamespace("limma", quietly = TRUE)) {
  stop("Package 'limma' is required for the metabolite panel recipe.")
}
analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
assay_matrix <- analysis_data$assay
recipe_parameters <- config$analyses$recipe_parameters[["metabolomics.linear_feature_scan"]] %||% list()
exposure <- config$variables$grouping
covariates <- intersect(
  unlist(config$variables$covariates %||% character(0), use.names = FALSE),
  names(metadata)
)
min_n <- as.integer(recipe_parameters$min_complete_cases %||% 20)
if (is.null(exposure) || !exposure %in% names(metadata)) {
  stop("The limma panel recipe requires a confirmed exposure/grouping variable.")
}
```

## Analysis contract

The recipe fits a common limma design across standardized metabolite features, uses model-specific available outcome values, and applies Benjamini–Hochberg correction across the complete feature–coefficient result family.

```{r}
contract <- data.frame(
  field = c("Samples", "Features", "Exposure", "Covariates", "Minimum observations"),
  value = c(
    ncol(assay_matrix),
    nrow(assay_matrix),
    exposure,
    if (length(covariates)) paste(covariates, collapse = ", ") else "None",
    min_n
  )
)
knitr::kable(contract)
```

## Feature-level limma models

```{r}
design_variables <- unique(c(exposure, covariates))
complete_design <- stats::complete.cases(metadata[, design_variables, drop = FALSE])
model_metadata <- metadata[complete_design, , drop = FALSE]
model_assay <- assay_matrix[, complete_design, drop = FALSE]
if (nrow(model_metadata) < min_n) {
  stop("Too few complete design rows for the configured minimum: ", nrow(model_metadata))
}

design_formula <- stats::reformulate(design_variables)
design <- stats::model.matrix(design_formula, data = model_metadata)
target_terms <- grep(paste0("^", make.names(exposure)), make.names(colnames(design)), value = FALSE)
target_terms <- setdiff(target_terms, which(colnames(design) == "(Intercept)"))
if (!length(target_terms)) stop("No exposure coefficient was estimable in the limma design.")

fit <- limma::lmFit(model_assay, design)
fit <- limma::eBayes(fit, trend = TRUE)
feature_n <- rowSums(is.finite(model_assay))

term_results <- lapply(target_terms, function(term_index) {
  table <- limma::topTable(
    fit,
    coef = term_index,
    number = Inf,
    sort.by = "none",
    confint = TRUE
  )
  data.frame(
    feature = rownames(table),
    exposure = exposure,
    term = colnames(design)[[term_index]],
    n = feature_n[rownames(table)],
    estimate = table$logFC,
    std.error = fit$stdev.unscaled[, term_index] * fit$sigma,
    conf.low = table$CI.L,
    conf.high = table$CI.R,
    p.value = table$P.Value,
    stringsAsFactors = FALSE
  )
})

panel_results <- do.call(rbind, term_results)
panel_results <- panel_results[panel_results$n >= min_n, , drop = FALSE]
panel_results$q.value <- stats::p.adjust(
  panel_results$p.value,
  method = recipe_parameters$fdr_method %||% "BH"
)
panel_results <- panel_results[order(panel_results$q.value, panel_results$p.value), ]
model_status <- data.frame(
  feature = rownames(model_assay),
  status = ifelse(feature_n >= min_n, "completed", "skipped"),
  n = feature_n,
  note = ifelse(feature_n >= min_n, "ok", "too_few_observed_outcomes"),
  stringsAsFactors = FALSE
)

dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
utils::write.csv(panel_results, "../output/results/metabolite_panel.csv", row.names = FALSE)
utils::write.csv(model_status, "../output/results/model_status.csv", row.names = FALSE)
knitr::kable(utils::head(panel_results, 25), digits = 4)
```

## Effect overview

```{r}
#| fig-width: 9
#| fig-height: 7
if (nrow(panel_results) && requireNamespace("ggplot2", quietly = TRUE)) {
  top_results <- utils::head(panel_results, 20)
  top_results$label <- factor(
    paste(top_results$feature, top_results$term, sep = " · "),
    levels = rev(paste(top_results$feature, top_results$term, sep = " · "))
  )
  print(
    ggplot2::ggplot(top_results, ggplot2::aes(x = estimate, y = label)) +
      ggplot2::geom_vline(xintercept = 0, color = "grey75") +
      ggplot2::geom_errorbarh(ggplot2::aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
      ggplot2::geom_point(ggplot2::aes(color = q.value < 0.05), size = 2) +
      ggplot2::scale_color_manual(values = c("TRUE" = "#0f766e", "FALSE" = "#71717a"), guide = "none") +
      ggplot2::labs(x = "Standardized effect estimate (95% CI)", y = NULL) +
      ggplot2::theme_minimal(base_size = 11)
  )
}
```

## Interpretation boundary

These are feature-wise associations rather than causal effects. Factor coefficients are interpreted relative to the reference level encoded in the model matrix.

## Reproducibility

```{r}
sessionInfo()
```
