---
title: "Longitudinal metabolite models"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
if (!requireNamespace("lmerTest", quietly = TRUE)) {
  stop("Package 'lmerTest' is required for the longitudinal metabolomics recipe.")
}
analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
assay_matrix <- analysis_data$assay
recipe_parameters <- config$analyses$recipe_parameters[["metabolomics.repeated_measures_mixed_model"]] %||% list()
exposure <- config$variables$grouping
subject_id <- config$identifiers$subject_id
visit <- config$identifiers$visit
covariates <- intersect(
  unlist(config$variables$covariates %||% character(0), use.names = FALSE),
  names(metadata)
)
min_n <- as.integer(recipe_parameters$min_complete_cases %||% 30)
identifier_config <- list(exposure = exposure, subject_id = subject_id, visit = visit)
missing_identifiers <- names(identifier_config)[
  vapply(identifier_config, function(value) is.null(value) || !value %in% names(metadata), logical(1))
]
if (length(missing_identifiers)) {
  stop("Longitudinal models require configured exposure, subject, and visit columns.")
}
```

## Model specification

Each standardized metabolite is modeled with exposure, visit, their interaction, configured covariates, and a participant-level random intercept. Every feature receives an explicit completed, skipped, or failed status.

```{r}
specification <- data.frame(
  field = c("Exposure", "Visit", "Subject ID", "Covariates", "Minimum observations"),
  value = c(
    exposure,
    visit,
    subject_id,
    if (length(covariates)) paste(covariates, collapse = ", ") else "None",
    min_n
  )
)
knitr::kable(specification)
```

## Mixed-model results

```{r}
fit_feature_lmm <- function(feature) {
  model_data <- metadata
  model_data$.outcome <- as.numeric(assay_matrix[feature, ])
  required <- unique(c(".outcome", exposure, visit, subject_id, covariates))
  complete <- stats::complete.cases(model_data[, required, drop = FALSE])
  model_data <- model_data[complete, required, drop = FALSE]

  if (nrow(model_data) < min_n) {
    return(list(
      result = NULL,
      status = data.frame(feature = feature, status = "skipped", n = nrow(model_data), note = "too_few_complete_cases")
    ))
  }
  if (length(unique(model_data[[subject_id]])) < 3 || !anyDuplicated(model_data[[subject_id]])) {
    return(list(
      result = NULL,
      status = data.frame(feature = feature, status = "skipped", n = nrow(model_data), note = "insufficient_repeated_subjects")
    ))
  }

  fixed_terms <- paste0("`", exposure, "` * `", visit, "`")
  if (length(covariates)) {
    fixed_terms <- paste(fixed_terms, paste(sprintf("`%s`", covariates), collapse = " + "), sep = " + ")
  }
  formula <- stats::as.formula(
    paste0(".outcome ~ ", fixed_terms, " + (1 | `", subject_id, "`)")
  )
  model <- try(
    lmerTest::lmer(formula, data = model_data, REML = FALSE),
    silent = TRUE
  )
  if (inherits(model, "try-error")) {
    return(list(
      result = NULL,
      status = data.frame(feature = feature, status = "failed", n = nrow(model_data), note = as.character(model))
    ))
  }

  coefficients <- summary(model)$coefficients
  keep_terms <- setdiff(rownames(coefficients), "(Intercept)")
  intervals <- suppressMessages(stats::confint(model, parm = keep_terms, method = "Wald"))
  result <- data.frame(
    feature = feature,
    exposure = exposure,
    term = keep_terms,
    n = nrow(model_data),
    subjects = length(unique(model_data[[subject_id]])),
    estimate = coefficients[keep_terms, "Estimate"],
    std.error = coefficients[keep_terms, "Std. Error"],
    conf.low = intervals[keep_terms, 1],
    conf.high = intervals[keep_terms, 2],
    p.value = coefficients[keep_terms, "Pr(>|t|)"],
    stringsAsFactors = FALSE
  )
  list(
    result = result,
    status = data.frame(
      feature = feature,
      status = if (lme4::isSingular(model)) "warning" else "completed",
      n = nrow(model_data),
      note = if (lme4::isSingular(model)) "singular_fit" else "ok"
    )
  )
}

model_runs <- lapply(rownames(assay_matrix), fit_feature_lmm)
result_parts <- lapply(model_runs, `[[`, "result")
result_parts <- result_parts[!vapply(result_parts, is.null, logical(1))]
lmm_results <- if (length(result_parts)) do.call(rbind, result_parts) else data.frame()
lmm_status <- do.call(rbind, lapply(model_runs, `[[`, "status"))
if (nrow(lmm_results)) {
  lmm_results$q.value <- stats::p.adjust(
    lmm_results$p.value,
    method = recipe_parameters$fdr_method %||% "BH"
  )
  lmm_results <- lmm_results[order(lmm_results$q.value, lmm_results$p.value), ]
}

dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
utils::write.csv(lmm_results, "../output/results/metabolite_lmm.csv", row.names = FALSE)
utils::write.csv(lmm_status, "../output/results/lmm_model_status.csv", row.names = FALSE)
if (nrow(lmm_results)) {
  knitr::kable(utils::head(lmm_results, 30), digits = 4)
} else {
  cat("No longitudinal model completed. See `lmm_model_status.csv`.")
}
```

## Interpretation boundary

The exposure-by-visit interaction tests whether longitudinal metabolite patterns differ by exposure group. Singular fits are retained as warnings and should not be treated as stable evidence.

## Reproducibility

```{r}
sessionInfo()
```
