---
title: "Metabolite panel associations"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
assay_matrix <- analysis_data$assay
exposure <- config$variables$grouping
covariates <- intersect(
  unlist(config$variables$covariates %||% character(0), use.names = FALSE),
  names(metadata)
)
min_n <- as.integer(config$analyses$min_complete_cases %||% 20)
```

## Analysis contract

This deterministic recipe fits one linear model per standardized metabolite feature. It uses model-specific complete cases and applies Benjamini–Hochberg correction across the feature-level family. Every attempted feature receives a model-status row.

```{r}
contract <- data.frame(
  field = c("Samples", "Metabolite features", "Exposure", "Covariates", "Minimum complete cases"),
  value = c(
    length(analysis_data$sample_ids),
    length(analysis_data$feature_ids),
    exposure %||% "Not configured",
    if (length(covariates)) paste(covariates, collapse = ", ") else "None",
    min_n
  )
)
knitr::kable(contract)
```

## Feature-level models

```{r}
fit_feature <- function(feature) {
  outcome <- as.numeric(assay_matrix[feature, ])
  model_data <- metadata
  model_data$.outcome <- outcome
  required <- unique(c(".outcome", exposure, covariates))

  if (is.null(exposure) || !exposure %in% names(model_data)) {
    return(list(
      results = NULL,
      status = data.frame(feature = feature, status = "skipped", n = 0L, note = "exposure_missing")
    ))
  }

  complete <- stats::complete.cases(model_data[, required, drop = FALSE])
  n_complete <- sum(complete)
  if (n_complete < min_n) {
    return(list(
      results = NULL,
      status = data.frame(feature = feature, status = "skipped", n = n_complete, note = "too_few_complete_cases")
    ))
  }

  analysis_frame <- model_data[complete, required, drop = FALSE]
  if (length(unique(analysis_frame[[exposure]])) < 2) {
    return(list(
      results = NULL,
      status = data.frame(feature = feature, status = "skipped", n = n_complete, note = "exposure_has_fewer_than_two_values")
    ))
  }

  quoted_terms <- paste(sprintf("`%s`", c(exposure, covariates)), collapse = " + ")
  model <- try(stats::lm(stats::as.formula(paste(".outcome ~", quoted_terms)), data = analysis_frame), silent = TRUE)
  if (inherits(model, "try-error")) {
    return(list(
      results = NULL,
      status = data.frame(feature = feature, status = "failed", n = n_complete, note = as.character(model))
    ))
  }

  coefficients <- summary(model)$coefficients
  keep_terms <- setdiff(rownames(coefficients), "(Intercept)")
  if (!length(keep_terms)) {
    return(list(
      results = NULL,
      status = data.frame(feature = feature, status = "skipped", n = n_complete, note = "no_estimable_exposure_term")
    ))
  }
  intervals <- suppressMessages(stats::confint(model))
  rows <- data.frame(
    feature = feature,
    exposure = exposure,
    term = keep_terms,
    n = n_complete,
    estimate = coefficients[keep_terms, "Estimate"],
    std.error = coefficients[keep_terms, "Std. Error"],
    conf.low = intervals[keep_terms, 1],
    conf.high = intervals[keep_terms, 2],
    p.value = coefficients[keep_terms, "Pr(>|t|)"],
    stringsAsFactors = FALSE
  )
  list(
    results = rows,
    status = data.frame(feature = feature, status = "completed", n = n_complete, note = "ok")
  )
}

model_runs <- lapply(rownames(assay_matrix), fit_feature)
result_parts <- lapply(model_runs, `[[`, "results")
result_parts <- result_parts[!vapply(result_parts, is.null, logical(1))]
panel_results <- if (length(result_parts)) do.call(rbind, result_parts) else data.frame()
model_status <- do.call(rbind, lapply(model_runs, `[[`, "status"))

if (nrow(panel_results)) {
  panel_results$q.value <- stats::p.adjust(panel_results$p.value, method = "BH")
  panel_results <- panel_results[order(panel_results$q.value, panel_results$p.value), ]
}
dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
utils::write.csv(panel_results, "../output/results/metabolite_panel.csv", row.names = FALSE)
utils::write.csv(model_status, "../output/results/model_status.csv", row.names = FALSE)

if (nrow(panel_results)) {
  knitr::kable(utils::head(panel_results, 25), digits = 4)
} else {
  cat("No feature-level model completed. See `model_status.csv` for the explicit reason for each feature.")
}
```

## Effect overview

```{r}
#| fig-width: 9
#| fig-height: 7
if (nrow(panel_results) && requireNamespace("ggplot2", quietly = TRUE)) {
  top_results <- utils::head(panel_results, 20)
  top_results$label <- paste(top_results$feature, top_results$term, sep = " · ")
  top_results$label <- factor(top_results$label, levels = rev(top_results$label))
  print(
    ggplot2::ggplot(top_results, ggplot2::aes(x = estimate, y = label)) +
      ggplot2::geom_vline(xintercept = 0, color = "grey75") +
      ggplot2::geom_errorbarh(ggplot2::aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
      ggplot2::geom_point(ggplot2::aes(color = q.value < 0.05), size = 2) +
      ggplot2::scale_color_manual(values = c("TRUE" = "#0f766e", "FALSE" = "#71717a"), guide = "none") +
      ggplot2::labs(x = "Standardized effect estimate (95% CI)", y = NULL) +
      ggplot2::theme_minimal(base_size = 11)
  )
}
```

## Interpretation boundary

These are feature-wise associations, not causal effects. The correction family includes all successfully fitted metabolite–term rows shown in the result object.

## Reproducibility

```{r}
sessionInfo()
```
