---
title: "Secondary figures: child diet and metabolites"
---

These figures show exposure-level detail, adjustment robustness, and cross-question pattern coherence for the linear-model components (`Q1`-`Q3`). They are not the primary inferential summaries and do not cover the warning-heavy repeated-measures `Q4` analysis.

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(MultiAssayExperiment)
library(SummarizedExperiment)

results <- readRDS("../output/results/targeted_v2/child_diet_study1_results_v2.rds")
q1 <- results$q1
q2 <- results$q2
q3 <- results$q3

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = c("visit_6", "visit_7")
)

mae <- analysis_obj$mae
df_all <- analysis_obj$data
metabolite_cols <- analysis_obj$metabolite_cols
tse6 <- experiments(mae)[["visit_6"]]
tse7 <- experiments(mae)[["visit_7"]]

feature_meta <- as.data.frame(
  SummarizedExperiment::rowData(experiments(mae)[["visit_all"]]),
  stringsAsFactors = FALSE
)
feature_meta$feature <- rownames(feature_meta)
feature_meta$feature_group <- if ("Group" %in% names(feature_meta)) as.character(feature_meta$Group) else "Unknown"
feature_meta$feature_group[is.na(feature_meta$feature_group) | feature_meta$feature_group == ""] <- "Unknown"

add_feature_meta <- function(df) {
  left_join(df, feature_meta[, c("feature", "feature_group")], by = "feature")
}

safe_logp <- function(p) {
  p <- as.numeric(p)
  p[p <= 0 | is.na(p)] <- NA_real_
  -log10(p)
}

label_exposure <- function(x) {
  gsub("_", " ", x, fixed = TRUE)
}

build_corr_heatmap_df <- function(df, feature_cols) {
  feature_cols <- intersect(feature_cols, names(df))
  if (length(feature_cols) < 2) return(NULL)

  mat <- as.matrix(df[, feature_cols, drop = FALSE])
  storage.mode(mat) <- "numeric"
  keep <- colSums(is.finite(mat)) >= 20
  mat <- mat[, keep, drop = FALSE]
  if (ncol(mat) < 2) return(NULL)

  cor_mat <- suppressWarnings(stats::cor(mat, use = "pairwise.complete.obs"))
  cor_mat[!is.finite(cor_mat)] <- 0
  diag(cor_mat) <- 1
  cor_mat <- pmax(pmin(cor_mat, 1), -1)

  ord <- stats::hclust(stats::as.dist(1 - cor_mat), method = "average")$order
  cor_mat <- cor_mat[ord, ord, drop = FALSE]

  expand.grid(
    row = seq_len(nrow(cor_mat)),
    col = seq_len(ncol(cor_mat))
  ) %>%
    mutate(correlation = as.vector(cor_mat))
}

strip_visit_suffix <- function(x) {
  sub("(6|7)$", "", as.character(x))
}

is_numeric_exposure <- function(tse, exposure) {
  cd <- as.data.frame(colData(tse), stringsAsFactors = FALSE)
  exposure %in% names(cd) && is.numeric(cd[[exposure]])
}

pick_representative_numeric_exposure <- function(res_df, tse) {
  if (is.null(res_df) || nrow(res_df) == 0) return(NA_character_)
  cand <- res_df %>%
    distinct(exposure, q.value) %>%
    filter(is.finite(q.value)) %>%
    group_by(exposure) %>%
    summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
    arrange(best_q)
  if (nrow(cand) == 0) return(NA_character_)
  for (exp_i in cand$exposure) {
    if (is_numeric_exposure(tse, exp_i)) return(exp_i)
  }
  NA_character_
}

fit_lm_exposure <- function(df, outcome, exposure, covars = character(0), min_n = 30L) {
  vars <- unique(c(outcome, exposure, covars))
  vars <- vars[vars %in% names(df)]
  if (!(outcome %in% vars) || !(exposure %in% vars)) return(NULL)

  sub_df <- df[, vars, drop = FALSE]
  sub_df <- sub_df[complete.cases(sub_df), , drop = FALSE]
  if (nrow(sub_df) < min_n) return(NULL)

  sub_df[[outcome]] <- suppressWarnings(as.numeric(sub_df[[outcome]]))
  sub_df[[exposure]] <- suppressWarnings(as.numeric(sub_df[[exposure]]))
  if (!is.numeric(sub_df[[exposure]]) || length(unique(sub_df[[exposure]][is.finite(sub_df[[exposure]])])) < 3) return(NULL)
  if (!is.numeric(sub_df[[outcome]]) || all(!is.finite(sub_df[[outcome]]))) return(NULL)

  y_sd <- stats::sd(sub_df[[outcome]], na.rm = TRUE)
  if (!is.finite(y_sd) || y_sd <= 0) return(NULL)
  sub_df[[outcome]] <- (sub_df[[outcome]] - mean(sub_df[[outcome]], na.rm = TRUE)) / y_sd

  covars <- setdiff(intersect(covars, names(sub_df)), c(outcome, exposure))
  rhs <- c(exposure, covars)
  f <- stats::as.formula(paste0("`", outcome, "` ~ ", paste(rhs, collapse = " + ")))
  fit <- tryCatch(stats::lm(f, data = sub_df), error = function(e) NULL)
  if (is.null(fit)) return(NULL)

  coef_tbl <- summary(fit)$coefficients
  term <- if (exposure %in% rownames(coef_tbl)) exposure else grep(paste0("^", exposure, "$"), rownames(coef_tbl), value = TRUE)[1]
  if (!is.character(term) || length(term) == 0 || is.na(term)) return(NULL)

  ci <- tryCatch(stats::confint(fit, parm = term, level = 0.95), error = function(e) c(NA_real_, NA_real_))
  data.frame(
    n = nrow(sub_df),
    estimate = unname(coef_tbl[term, "Estimate"]),
    conf.low = unname(ci[1]),
    conf.high = unname(ci[2]),
    p.value = unname(coef_tbl[term, "Pr(>|t|)"]),
    stringsAsFactors = FALSE
  )
}

as_panel_df <- function(tse) {
  cd <- as.data.frame(colData(tse), stringsAsFactors = FALSE)
  ft <- as.data.frame(t(assay(tse, "mbo")), check.names = FALSE)
  cd$sample_id <- rownames(cd)
  ft$sample_id <- rownames(ft)
  out <- left_join(cd, ft, by = "sample_id")
  out$StudyID <- as.character(out$StudyID)
  out
}

df6 <- as_panel_df(tse6)
df7 <- as_panel_df(tse7)

core_v6 <- c(
  "Intervention", "MPrimipara", "MUniEdu", "MGDMOGTT1OR2Fi",
  "CGender", "CWeightBirth", "Bfdurationm"
)
core_v7 <- c(
  "Intervention", "MPrimipara", "MUniEdu", "MGDMOGTT1OR2Fi",
  "CGender", "CWeightBirth", "Bfdurationm"
)
minimal_v6 <- c("Intervention", "CGender")
minimal_v7 <- c("Intervention", "CGender")
technical_v6 <- c(core_v6, "CHealthy6")
technical_v7 <- c(core_v7, "CHealthy7")

rep_q1 <- pick_representative_numeric_exposure(q1, tse6)
rep_q2 <- pick_representative_numeric_exposure(q2, tse7)
rep_q3 <- pick_representative_numeric_exposure(q3, tse6)

rep_tbl <- data.frame(
  analysis = c(
    "Concurrent: 2-year diet and 2-year metabolites",
    "Concurrent: 5-6-year diet and 5-6-year metabolites",
    "Prospective: 2-year diet and 5-6-year metabolites"
  ),
  selected_exposure = c(rep_q1, rep_q2, rep_q3),
  selection_rule = c(
    "Best q-value among numeric visit-6 exposures",
    "Best q-value among numeric visit-7 exposures",
    "Best q-value among numeric visit-6 exposures"
  ),
  stringsAsFactors = FALSE
)
```

Selected exposures used in this page. These are rule-based appendix exemplars, not hand-picked headline results:

```{r}
#| label: secondary-figures-01
make_table(rep_tbl, caption = "Selected representative exposures for secondary figures")
```

## Figure A1. Global association screen

```{r}
#| label: fig-global-screen
#| fig-cap: "Global association screen (p-value ranking) for selected dietary exposures at 2 and 5-6 years. Points colored by metabolite class."
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 10
screen_df <- bind_rows(
  q1 %>% filter(exposure == rep_q1) %>% mutate(analysis = "Concurrent: 2-year diet and 2-year metabolites"),
  q2 %>% filter(exposure == rep_q2) %>% mutate(analysis = "Concurrent: 5-6-year diet and 5-6-year metabolites"),
  q3 %>% filter(exposure == rep_q3) %>% mutate(analysis = "Prospective: 2-year diet and 5-6-year metabolites")
) %>%
  add_feature_meta() %>%
  group_by(analysis) %>%
  arrange(p.value, .by_group = TRUE) %>%
  mutate(rank = row_number(), logp = safe_logp(p.value), sig = q.value < 0.1) %>%
  ungroup()

screen_df %>%
  ggplot(aes(x = rank, y = logp, color = feature_group)) +
  geom_point(alpha = 0.7, size = 1.4) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey45") +
  facet_wrap(~ analysis, scales = "free_x") +
  labs(
    x = "Ordered metabolite rank",
    y = "-log10(p)",
    color = "Metabolite class",
    title = "Global association screens for selected exposures"
  ) +
  theme_minimal(base_size = 16)
```

## Figure A2. Main forest plot of top hits

```{r}
#| label: fig-forest-main
#| fig-cap: "Forest plots showing standardized beta coefficients and 95% confidence intervals for top metabolic associations."
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 12
forest_primary <- bind_rows(
  q1 %>% filter(exposure == rep_q1) %>% mutate(analysis = "Concurrent: 2-year diet and 2-year metabolites"),
  q2 %>% filter(exposure == rep_q2) %>% mutate(analysis = "Concurrent: 5-6-year diet and 5-6-year metabolites"),
  q3 %>% filter(exposure == rep_q3) %>% mutate(analysis = "Prospective: 2-year diet and 5-6-year metabolites")
) %>%
  add_feature_meta() %>%
  group_by(analysis) %>%
  slice_min(q.value, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    feature_lab = paste0(feature_group, " | ", feature)
  )

analysis_order <- c(
  "Concurrent: 2-year diet and 2-year metabolites",
  "Concurrent: 5-6-year diet and 5-6-year metabolites",
  "Prospective: 2-year diet and 5-6-year metabolites"
)

for (analysis_i in analysis_order) {
  panel_df <- forest_primary %>% filter(analysis == analysis_i)
  exp_name <- unique(panel_df$exposure)
  exp_label <- if (length(exp_name) == 1) pretty_exposure(exp_name) else paste(pretty_exposure(exp_name), collapse = ", ")
  p <- panel_df %>%
    ggplot(aes(x = estimate, y = fct_reorder(feature_lab, estimate), color = feature_group)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey45") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "grey55") +
    geom_point(size = 1.9) +
    labs(
      x = "Standardized beta (95% CI)",
      y = NULL,
      color = "Metabolite class",
      title = paste0("Main forest plot: ", analysis_i),
      subtitle = paste0("Selected diet exposure: ", exp_label, " (top 15 most significant associations shown)")
    ) +
    theme_minimal(base_size = 16)
  print(p)
}
```

## Figure A3. Sensitivity and model comparison

```{r}
#| label: fig-sensitivity
#| fig-cap: "Sensitivity analysis comparing coefficient estimates across minimal, primary, and technical adjustment sets."
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 11
top_features_q1 <- q1 %>%
  filter(exposure == rep_q1) %>%
  slice_min(q.value, n = 10, with_ties = FALSE) %>%
  pull(feature)

top_features_q2 <- q2 %>%
  filter(exposure == rep_q2) %>%
  slice_min(q.value, n = 10, with_ties = FALSE) %>%
  pull(feature)

top_features_q3 <- q3 %>%
  filter(exposure == rep_q3) %>%
  slice_min(q.value, n = 10, with_ties = FALSE) %>%
  pull(feature)

fit_cross_models <- function(df, exposure, features, model_sets, analysis_name) {
  bind_rows(lapply(features, function(ft) {
    bind_rows(lapply(names(model_sets), function(model_nm) {
      fit <- fit_lm_exposure(
        df = df,
        outcome = ft,
        exposure = exposure,
        covars = model_sets[[model_nm]]
      )
      if (is.null(fit)) return(NULL)
      fit %>%
        mutate(
          analysis = analysis_name,
          model = model_nm,
          feature = ft
        )
    }))
  }))
}

comp_q1 <- fit_cross_models(
  df = df6,
  exposure = rep_q1,
  features = top_features_q1,
  model_sets = list(
    "Minimally adjusted" = minimal_v6,
    "Primary adjusted" = core_v6,
    "Technical/acute-state adjusted" = technical_v6
  ),
  analysis_name = "Concurrent: 2-year diet and 2-year metabolites"
)

comp_q2 <- fit_cross_models(
  df = df7,
  exposure = rep_q2,
  features = top_features_q2,
  model_sets = list(
    "Minimally adjusted" = minimal_v7,
    "Primary adjusted" = core_v7,
    "Technical/acute-state adjusted" = technical_v7
  ),
  analysis_name = "Concurrent: 5-6-year diet and 5-6-year metabolites"
)

fit_prospective_models <- function(df6, df7, exposure, features, core_covars_v6) {
  bind_rows(lapply(features, function(ft) {
    if (!(ft %in% names(df6)) || !(ft %in% names(df7))) return(NULL)

    dat <- inner_join(
      df7 %>% select(StudyID, outcome = all_of(ft)),
      df6 %>% select(StudyID, exposure_value = all_of(exposure), all_of(core_covars_v6), baseline = all_of(ft)),
      by = "StudyID"
    )

    base_covars <- intersect(core_covars_v6, names(dat))
    bind_rows(
      fit_lm_exposure(dat, outcome = "outcome", exposure = "exposure_value", covars = character(0)) %>%
        mutate(model = "Unadjusted"),
      fit_lm_exposure(dat, outcome = "outcome", exposure = "exposure_value", covars = base_covars) %>%
        mutate(model = "Primary adjusted"),
      fit_lm_exposure(dat, outcome = "outcome", exposure = "exposure_value", covars = c(base_covars, "baseline")) %>%
        mutate(model = "Baseline-metabolite adjusted")
    ) %>%
      mutate(feature = ft)
  }))
}

comp_q3 <- fit_prospective_models(
  df6 = df6,
  df7 = df7,
  exposure = rep_q3,
  features = top_features_q3,
  core_covars_v6 = core_v6
) %>%
  mutate(analysis = "Prospective: 2-year diet and 5-6-year metabolites")

comp_all <- bind_rows(comp_q1, comp_q2, comp_q3) %>%
  add_feature_meta() %>%
  mutate(
    feature_lab = paste0(feature_group, " | ", feature),
    model = factor(
      model,
      levels = c(
        "Unadjusted",
        "Minimally adjusted",
        "Primary adjusted",
        "Technical/acute-state adjusted",
        "Baseline-metabolite adjusted"
      )
    )
  )

for (analysis_i in analysis_order) {
  panel_df <- comp_all %>% filter(analysis == analysis_i)
  if (nrow(panel_df) == 0) next

  p <- panel_df %>%
    ggplot(aes(x = feature_lab, y = estimate, color = model)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
    geom_pointrange(
      aes(ymin = conf.low, ymax = conf.high),
      position = position_dodge(width = 0.7),
      fatten = 1.5,
      linewidth = 0.35
    ) +
    coord_flip() +
    labs(
      x = NULL,
      y = "Standardized beta (95% CI)",
      color = "Model",
      title = paste0("Sensitivity/model comparison: ", analysis_i)
    ) +
    theme_minimal(base_size = 16)
  print(p)
}
```

## Figure A4. Comparative pattern across Q1-Q3

```{r}
#| label: fig-age-comp
#| fig-cap: "Comparative effect sizes across the two concurrent models and the prospective model for one matched exposure family selected by the best observed q-value across Q1-Q3."
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 11
bases_q1 <- unique(strip_visit_suffix(q1$exposure))
bases_q2 <- unique(strip_visit_suffix(q2$exposure))
bases_q3 <- unique(strip_visit_suffix(q3$exposure))
common_bases <- intersect(intersect(bases_q1, bases_q2), bases_q3)

base_score <- bind_rows(
  q1 %>% mutate(base = strip_visit_suffix(exposure)),
  q2 %>% mutate(base = strip_visit_suffix(exposure)),
  q3 %>% mutate(base = strip_visit_suffix(exposure))
) %>%
  filter(base %in% common_bases) %>%
  group_by(base) %>%
  summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
  arrange(best_q)

selected_base <- if (nrow(base_score) > 0) base_score$base[[1]] else NA_character_

exp_q1 <- q1 %>% filter(strip_visit_suffix(exposure) == selected_base) %>% distinct(exposure) %>% slice_head(n = 1) %>% pull(exposure)
exp_q2 <- q2 %>% filter(strip_visit_suffix(exposure) == selected_base) %>% distinct(exposure) %>% slice_head(n = 1) %>% pull(exposure)
exp_q3 <- q3 %>% filter(strip_visit_suffix(exposure) == selected_base) %>% distinct(exposure) %>% slice_head(n = 1) %>% pull(exposure)

comp_age <- bind_rows(
  q1 %>% filter(exposure == exp_q1) %>% mutate(model = "Concurrent at 2 years"),
  q2 %>% filter(exposure == exp_q2) %>% mutate(model = "Concurrent at 5-6 years"),
  q3 %>% filter(exposure == exp_q3) %>% mutate(model = "Prospective 2-year to 5-6-year")
) %>%
  add_feature_meta()

top_comp_features <- comp_age %>%
  group_by(feature) %>%
  summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
  slice_min(best_q, n = 20, with_ties = FALSE)

comp_age_plot <- comp_age %>%
  semi_join(top_comp_features, by = "feature") %>%
  mutate(
    feature_lab = paste0(feature_group, " | ", feature),
    model = factor(
      model,
      levels = c("Concurrent at 2 years", "Concurrent at 5-6 years", "Prospective 2-year to 5-6-year")
    )
  )

comp_age_plot %>%
  ggplot(aes(x = feature_lab, y = estimate, color = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_pointrange(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.7),
    fatten = 1.4,
    linewidth = 0.35
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Standardized beta (95% CI)",
    color = "Analysis frame",
    title = paste0("Comparative summary across ages for exposure family: ", selected_base)
  ) +
  theme_minimal(base_size = 16)
```

## Figure A5. Metabolite-class composition of significant hits

```{r}
#| label: fig-class-composition
#| fig-cap: "Metabolite-class composition of significant associations (q < 0.10) across the three study linear-model question families."
class_comp <- bind_rows(
  add_exposure_family(q1) %>% mutate(analysis = "Q1 concurrent"),
  add_exposure_family(q2) %>% mutate(analysis = "Q2 concurrent"),
  add_exposure_family(q3) %>% mutate(analysis = "Q3 prospective")
) %>%
  filter(q.value < 0.1) %>%
  add_feature_meta() %>%
  count(analysis, exposure_family, feature_group, name = "n_hits") %>%
  group_by(analysis, exposure_family) %>%
  mutate(hit_fraction = n_hits / sum(n_hits)) %>%
  ungroup() %>%
  mutate(exposure_family = factor(exposure_family, levels = family_order))

make_table(
  class_comp %>% mutate(hit_fraction = round(hit_fraction, 3)),
  caption = "Metabolite-class composition of significant study hits"
)

class_comp %>%
  ggplot(aes(x = analysis, y = hit_fraction, fill = feature_group)) +
  geom_col(position = "fill", width = 0.65) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ exposure_family) +
  labs(
    x = "Question family",
    y = "Share of significant hits",
    fill = "Metabolite class",
    title = "Class composition of significant study findings"
  ) +
  theme_minimal(base_size = 15)
```

## Figure A6. Metabolite correlation heatmap

```{r}
#| label: fig-metabolite-correlation
#| fig-cap: "Correlation heatmap across all 250 metabolite features. Correlations are computed from the pooled standardized metabolite matrix and ordered by official NMR group."
#| classes: fig-medium
#| fig-width: 11
#| fig-height: 9

mat_data <- as.matrix(df_all[, intersect(metabolite_cols, names(df_all)), drop = FALSE])
storage.mode(mat_data) <- "numeric"
keep_cols <- colSums(is.finite(mat_data)) >= 20
mat_data <- mat_data[, keep_cols, drop = FALSE]

if (ncol(mat_data) < 2) {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "Metabolite correlation heatmap unavailable") +
    theme_void()
} else {
  cor_mat <- suppressWarnings(cor(mat_data, use = "pairwise.complete.obs"))
  cor_mat[!is.finite(cor_mat)] <- 0
  diag(cor_mat) <- 1
  cor_mat <- pmax(pmin(cor_mat, 1), -1)

  # Fetch NMR group metadata
  rd <- as.data.frame(rowData(experiments(mae)[["visit_all"]]), stringsAsFactors = FALSE)
  rd$Feature <- rownames(rd)
  rd$Group <- as.character(rd$Group)
  rd$Group[is.na(rd$Group) | rd$Group == ""] <- "Unknown"

  feature_meta_here <- rd %>%
    filter(Feature %in% colnames(cor_mat)) %>%
    arrange(match(Feature, colnames(cor_mat)))

  group_order <- unique(feature_meta_here$Group)
  
  ordered_names <- unlist(lapply(group_order, function(group_nm) {
    members <- feature_meta_here$Feature[feature_meta_here$Group == group_nm]
    if (length(members) > 2) {
      group_cor <- cor_mat[members, members, drop = FALSE]
      members[hclust(as.dist(1 - abs(group_cor)))$order]
    } else {
      members
    }
  }), use.names = FALSE)
  ordered_names <- intersect(ordered_names, colnames(cor_mat))

  feature_order <- data.frame(
    Feature = ordered_names,
    Position = seq_along(ordered_names),
    stringsAsFactors = FALSE
  ) %>%
    left_join(feature_meta_here[, c("Feature", "Group")], by = "Feature")

  group_ranges <- feature_order %>%
    group_by(Group) %>%
    summarise(
      start = min(Position) - 0.5,
      end = max(Position) + 0.5,
      center = mean(Position),
      n_features = n(),
      .groups = "drop"
    ) %>%
    mutate(Group = factor(Group, levels = group_order)) %>%
    arrange(Group)

  group_label_df <- group_ranges %>%
    mutate(Label = stringr::str_wrap(as.character(Group), width = 24))

  boundary_pos <- sort(unique(c(group_ranges$start, group_ranges$end)))
  boundary_pos <- boundary_pos[boundary_pos > 0.5 & boundary_pos < length(ordered_names) + 0.5]

  cor_long_df <- as.data.frame(cor_mat[ordered_names, ordered_names]) %>%
    mutate(Feature1 = rownames(cor_mat[ordered_names, ordered_names])) %>%
    pivot_longer(-Feature1, names_to = "Feature2", values_to = "Correlation") %>%
    left_join(feature_order %>% select(Feature1 = Feature, X = Position), by = "Feature1") %>%
    left_join(feature_order %>% select(Feature2 = Feature, Y = Position), by = "Feature2")

  ggplot(cor_long_df, aes(x = X, y = Y, fill = Correlation)) +
    geom_tile() +
    geom_vline(xintercept = boundary_pos, color = "grey45", linewidth = 0.18) +
    geom_hline(yintercept = boundary_pos, color = "grey45", linewidth = 0.18) +
    ggrepel::geom_text_repel(
      data = group_label_df,
      aes(x = 0.5, y = center, label = Label),
      inherit.aes = FALSE,
      nudge_x = -45,
      direction = "y",
      hjust = 1,
      size = 2.35,
      segment.color = "grey35",
      segment.size = 0.18,
      min.segment.length = 0,
      box.padding = 0.12,
      point.padding = 0.05,
      max.overlaps = Inf
    ) +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    scale_x_continuous(limits = c(-70, length(ordered_names) + 0.5), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0.5, length(ordered_names) + 0.5), expand = c(0, 0)) +
    coord_fixed(clip = "off") +
    labs(x = NULL, y = NULL, fill = "r", title = "Metabolite correlation structure") +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      plot.margin = margin(5.5, 5.5, 5.5, 120),
      legend.position = "right"
    )
}
```
