---
title: "Supplementary tables"
---

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")
library(dplyr)
library(tidyr)

results <- readRDS("../output/results/targeted_v2/child_diet_study1_results_v2.rds")
model_status <- results$model_status

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = c("visit_6", "visit_7")
)

raw_df <- extract_long_from_mae(
  analysis_obj$mae,
  visits = c("visit_6", "visit_7")
)

q1 <- results$q1
q2 <- results$q2
q3 <- results$q3
q4 <- results$q4
q4_split <- split_q4_terms(q4)
q4_main <- q4_split$main
q4_interaction <- q4_split$interaction
q1_sens <- results$q1_sensitivity
q2_sens <- results$q2_sensitivity
q3_sens <- results$q3_sensitivity
q4_sens <- results$q4_sensitivity

q1_expos <- sort(unique(model_status$exposure[model_status$question == "Q1"]))
q2_expos <- sort(unique(model_status$exposure[model_status$question == "Q2"]))
q4_pairs <- tibble(
  exposure_2y = q1_expos,
  exposure_56y = sub("6$", "7", q1_expos),
  exposure_family = sub("6$", "", gsub("_6$", "", q1_expos))
)

v6 <- raw_df %>%
  filter(visit_num == 6) %>%
  select(StudyID, all_of(q1_expos))
v6$has_q1_exposure <- has_any_nonmissing(v6, q1_expos)
v6 <- v6 %>% select(StudyID, has_q1_exposure, all_of(q1_expos))

v7 <- raw_df %>%
  filter(visit_num == 7) %>%
  select(StudyID, all_of(q2_expos))
v7$has_q2_exposure <- has_any_nonmissing(v7, q2_expos)
v7 <- v7 %>% select(StudyID, has_q2_exposure, all_of(q2_expos))

both_visits <- inner_join(v6, v7, by = "StudyID", suffix = c("_2y", "_56y"))

has_paired_exposure <- function(df, pairs) {
  keep <- pairs$exposure_2y %in% names(df) & pairs$exposure_56y %in% names(df)
  pairs <- pairs[keep, , drop = FALSE]
  if (nrow(pairs) == 0) return(rep(FALSE, nrow(df)))
  Reduce(`|`, lapply(seq_len(nrow(pairs)), function(i) {
    !is.na(df[[pairs$exposure_2y[i]]]) & !is.na(df[[pairs$exposure_56y[i]]])
  }))
}

q4_eligible <- has_paired_exposure(both_visits, q4_pairs)
```

## Supplementary Table S1. Analytic sample derivation

```{r}
#| label: supplementary-tables-01
sample_tbl <- tibble(
  analytic_step = c(
    "Children with at least one metabolomics panel",
    "Children with a 2-year metabolomics panel",
    "Children with a 5-6-year metabolomics panel",
    "Children with metabolomics at both 2 years and 5-6 years",
    "Children with any analyzed 2-year diet exposure in the 2-year panel",
    "Children with any analyzed 5-6-year diet exposure in the 5-6-year panel",
    "Children contributing to the prospective question set",
    "Children with paired diet exposure data across both ages for repeated-measures models"
  ),
  n_children = c(
    n_distinct(raw_df$StudyID),
    n_distinct(raw_df$StudyID[raw_df$visit_num == 6]),
    n_distinct(raw_df$StudyID[raw_df$visit_num == 7]),
    nrow(both_visits),
    sum(v6$has_q1_exposure, na.rm = TRUE),
    sum(v7$has_q2_exposure, na.rm = TRUE),
    sum(both_visits$has_q1_exposure, na.rm = TRUE),
    sum(q4_eligible, na.rm = TRUE)
  )
)

make_table(sample_tbl, caption = "Supplementary Table S1. Analytic sample derivation")
```

## Supplementary Table S2. Exposure definitions

```{r}
#| label: supplementary-tables-02
exposure_tbl <- bind_rows(
  tibble(
    component = "Concurrent at 2 Years",
    exposure_timing = "2-year child diet",
    exposure_code = q1_expos
  ),
  tibble(
    component = "Concurrent at 5-6 Years",
    exposure_timing = "5-6-year child diet",
    exposure_code = q2_expos
  ),
  tibble(
    component = "Prospective 2-Year to 5-6-Year Models",
    exposure_timing = "2-year child diet",
    exposure_code = q1_expos
  ),
  q4_pairs %>%
    transmute(
      component = "Repeated-Measures Models",
      exposure_timing = "Paired 2-year and 5-6-year child diet",
      exposure_code = exposure_family,
      paired_codes = paste0(exposure_2y, " / ", exposure_56y)
    )
) %>%
  mutate(
    exposure_label = pretty_var(exposure_code),
    scale = case_when(
      grepl("classes", exposure_code, ignore.case = TRUE) ~ "Categorical class variable",
      grepl("points", exposure_code, ignore.case = TRUE) ~ "Continuous score",
      TRUE ~ "Continuous percentage or intake metric"
    )
  ) %>%
  relocate(exposure_label, .after = exposure_code)

make_table(exposure_tbl, caption = "Supplementary Table S2. Exposure definitions and coding")
```

## Supplementary Table S3. Primary model specifications

```{r}
#| label: supplementary-tables-03
model_tbl <- tibble(
  analysis_component = c(
    "Concurrent at 2 Years",
    "Concurrent at 5-6 Years",
    "Prospective: 2-Year Diet to 5-6-Year Metabolites",
    "Repeated-Measures Across 2 Years and 5-6 Years"
  ),
  outcome = c(
    "2-year metabolite",
    "5-6-year metabolite",
    "5-6-year metabolite with baseline metabolite adjustment",
    "Repeated metabolite across 2 years and 5-6 years"
  ),
  model_family = c(
    "Feature-wise linear model",
    "Feature-wise linear model",
    "Feature-wise linear model",
    "Feature-wise mixed model"
  ),
  primary_covariates = c(
    "Intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration",
    "Intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration",
    "2-year exposure; 2-year baseline metabolite; intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration",
    "Paired diet exposure across child ages; intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration"
  ),
  sensitivity_additions = c(
    "Small-for-gestational-age; macrosomy; healthy-diet score at 2 years",
    "Maternal smoking; prematurity; small-for-gestational-age; macrosomy; healthy-diet score at 5-6 years",
    "Primary prospective block plus small-for-gestational-age; macrosomy; healthy-diet score at 2 years",
    "Primary repeated-measures block plus small-for-gestational-age; macrosomy; time-varying healthy-diet score"
  ),
  multiple_testing = c(
    "Benjamini-Hochberg across metabolites within each exposure-specific concurrent model",
    "Benjamini-Hochberg across metabolites within each exposure-specific concurrent model",
    "Benjamini-Hochberg across metabolites within each exposure-specific prospective model",
    "Benjamini-Hochberg across metabolites applied separately within each repeated-measures exposure and coefficient family"
  )
)

make_table(model_tbl, caption = "Supplementary Table S3. Primary model specifications")
```

## Supplementary Table S4. Returned result inventory

```{r}
#| label: supplementary-tables-04
inventory_tbl <- build_result_inventory(results)

make_table(inventory_tbl, caption = "Supplementary Table S4. Returned result inventory")
```

## Supplementary Table S5. Candidate associations with q < 0.10

```{r}
#| label: supplementary-tables-05
candidate_linear <- bind_rows(
  q1 %>% mutate(component = "Concurrent at 2 Years"),
  q2 %>% mutate(component = "Concurrent at 5-6 Years"),
  q3 %>% mutate(component = "Prospective: 2-Year Diet to 5-6-Year Metabolites")
) %>%
  filter(q.value < 0.10) %>%
  mutate(multilevel_factor = grepl("\\|", term)) %>%
  transmute(
    component,
    exposure_family = exposure_family(exposure),
    exposure = pretty_var(exposure),
    feature,
    term = if_else(multilevel_factor, "Omnibus multi-level factor test", NA_character_),
    n,
    beta = if_else(multilevel_factor, NA_real_, estimate),
    conf_low = if_else(multilevel_factor, NA_real_, conf.low),
    conf_high = if_else(multilevel_factor, NA_real_, conf.high),
    p_value = p.value,
    q_value = q.value
  )

candidate_q4 <- bind_rows(
  q4_main %>% mutate(component = "Repeated-Measures Main Effect"),
  q4_interaction %>% mutate(component = "Repeated-Measures Diet-by-Time Interaction")
) %>%
  filter(q.value < 0.10) %>%
  transmute(
    component,
    exposure_family = as.character(exposure_family(exposure)),
    exposure = pretty_var(exposure),
    feature,
    term,
    n_obs,
    n_ind,
    beta = estimate,
    conf_low = estimate - 1.96 * std.error,
    conf_high = estimate + 1.96 * std.error,
    p_value = p.value,
    q_value = q.value
  )

candidate_tbl <- bind_rows(
  candidate_linear %>% mutate(n_obs = n, n_ind = NA_real_),
  candidate_q4 %>% rename(n = n_obs)
) %>%
  mutate(exposure_family = factor(exposure_family, levels = family_order)) %>%
  arrange(exposure_family, q_value, p_value)

make_table(candidate_tbl, caption = "Supplementary Table S5. Candidate associations meeting q < 0.10")
```

## Supplementary Table S6. Primary-vs-sensitivity stability summary

```{r}
#| label: supplementary-tables-06
stability_tbl <- bind_rows(
  compare_primary_sensitivity(q1, q1_sens, c("exposure", "feature", "visit")) %>% mutate(component = "Concurrent at 2 Years"),
  compare_primary_sensitivity(q2, q2_sens, c("exposure", "feature", "visit")) %>% mutate(component = "Concurrent at 5-6 Years"),
  compare_primary_sensitivity(q3, q3_sens, c("exposure", "feature", "visit_exposure", "visit_outcome")) %>% mutate(component = "Prospective: 2-Year Diet to 5-6-Year Metabolites"),
  compare_primary_sensitivity(q4, q4_sens, c("exposure", "feature", "term")) %>% mutate(component = "Repeated-Measures")
) %>%
  mutate(exposure_family = exposure_family(exposure)) %>%
  group_by(component, exposure_family) %>%
  summarise(
    matched_rows = n(),
    median_abs_beta_change = median(abs_delta, na.rm = TRUE),
    p90_abs_beta_change = quantile(abs_delta, 0.90, na.rm = TRUE),
    direction_flip_fraction = mean(direction_flip, na.rm = TRUE),
    q_reclassification_fraction = mean(q_reclassified, na.rm = TRUE),
    .groups = "drop"
  )

make_table(stability_tbl, caption = "Supplementary Table S6. Primary-vs-sensitivity stability summary")
```

## Supplementary Table S7. Saved execution status inventory

```{r}
#| label: supplementary-tables-07
status_tbl <- model_status %>%
  mutate(
    component = case_when(
      question == "Q1" ~ "Concurrent at 2 Years",
      question == "Q2" ~ "Concurrent at 5-6 Years",
      question == "Q3" ~ "Prospective: 2-Year Diet to 5-6-Year Metabolites",
      question == "Q4" ~ "Repeated-Measures",
      TRUE ~ question
    ),
    exposure = pretty_var(exposure),
    complete_case_n = ifelse(
      is.na(n_complete_min) | !is.finite(n_complete_min),
      NA_character_,
      ifelse(n_complete_min == n_complete_max, as.character(n_complete_min), paste0(n_complete_min, "-", n_complete_max))
    )
  ) %>%
  select(component, analysis, exposure, term_family, status, reason, complete_case_n, n_features_tested)

make_table(status_tbl, caption = "Supplementary Table S7. Saved execution-status inventory")
```
