---
title: "Clinical characteristics"
---

## Setup

```{r}
#| label: clinical-characteristics-01
source("analysis_data_utils.R")
source("../../scripts/methods_common.R")
library(dplyr)
library(tidyr)

visits <- c(6L, 7L)

visit_label <- function(v) ifelse(v == 6L, "2 Years", "5-6 Years")

pretty_var <- function(x) {
  x <- gsub("CEPercent", "Percent ", x)
  x <- gsub("CFiball", "Fibre ", x)
  x <- gsub("C_", "", x)
  x <- gsub("DietQuality", "Diet Quality ", x)
  x <- gsub("CHO", "Carbohydrate", x)
  x <- gsub("Pro", "Protein", x)
  x <- gsub("FAn3", "n-3 fatty acids", x)
  x <- gsub("FAn6", "n-6 fatty acids", x)
  x <- gsub("SFA", "SFA", x)
  x <- gsub("MUFA", "MUFA", x)
  x <- gsub("PUFA", "PUFA", x)
  x <- gsub("_", " ", x, fixed = TRUE)
  x <- gsub("\\s+", " ", trimws(x))
  tools::toTitleCase(tolower(x))
}

exposure_map <- list(
  "6" = c(
    "CEPercentProDiet6", "CEPercentCHODiet6", "CEPercentFatDiet6", "CEPercentSFADiet6",
    "CEPercentMUFADiet6", "CEPercentPUFADiet6", "CEPercentFAn3Diet6", "CEPercentFAn6Diet6",
    "CFiballDiet6", "C_DietQuality_points_6", "C_DietQuality_classes_6"
  ),
  "7" = c(
    "CEPercentProDiet7", "CEPercentCHODiet7", "CEPercentFatDiet7", "CEPercentSFADiet7",
    "CEPercentMUFADiet7", "CEPercentPUFADiet7", "CEPercentFAn3Diet7", "CEPercentFAn6Diet7",
    "CFiballDiet7", "C_DietQuality_points_7", "C_DietQuality_classes_7"
  )
)

core_v6 <- c(
  "Intervention", "MPrimipara", "MUniEdu", "MGDMOGTT1OR2Fi",
  "CGender", "CWeightBirth", "Bfdurationm"
)

core_v7 <- c(
  "Intervention", "MPrimipara", "MUniEdu", "MGDMOGTT1OR2Fi",
  "CGender", "CWeightBirth", "Bfdurationm"
)

sensitivity_covars <- c("MPRSmoke", "Premature", "CSGA", "CMacrosomy", "CHealthy6", "CHealthy7")

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = paste0("visit_", visits)
)

mae <- analysis_obj$mae
```

## Objective-specific context

Child diet exposures, clinical characteristics, and analytic panel characteristics are summarized here for 2 years and 5-6 years.

## Sample counts by visit and intervention

```{r}
#| label: clinical-characteristics-02
counts_tbl <- bind_rows(lapply(visits, function(v) {
  vn <- paste0("visit_", v)
  if (!vn %in% names(experiments(mae))) return(NULL)

  cd <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[[vn]]), stringsAsFactors = FALSE)
  if (!"Intervention" %in% names(cd)) {
    return(data.frame(
      visit = v,
      Intervention = "missing",
      n = nrow(cd),
      stringsAsFactors = FALSE
    ))
  }

  cd %>%
    mutate(Intervention = as.character(Intervention)) %>%
    group_by(child_age = visit_label(v), Intervention) %>%
    summarise(n = n(), .groups = "drop")
}))

make_table(counts_tbl, caption = "Child-age panel counts by intervention")
```

## Continuous variables by visit

```{r}
#| label: clinical-characteristics-03
continuous_summary <- function(df, vars, group_var = "Intervention") {
  vars <- intersect(vars, names(df))
  if (length(vars) == 0) return(data.frame())

  num_vars <- vars[vapply(df[vars], function(x) {
    x_num <- suppressWarnings(as.numeric(x))
    sum(is.finite(x_num)) >= 20L && length(unique(x_num[is.finite(x_num)])) >= 8L
  }, logical(1))]

  if (length(num_vars) == 0) return(data.frame())

  out <- bind_rows(lapply(num_vars, function(vn) {
    x <- suppressWarnings(as.numeric(df[[vn]]))

    overall <- data.frame(
      variable = pretty_var(vn),
      variable_code = vn,
      group = "Overall",
      n = sum(is.finite(x)),
      mean = mean(x, na.rm = TRUE),
      sd = stats::sd(x, na.rm = TRUE),
      median = stats::median(x, na.rm = TRUE),
      q1 = stats::quantile(x, 0.25, na.rm = TRUE),
      q3 = stats::quantile(x, 0.75, na.rm = TRUE),
      stringsAsFactors = FALSE
    )

    by_group <- data.frame()
    if (group_var %in% names(df)) {
      by_group <- bind_rows(lapply(sort(unique(as.character(df[[group_var]]))), function(g) {
        idx <- as.character(df[[group_var]]) == g
        gx <- x[idx]
        data.frame(
          variable = pretty_var(vn),
          variable_code = vn,
          group = g,
          n = sum(is.finite(gx)),
          mean = mean(gx, na.rm = TRUE),
          sd = stats::sd(gx, na.rm = TRUE),
          median = stats::median(gx, na.rm = TRUE),
          q1 = stats::quantile(gx, 0.25, na.rm = TRUE),
          q3 = stats::quantile(gx, 0.75, na.rm = TRUE),
          stringsAsFactors = FALSE
        )
      }))
    }

    bind_rows(overall, by_group)
  }))

  out
}

cont_tbl <- bind_rows(lapply(visits, function(v) {
  vn <- paste0("visit_", v)
  if (!vn %in% names(experiments(mae))) return(NULL)
  cd <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[[vn]]), stringsAsFactors = FALSE)
  vars <- unique(c(exposure_map[[as.character(v)]], if (v == 6L) core_v6 else core_v7, sensitivity_covars))
  tab <- continuous_summary(cd, vars, group_var = "Intervention")
  if (nrow(tab) == 0) return(NULL)
  tab$child_age <- visit_label(v)
  tab
})) %>%
  select(child_age, variable, variable_code, group, n, mean, sd, median, q1, q3)

make_table(round_numeric(cont_tbl, 3), caption = "Continuous clinical and exposure summaries by child age")
```

## Categorical variables by visit

```{r}
#| label: clinical-characteristics-04
categorical_summary <- function(df, vars, group_var = "Intervention") {
  vars <- intersect(vars, names(df))
  if (length(vars) == 0) return(data.frame())

  cat_vars <- vars[vapply(df[vars], function(x) {
    x_chr <- as.character(x)
    n_non_na <- sum(!is.na(x_chr) & nzchar(x_chr))
    n_non_na >= 20L && length(unique(x_chr[!is.na(x_chr) & nzchar(x_chr)])) <= 10L
  }, logical(1))]

  if (length(cat_vars) == 0) return(data.frame())

  bind_rows(lapply(cat_vars, function(vn) {
    xv <- as.character(df[[vn]])
    xv[is.na(xv) | xv == ""] <- "<missing>"

    overall <- as.data.frame(table(xv), stringsAsFactors = FALSE) %>%
      transmute(
        variable = pretty_var(vn),
        variable_code = vn,
        group = "Overall",
        level = xv,
        n = Freq,
        pct = 100 * Freq / sum(Freq)
      )

    by_group <- data.frame()
    if (group_var %in% names(df)) {
      g <- as.character(df[[group_var]])
      g[is.na(g) | g == ""] <- "<missing_group>"
      by_group <- as.data.frame(table(g, xv), stringsAsFactors = FALSE) %>%
        group_by(g) %>%
        mutate(pct = 100 * Freq / sum(Freq)) %>%
        ungroup() %>%
        transmute(
          variable = pretty_var(vn),
          variable_code = vn,
          group = g,
          level = xv,
          n = Freq,
          pct = pct
        )
    }

    bind_rows(overall, by_group)
  }))
}

cat_tbl <- bind_rows(lapply(visits, function(v) {
  vn <- paste0("visit_", v)
  if (!vn %in% names(experiments(mae))) return(NULL)
  cd <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[[vn]]), stringsAsFactors = FALSE)
  vars <- unique(c(exposure_map[[as.character(v)]], if (v == 6L) core_v6 else core_v7, sensitivity_covars))
  tab <- categorical_summary(cd, vars, group_var = "Intervention")
  if (nrow(tab) == 0) return(NULL)
  tab$child_age <- visit_label(v)
  tab
})) %>%
  select(child_age, variable, variable_code, group, level, n, pct)

make_table(round_numeric(cat_tbl, 3), caption = "Categorical clinical and exposure summaries by child age")
```

## Requested table of characteristics

These tables summarize visit-specific characteristics overall and by child diet quality group. Continuous variables are median (IQR); categorical variables are n (%).

P-values compare Good, Moderate, and Poor diet quality groups within the same visit. Continuous variables use Kruskal-Wallis tests; categorical variables use Fisher exact tests, with chi-square fallback if Fisher exact fails.

```{r}
#| label: clinical-characteristics-04b
child_diet_visit6_path <- "../data/child_diet_characteristics_visit6.csv"
child_diet_visit7_path <- "../data/child_diet_characteristics_visit7.csv"

if (file.exists(child_diet_visit6_path)) {
  make_table(
    read.csv(child_diet_visit6_path, check.names = FALSE, colClasses = "character", na.strings = character()),
    caption = "Table of characteristics: visit 6, 2 years"
  )
}

if (file.exists(child_diet_visit7_path)) {
  make_table(
    read.csv(child_diet_visit7_path, check.names = FALSE, colClasses = "character", na.strings = character()),
    caption = "Table of characteristics: visit 7, 5-6 years"
  )
}
```

## Dietary intake and diet quality

This table shows the descriptive summary of nutrient intakes and diet quality metrics across the two study visits (Visit 6 at 2 years of age and Visit 7 at 5–6 years of age) for the children included in the primary analyses. Continuous variables are represented as mean (SD); categorical diet quality is represented as n (%).

```{r}
#| label: clinical-characteristics-diet-descriptive
# 1. Summarise continuous
get_cont_summary <- function(df, vname) {
  val <- as.numeric(df[[vname]])
  val <- val[is.finite(val)]
  sprintf("%.1f (%.1f)", mean(val), sd(val))
}

vars_v6 <- c(
  "CEPercentProDiet6", "CEPercentCHODiet6", "CEPercentFatDiet6", "CEPercentSFADiet6",
  "CEPercentMUFADiet6", "CEPercentPUFADiet6", "CEPercentFAn3Diet6", "CEPercentFAn6Diet6",
  "CFiballDiet6", "C_DietQuality_points_6"
)

vars_v7 <- c(
  "CEPercentProDiet7", "CEPercentCHODiet7", "CEPercentFatDiet7", "CEPercentSFADiet7",
  "CEPercentMUFADiet7", "CEPercentPUFADiet7", "CEPercentFAn3Diet7", "CEPercentFAn6Diet7",
  "CFiballDiet7", "C_DietQuality_points_7"
)

row_labels <- c(
  "Protein (E%)", "Carbohydrate (E%)", "Total fat (E%)", "SFA (E%)",
  "MUFA (E%)", "PUFA (E%)", "n-3 fatty acids (E%)", "n-6 fatty acids (E%)",
  "Fiber (g)", "Diet quality score (points)"
)

cd6 <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[["visit_6"]]))
cd7 <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[["visit_7"]]))

cont_tbl_6 <- sapply(vars_v6, function(vn) get_cont_summary(cd6, vn))
cont_tbl_7 <- sapply(vars_v7, function(vn) get_cont_summary(cd7, vn))

cont_rows <- data.frame(
  `Dietary Factor` = row_labels,
  `2 Years (Visit 6)` = as.character(cont_tbl_6),
  `5-6 Years (Visit 7)` = as.character(cont_tbl_7),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

# 2. Summarise categorical
cat_summary <- function(df, vname) {
  x <- tolower(trimws(as.character(haven::as_factor(df[[vname]]))))
  total_nonmiss <- sum(!is.na(x) & x != "")
  
  classes <- c("good", "moderate", "poor")
  res <- sapply(classes, function(cl) {
    if (cl == "good") {
      n <- sum(grepl("good", x))
    } else if (cl == "moderate") {
      n <- sum(grepl("moderate", x))
    } else {
      n <- sum(grepl("poor|needs to be vastly improved", x))
    }
    pct <- 100 * n / total_nonmiss
    sprintf("%d (%.1f%%)", n, pct)
  })
  res
}

cat_v6 <- cat_summary(cd6, "C_DietQuality_classes_6")
cat_v7 <- cat_summary(cd7, "C_DietQuality_classes_7")

cat_rows <- data.frame(
  `Dietary Factor` = c("Diet quality category (n, %)", "  Good", "  Moderate", "  Poor"),
  `2 Years (Visit 6)` = c("", cat_v6[["good"]], cat_v6[["moderate"]], cat_v6[["poor"]]),
  `5-6 Years (Visit 7)` = c("", cat_v7[["good"]], cat_v7[["moderate"]], cat_v7[["poor"]]),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

diet_tbl <- rbind(cont_rows, cat_rows)
make_table(diet_tbl, caption = "Summary of dietary factors at baseline and follow-up visits")
```


## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r}
#| label: clinical-characteristics-05
#| echo: false
sessionInfo()
```
:::
