---
title: "Alpha diversity"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
group_variable <- config$variables$grouping
recipe_parameters <- config$analyses$recipe_parameters[["microbiome.alpha_diversity"]] %||% list()
requested_metrics <- unlist(
  recipe_parameters$metrics %||% c("observed", "shannon", "simpson"),
  use.names = FALSE
)
metrics <- intersect(requested_metrics, names(metadata))
if (!length(metrics)) stop("No configured alpha-diversity metrics are available.")
```

## Analysis contract

This deterministic recipe calculates observed richness, Shannon diversity, and Simpson diversity from the validated abundance matrix. Sample identifiers were aligned before calculation; unmatched samples are recorded in `data_validation.json`.

```{r}
contract <- data.frame(
  field = c("Samples", "Features", "Input scale", "Grouping variable"),
  value = c(
    length(analysis_data$sample_ids),
    length(analysis_data$feature_ids),
    analysis_data$input_scale,
    group_variable %||% "Not configured"
  )
)
knitr::kable(contract)
```

## Diversity distributions

```{r}
#| fig-width: 9
#| fig-height: 4
plot_rows <- do.call(rbind, lapply(metrics, function(metric) {
  data.frame(
    sample_id = metadata$.sample_id,
    metric = metric,
    value = metadata[[metric]],
    group = if (!is.null(group_variable) && group_variable %in% names(metadata)) {
      as.character(metadata[[group_variable]])
    } else {
      "All samples"
    },
    stringsAsFactors = FALSE
  )
}))

if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(
    ggplot2::ggplot(plot_rows, ggplot2::aes(x = group, y = value, fill = group)) +
      ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.7) +
      ggplot2::geom_jitter(width = 0.12, alpha = 0.65, size = 1.5) +
      ggplot2::facet_wrap(~metric, scales = "free_y") +
      ggplot2::labs(x = group_variable %||% NULL, y = NULL) +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::theme(legend.position = "none", axis.text.x = ggplot2::element_text(angle = 30, hjust = 1))
  )
}
```

## Group comparisons

```{r}
result_rows <- lapply(metrics, function(metric) {
  if (is.null(group_variable) || !group_variable %in% names(metadata)) {
    return(data.frame(
      metric = metric, test = "not_run", statistic = NA_real_,
      p.value = NA_real_, n = sum(is.finite(metadata[[metric]])),
      status = "grouping_variable_not_configured"
    ))
  }

  keep <- !is.na(metadata[[group_variable]]) & is.finite(metadata[[metric]])
  values <- metadata[[metric]][keep]
  groups <- droplevels(factor(metadata[[group_variable]][keep]))
  if (length(levels(groups)) < 2) {
    return(data.frame(
      metric = metric, test = "not_run", statistic = NA_real_,
      p.value = NA_real_, n = length(values), status = "fewer_than_two_groups"
    ))
  }

  if (length(levels(groups)) == 2) {
    test_result <- stats::wilcox.test(values ~ groups, exact = FALSE)
    test_name <- "Wilcoxon rank-sum"
  } else {
    test_result <- stats::kruskal.test(values ~ groups)
    test_name <- "Kruskal-Wallis"
  }
  data.frame(
    metric = metric,
    test = test_name,
    statistic = unname(test_result$statistic),
    p.value = test_result$p.value,
    n = length(values),
    status = "completed"
  )
})

alpha_results <- do.call(rbind, result_rows)
alpha_results$q.value <- stats::p.adjust(
  alpha_results$p.value,
  method = recipe_parameters$fdr_method %||% "BH"
)
dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
utils::write.csv(alpha_results, "../output/results/alpha_diversity.csv", row.names = FALSE)
knitr::kable(alpha_results, digits = 4)
```

## Interpretation boundary

P-values describe evidence for distributional differences in the configured groups. They do not identify causal effects, and alpha diversity alone does not summarize community composition.

## Reproducibility

```{r}
sessionInfo()
```
