---
title: "Beta diversity and ordination"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
if (!requireNamespace("vegan", quietly = TRUE)) {
  stop("Package 'vegan' is required for the beta-diversity recipe.")
}
analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
group_variable <- config$variables$grouping
relative_abundance <- analysis_data$relative_abundance
recipe_parameters <- config$analyses$recipe_parameters[["microbiome.beta_diversity"]] %||% list()
distance_method <- recipe_parameters$distance %||% "bray"
```

## Analysis contract

The configured `r distance_method` distances are calculated from the validated sample-aligned relative-abundance matrix. Principal coordinates analysis provides a two-dimensional representation of between-sample compositional differences.

```{r}
beta_distance <- vegan::vegdist(t(relative_abundance), method = distance_method)
sample_count <- attr(beta_distance, "Size")
if (sample_count < 3) stop("At least three samples are required for beta-diversity ordination.")

ordination <- stats::cmdscale(beta_distance, k = 2, eig = TRUE, add = TRUE)
coordinates <- as.data.frame(ordination$points)
names(coordinates) <- c("PCoA1", "PCoA2")
coordinates$sample_id <- rownames(coordinates)
metadata_columns <- setdiff(names(metadata), c(".sample_id", names(coordinates)))
coordinates <- cbind(
  coordinates,
  metadata[match(coordinates$sample_id, metadata$.sample_id), metadata_columns, drop = FALSE]
)

positive_eigenvalues <- ordination$eig[ordination$eig > 0]
variance_explained <- positive_eigenvalues / sum(positive_eigenvalues)
variance_explained <- c(variance_explained, 0, 0)
axis_labels <- c(
  sprintf("PCoA 1 (%.1f%%)", 100 * variance_explained[[1]]),
  sprintf("PCoA 2 (%.1f%%)", 100 * variance_explained[[2]])
)

dir.create("../output/derived", recursive = TRUE, showWarnings = FALSE)
dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
saveRDS(beta_distance, "../output/derived/beta_distance.rds")
utils::write.csv(coordinates, "../output/results/beta_ordination.csv", row.names = FALSE)
```

## Distance-based PCoA

```{r}
#| fig-width: 8
#| fig-height: 6
if (requireNamespace("ggplot2", quietly = TRUE)) {
  if (!is.null(group_variable) && group_variable %in% names(coordinates)) {
    plot <- ggplot2::ggplot(
      coordinates,
      ggplot2::aes(x = PCoA1, y = PCoA2, color = .data[[group_variable]])
    )
  } else {
    plot <- ggplot2::ggplot(coordinates, ggplot2::aes(x = PCoA1, y = PCoA2))
  }
  print(
    plot +
      ggplot2::geom_point(size = 2.8, alpha = 0.8) +
      ggplot2::labs(x = axis_labels[[1]], y = axis_labels[[2]], color = group_variable) +
      ggplot2::theme_minimal(base_size = 12)
  )
}
```

```{r}
knitr::kable(utils::head(coordinates, 20), digits = 4)
```

## Interpretation boundary

Distances and ordination are descriptive. Visual separation is not itself a statistical test; the paired PERMANOVA recipe evaluates the configured group association and reports dispersion diagnostics. The configured distance method was `r distance_method`.

## Reproducibility

```{r}
sessionInfo()
```
