---
title: "LimROTS differential abundance"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
required_packages <- c("LimROTS", "SummarizedExperiment", "S4Vectors")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Missing packages for LimROTS: ", paste(missing_packages, collapse = ", "))
}

analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
relative_abundance <- analysis_data$relative_abundance
group_variable <- config$variables$grouping
covariates <- intersect(
  unlist(config$variables$covariates %||% character(0), use.names = FALSE),
  names(metadata)
)
recipe_parameters <- config$analyses$recipe_parameters[["microbiome.limrots_differential_abundance"]] %||% list()
niter <- as.integer(recipe_parameters$bootstrap_iterations %||% 200)
pseudocount <- as.numeric(recipe_parameters$pseudocount %||% 0.000001)
if (is.null(group_variable) || !group_variable %in% names(metadata)) {
  stop("LimROTS requires a confirmed grouping variable in study_config.yml.")
}
```

## Analysis contract

LimROTS combines limma-style empirical Bayes modeling with a reproducibility-optimized test statistic. This recipe performs all pairwise contrasts among configured groups using centered log-ratio transformed relative abundances. LimROTS permutation-based FDR and Benjamini–Hochberg values are both retained.

```{r}
log_abundance <- log(relative_abundance + pseudocount)
clr_matrix <- sweep(log_abundance, 2, colMeans(log_abundance), FUN = "-")
group <- droplevels(factor(metadata[[group_variable]]))
group_levels <- levels(group)
if (length(group_levels) < 2) stop("LimROTS requires at least two observed groups.")
comparisons <- utils::combn(group_levels, 2, simplify = FALSE)

contract <- data.frame(
  field = c(
    "Features", "Samples", "Grouping variable", "Contrasts",
    "Transformation", "Bootstrap iterations", "Covariates"
  ),
  value = c(
    nrow(clr_matrix),
    ncol(clr_matrix),
    group_variable,
    length(comparisons),
    paste0("CLR; pseudocount=", pseudocount),
    niter,
    if (length(covariates)) paste(covariates, collapse = ", ") else "None"
  )
)
knitr::kable(contract)
```

## Pairwise differential-abundance results

```{r}
row_value <- function(row_data, candidates, default = NA_real_) {
  column <- candidates[candidates %in% names(row_data)][1]
  if (is.na(column)) rep(default, nrow(row_data)) else as.numeric(row_data[[column]])
}

run_limrots_contrast <- function(levels) {
  reference <- levels[[1]]
  comparison <- levels[[2]]
  keep <- group %in% c(reference, comparison)
  contrast_metadata <- metadata[keep, , drop = FALSE]
  contrast_metadata$.group <- factor(
    contrast_metadata[[group_variable]],
    levels = c(reference, comparison)
  )
  required_model_columns <- c(".group", covariates)
  complete_design <- stats::complete.cases(
    contrast_metadata[, required_model_columns, drop = FALSE]
  )
  contrast_metadata <- contrast_metadata[complete_design, , drop = FALSE]
  contrast_assay <- clr_matrix[, keep, drop = FALSE][, complete_design, drop = FALSE]

  if (nrow(contrast_metadata) < 6 || any(table(contrast_metadata$.group) < 3)) {
    return(list(
      result = NULL,
      status = data.frame(
        comparison = paste(comparison, "vs", reference),
        status = "skipped",
        n = nrow(contrast_metadata),
        note = "fewer_than_three_samples_in_a_group"
      )
    ))
  }

  colnames(contrast_assay) <- rownames(contrast_metadata)
  experiment <- SummarizedExperiment::SummarizedExperiment(
    assays = list(exprs = contrast_assay),
    colData = S4Vectors::DataFrame(contrast_metadata)
  )
  formula_terms <- c(".group", covariates)
  formula_string <- paste("~ 0 +", paste(formula_terms, collapse = " + "))
  fit <- try(
    LimROTS::LimROTS(
      x = experiment,
      niter = niter,
      K = min(100L, nrow(experiment)),
      log = FALSE,
      verbose = FALSE,
      meta.info = formula_terms,
      BPPARAM = NULL,
      group.name = ".group",
      formula.str = formula_string,
      robust = isTRUE(recipe_parameters$robust %||% TRUE),
      trend = isTRUE(recipe_parameters$trend %||% TRUE),
      permutating.group = FALSE
    ),
    silent = TRUE
  )
  if (inherits(fit, "try-error")) {
    return(list(
      result = NULL,
      status = data.frame(
        comparison = paste(comparison, "vs", reference),
        status = "failed",
        n = nrow(contrast_metadata),
        note = as.character(fit)
      )
    ))
  }

  row_data <- as.data.frame(SummarizedExperiment::rowData(fit))
  results <- data.frame(
    feature = rownames(fit),
    comparison = paste(comparison, "vs", reference),
    reference = reference,
    group = comparison,
    n = nrow(contrast_metadata),
    logFC = row_value(row_data, c("corrected.logfc", "logfc", "logFC")),
    p.value = row_value(row_data, c("pvalue", "p.value", "P.Value")),
    FDR = row_value(row_data, c("FDR", "fdr")),
    q.value = row_value(row_data, c("qvalue", "q.value")),
    BH.p.value = row_value(row_data, c("BH.pvalue", "adj.P.Val")),
    stringsAsFactors = FALSE
  )
  list(
    result = results,
    status = data.frame(
      comparison = paste(comparison, "vs", reference),
      status = "completed",
      n = nrow(contrast_metadata),
      note = "ok"
    )
  )
}

model_runs <- lapply(comparisons, run_limrots_contrast)
result_parts <- lapply(model_runs, `[[`, "result")
result_parts <- result_parts[!vapply(result_parts, is.null, logical(1))]
limrots_results <- if (length(result_parts)) do.call(rbind, result_parts) else data.frame()
model_status <- do.call(rbind, lapply(model_runs, `[[`, "status"))

if (nrow(limrots_results)) {
  limrots_results <- limrots_results[
    order(limrots_results$FDR, limrots_results$p.value, na.last = TRUE),
    ,
    drop = FALSE
  ]
}
dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
utils::write.csv(
  limrots_results,
  "../output/results/limrots_differential_abundance.csv",
  row.names = FALSE
)
utils::write.csv(
  model_status,
  "../output/results/limrots_model_status.csv",
  row.names = FALSE
)

if (nrow(limrots_results)) {
  knitr::kable(utils::head(limrots_results, 30), digits = 4)
} else {
  cat("No LimROTS contrast completed. See `limrots_model_status.csv`.")
}
```

## Effect overview

```{r}
#| fig-width: 10
#| fig-height: 6
if (nrow(limrots_results) && requireNamespace("ggplot2", quietly = TRUE)) {
  plot_data <- limrots_results
  plot_data$significant <- !is.na(plot_data$FDR) & plot_data$FDR < 0.05
  print(
    ggplot2::ggplot(
      plot_data,
      ggplot2::aes(x = logFC, y = -log10(p.value), color = significant)
    ) +
      ggplot2::geom_point(alpha = 0.75) +
      ggplot2::facet_wrap(~ comparison, scales = "free") +
      ggplot2::scale_color_manual(
        values = c("TRUE" = "#0f766e", "FALSE" = "#71717a"),
        guide = "none"
      ) +
      ggplot2::labs(x = "LimROTS log fold-change", y = "-log10(p-value)") +
      ggplot2::theme_minimal(base_size = 11)
  )
}
```

## Interpretation boundary

Differential-abundance conclusions can depend materially on normalization and model choice. LimROTS is presented as a sensitivity method alongside composition-aware primary methods; isolated LimROTS discoveries should not be treated as method-independent findings.

## Reproducibility

```{r}
sessionInfo()
```
