---
title: "PERMANOVA community comparison"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
source("recipe_runtime.R")
if (!requireNamespace("vegan", quietly = TRUE)) {
  stop("Package 'vegan' is required for the PERMANOVA recipe.")
}
analysis_data <- readRDS("../output/derived/analysis_data.rds")
config <- analysis_data$config
metadata <- analysis_data$metadata
group_variable <- config$variables$grouping
recipe_parameters <- config$analyses$recipe_parameters[["microbiome.permanova"]] %||% list()
beta_parameters <- config$analyses$recipe_parameters[["microbiome.beta_diversity"]] %||% list()
distance_method <- beta_parameters$distance %||% "bray"
if (is.null(group_variable) || !group_variable %in% names(metadata)) {
  stop("PERMANOVA requires a confirmed grouping variable in study_config.yml.")
}
if (!file.exists("../output/derived/beta_distance.rds")) {
  beta_distance <- vegan::vegdist(
    t(analysis_data$relative_abundance),
    method = distance_method
  )
} else {
  beta_distance <- readRDS("../output/derived/beta_distance.rds")
}
```

## Model specification

```{r}
grouping <- droplevels(factor(metadata[[group_variable]]))
if (length(levels(grouping)) < 2) stop("PERMANOVA requires at least two observed groups.")
model_data <- data.frame(grouping = grouping, row.names = metadata$.sample_id)
distance_labels <- attr(beta_distance, "Labels")
model_data <- model_data[distance_labels, , drop = FALSE]
permutations <- as.integer(recipe_parameters$permutations %||% 999)
seed <- as.integer(recipe_parameters$seed %||% 20260730)
set.seed(seed)

permanova_fit <- vegan::adonis2(
  beta_distance ~ grouping,
  data = model_data,
  permutations = permutations
)
permanova_results <- data.frame(
  term = rownames(permanova_fit),
  Df = permanova_fit$Df,
  SumOfSqs = permanova_fit$SumOfSqs,
  R2 = permanova_fit$R2,
  F = permanova_fit$F,
  p.value = permanova_fit$`Pr(>F)`,
  row.names = NULL,
  check.names = FALSE
)

dispersion <- vegan::betadisper(beta_distance, model_data$grouping)
set.seed(seed)
dispersion_test <- vegan::permutest(dispersion, permutations = permutations)
dispersion_table <- data.frame(
  term = rownames(dispersion_test$tab),
  Df = dispersion_test$tab[, "Df"],
  SumOfSqs = dispersion_test$tab[, "Sum Sq"],
  F = dispersion_test$tab[, "F"],
  p.value = dispersion_test$tab[, "Pr(>F)"],
  row.names = NULL,
  check.names = FALSE
)

dir.create("../output/results", recursive = TRUE, showWarnings = FALSE)
utils::write.csv(permanova_results, "../output/results/permanova.csv", row.names = FALSE)
utils::write.csv(dispersion_table, "../output/results/beta_dispersion.csv", row.names = FALSE)
```

## PERMANOVA results

```{r}
knitr::kable(permanova_results, digits = 4)
```

PERMANOVA tests whether the centroids of the configured groups differ in the configured `r distance_method` distance space. The reported R² is the proportion of distance variation associated with the grouping variable.

## Dispersion diagnostic

```{r}
knitr::kable(dispersion_table, digits = 4)
```

A significant dispersion result indicates that differing within-group variability may contribute to the PERMANOVA result and should be considered during interpretation.

## Reproducibility

```{r}
reproducibility <- data.frame(
  parameter = c("Distance", "Permutations", "Random seed", "Grouping variable"),
  value = c(distance_method, permutations, seed, group_variable)
)
knitr::kable(reproducibility)
sessionInfo()
```
