---
title: "Figures: maternal diet and child metabolites"
---

Figures from the early-pregnancy, late-pregnancy, mutual-timing, and longitudinal analyses. The longitudinal component now returns coefficient-level output across all attempted exposures, but it still generated substantial execution warnings, so the figures below should be read as supporting context rather than a strong standalone inferential block.

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(MultiAssayExperiment)
library(SummarizedExperiment)

results <- readRDS("../output/results/targeted/prenatal_diet_study2_results.rds")
age_res <- results$age_specific
long_res <- results$longitudinal
age_res_sens <- results$age_specific_sensitivity
long_res_sens <- results$longitudinal_sensitivity
model_status <- results$model_status

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = c("visit_4", "visit_5", "visit_6", "visit_7")
)

mae <- analysis_obj$mae
df_long <- analysis_obj$data

feature_meta <- as.data.frame(
  SummarizedExperiment::rowData(experiments(mae)[["visit_all"]]),
  stringsAsFactors = FALSE
)
feature_meta$feature <- rownames(feature_meta)
feature_meta$feature_group <- if ("Group" %in% names(feature_meta)) as.character(feature_meta$Group) else "Unknown"
feature_meta$feature_group[is.na(feature_meta$feature_group) | feature_meta$feature_group == ""] <- "Unknown"
feature_meta$metabolite_label <- if ("Biomarker.name" %in% names(feature_meta)) as.character(feature_meta$Biomarker.name) else feature_meta$feature

add_feature_meta <- function(df) {
  left_join(df, feature_meta[, c("feature", "feature_group", "metabolite_label")], by = "feature")
}

strip_window_suffix <- function(x) {
  x <- sub("_[12]$", "", x)
  sub("[12]$", "", x)
}

label_exposure <- function(x) {
  gsub("_", " ", x, fixed = TRUE)
}

safe_logp <- function(p) {
  p <- as.numeric(p)
  p[p <= 0 | is.na(p)] <- NA_real_
  -log10(p)
}

early_res <- flatten_age(age_res, "early")
late_res <- flatten_age(age_res, "late")
mutual_res <- flatten_age(age_res, "mutual")
early_res_sens <- flatten_age(age_res_sens, "early")
late_res_sens <- flatten_age(age_res_sens, "late")
mutual_res_sens <- flatten_age(age_res_sens, "mutual")
mutual_early_res <- flatten_mutual_term(mutual_res, "early")
mutual_late_res <- flatten_mutual_term(mutual_res, "late")
mutual_early_res_sens <- flatten_mutual_term(mutual_res_sens, "early")
mutual_late_res_sens <- flatten_mutual_term(mutual_res_sens, "late")
all_age <- bind_rows(early_res, late_res)
long_coef_res <- if (!is.null(long_res) && all(c("term", "estimate", "q.value", "feature", "timing") %in% names(long_res))) long_res else data.frame()
long_coef_res_sens <- if (!is.null(long_res_sens) && all(c("term", "estimate", "q.value", "feature", "timing") %in% names(long_res_sens))) long_res_sens else data.frame()

pick_representative_exposure <- function(df) {
  cand <- df %>%
    filter(is.finite(q.value)) %>%
    group_by(exposure) %>%
    summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
    arrange(best_q)
  if (nrow(cand) == 0) return(NA_character_)
  cand$exposure[[1]]
}
```

## Analytic sample structure

```{r}
#| label: fig-visit-counts
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Visit-level sample sizes for the metabolomics analytical population. N indicates the number of unique participants with valid serum metabolite profiles at each study visit."
visit_counts <- data.frame(
  visit = c(4, 5, 6, 7),
  n = c(
    nrow(as.data.frame(colData(experiments(mae)[["visit_4"]]))),
    nrow(as.data.frame(colData(experiments(mae)[["visit_5"]]))),
    nrow(as.data.frame(colData(experiments(mae)[["visit_6"]]))),
    nrow(as.data.frame(colData(experiments(mae)[["visit_7"]])))
  ),
  stringsAsFactors = FALSE
)
visit_counts$child_age <- visit_label(visit_counts$visit)

visit_counts %>%
  ggplot(aes(x = child_age, y = n, fill = child_age)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.25, size = 3.5) +
  labs(
    x = "Child age",
    y = "Participants"
  ) +
  theme_minimal(base_size = 15)
```

```{r}
#| label: fig-repeated-measures
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Participation profile across visits. This plot shows the number of children who contributed to one, two, three, or all four metabolomics panels."
all_ids <- bind_rows(lapply(c(4, 5, 6, 7), function(v) {
  cd <- as.data.frame(colData(experiments(mae)[[paste0("visit_", v)]]), stringsAsFactors = FALSE)
  data.frame(StudyID = as.character(cd$StudyID), visit = v, stringsAsFactors = FALSE)
}))

profile_tbl <- all_ids %>%
  distinct(StudyID, visit) %>%
  count(StudyID, name = "n_visits") %>%
  count(n_visits, name = "n_children")

profile_tbl %>%
  ggplot(aes(x = factor(n_visits), y = n_children)) +
  geom_col(fill = "#1B6CA8", width = 0.65) +
  geom_text(aes(label = n_children), vjust = -0.25, size = 3.3) +
  labs(
    x = "Number of visits contributed",
    y = "Children"
  ) +
  theme_minimal(base_size = 15)
```

```{r}
#| label: fig-cc-boxplot
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Complete-case sample sizes for regression models. Each point represents one exposure-metabolite pair, showing the final N used after aligning diet data and confounders."
cc_tbl <- bind_rows(
  early_res %>% distinct(timing, exposure, visit, n),
  late_res %>% distinct(timing, exposure, visit, n)
)

cc_tbl %>%
  ggplot(aes(x = factor(visit), y = n, color = timing)) +
  geom_boxplot(outlier.alpha = 0.35, width = 0.6) +
  geom_jitter(width = 0.12, alpha = 0.55) +
  labs(
    x = "Child age",
    y = "Complete-case n",
    color = "Window"
  ) +
  theme_minimal(base_size = 15) +
  scale_x_discrete(labels = visit_label)
```

## Maternal exposure distributions and timing windows

```{r}
#| label: fig-exposure-dist
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 10
#| fig-cap: "Distributions of maternal dietary exposures in early and late pregnancy. Standardized E% values are shown for all nutrients included in the targeted models."
cd4 <- as.data.frame(colData(experiments(mae)[["visit_4"]]), stringsAsFactors = FALSE)

exp_early <- sort(unique(early_res$exposure))
exp_late <- sort(unique(late_res$exposure))

num_early <- exp_early[vapply(exp_early, function(v) v %in% names(cd4) && is.numeric(cd4[[v]]), logical(1))]
num_late <- exp_late[vapply(exp_late, function(v) v %in% names(cd4) && is.numeric(cd4[[v]]), logical(1))]

dist_early <- bind_rows(lapply(num_early, function(v) {
  data.frame(window = "Early", exposure = v, value = suppressWarnings(as.numeric(cd4[[v]])), stringsAsFactors = FALSE)
}))

dist_late <- bind_rows(lapply(num_late, function(v) {
  data.frame(window = "Late", exposure = v, value = suppressWarnings(as.numeric(cd4[[v]])), stringsAsFactors = FALSE)
}))

dist_all <- bind_rows(dist_early, dist_late)

if (nrow(dist_all) > 0) {
  dist_all %>%
    filter(is.finite(value)) %>%
    mutate(exposure = label_exposure(exposure)) %>%
    ggplot(aes(x = exposure, y = value, fill = window)) +
    geom_boxplot(outlier.alpha = 0.2) +
    coord_flip() +
    facet_wrap(~ window, scales = "free_y") +
    labs(
      x = NULL,
      y = "Value"
    ) +
    theme_minimal(base_size = 15)
}
```

```{r}
#| label: fig-pairing-scatter
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 10
#| fig-cap: "Intra-individual correlations between early and late pregnancy diet measurements. These pairings inform the mutual-timing models."
pair_key <- intersect(strip_window_suffix(exp_early), strip_window_suffix(exp_late))

pair_df <- bind_rows(lapply(pair_key, function(k) {
  e <- exp_early[strip_window_suffix(exp_early) == k][1]
  l <- exp_late[strip_window_suffix(exp_late) == k][1]
  if (is.na(e) || is.na(l) || !(e %in% names(cd4)) || !(l %in% names(cd4))) return(NULL)

  data.frame(
    pair = k,
    early = suppressWarnings(as.numeric(cd4[[e]])),
    late = suppressWarnings(as.numeric(cd4[[l]])),
    stringsAsFactors = FALSE
  )
})) %>%
  filter(is.finite(early), is.finite(late))

if (nrow(pair_df) > 0) {
  pair_df %>%
    mutate(pair = label_exposure(pair)) %>%
    ggplot(aes(x = early, y = late)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(alpha = 0.55, size = 1.3, color = "#1B6CA8") +
    facet_wrap(~ pair, scales = "free") +
    labs(
      x = "Early pregnancy value",
      y = "Late pregnancy value"
    ) +
    theme_minimal(base_size = 15)
}
```

## Reported result rows

```{r}
#| label: fig-testing-burden
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Count of reported result rows by analysis component. This avoids mixing feature-level association rows with exposure-level longitudinal attempts in one metric."
burden_tbl <- data.frame(
  component = c(
    "Q1 early pregnancy age-specific",
    "Q2 late pregnancy age-specific",
    "Q3 mutual timing",
    "Q4 longitudinal"
  ),
  result_rows = c(nrow(early_res), nrow(late_res), nrow(mutual_res), nrow(long_res)),
  stringsAsFactors = FALSE
)

burden_tbl %>%
  ggplot(aes(x = fct_reorder(component, result_rows), y = result_rows, fill = component)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(result_rows)), hjust = -0.1, size = 3.2) +
  coord_flip() +
  labs(x = NULL, y = "Reported result rows") +
  theme_minimal(base_size = 16)
```

## Inferential signal summary

```{r}
#| label: fig-signal-summary
#| classes: fig-wide
#| fig-cap: "Significant result-row counts and fractions across analysis components. The mutual model is split into early-term and late-term summaries because it yields two coefficient families."
signal_tbl <- bind_rows(
  early_res %>% mutate(component = paste0("Q1 early pregnancy | ", visit_label(visit))) %>% group_by(component) %>% summarise(significant_rows = sum(q.value < 0.1, na.rm = TRUE), total_rows = n(), .groups = "drop"),
  late_res %>% mutate(component = paste0("Q2 late pregnancy | ", visit_label(visit))) %>% group_by(component) %>% summarise(significant_rows = sum(q.value < 0.1, na.rm = TRUE), total_rows = n(), .groups = "drop"),
  mutual_res %>% mutate(component = paste0("Q3 mutual timing: early term | ", visit_label(visit))) %>% group_by(component) %>% summarise(significant_rows = sum(q_early < 0.1, na.rm = TRUE), total_rows = n(), .groups = "drop"),
  mutual_res %>% mutate(component = paste0("Q3 mutual timing: late term | ", visit_label(visit))) %>% group_by(component) %>% summarise(significant_rows = sum(q_late < 0.1, na.rm = TRUE), total_rows = n(), .groups = "drop"),
  if (nrow(long_coef_res) > 0) {
    long_coef_res %>%
      group_by(timing) %>%
      summarise(component = paste0("Q4 longitudinal | ", timing), significant_rows = sum(q.value < 0.1, na.rm = TRUE), total_rows = n(), .groups = "drop")
  } else {
    data.frame(component = "Q4 longitudinal | no coefficient rows", significant_rows = 0L, total_rows = 0L, stringsAsFactors = FALSE)
  }
) %>%
  mutate(significant_fraction = ifelse(total_rows > 0, significant_rows / total_rows, NA_real_))

signal_tbl %>%
  filter(!is.na(significant_fraction)) %>%
  ggplot(aes(x = fct_reorder(component, significant_fraction), y = significant_fraction, fill = component)) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = paste0(scales::percent(significant_fraction, accuracy = 0.1), "\n(", significant_rows, "/", total_rows, ")")),
    hjust = -0.05,
    size = 2.9
  ) +
  coord_flip() +
  labs(x = NULL, y = "Fraction of rows with q < 0.10") +
  theme_minimal(base_size = 15)
```

## Primary-vs-sensitivity stability

```{r}
#| label: fig-stability
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Visual summary of how much reported estimates move under expanded covariate adjustment. Reclassification uses the q < 0.10 threshold."
stability_comp <- bind_rows(
  compare_primary_sensitivity(early_res, early_res_sens, c("exposure", "feature", "visit")) %>% mutate(component = "Q1 early pregnancy age-specific"),
  compare_primary_sensitivity(late_res, late_res_sens, c("exposure", "feature", "visit")) %>% mutate(component = "Q2 late pregnancy age-specific"),
  compare_primary_sensitivity(mutual_early_res, mutual_early_res_sens, c("exposure_pair", "feature", "visit")) %>% mutate(component = "Q3 mutual timing: early term"),
  compare_primary_sensitivity(mutual_late_res, mutual_late_res_sens, c("exposure_pair", "feature", "visit")) %>% mutate(component = "Q3 mutual timing: late term"),
  compare_primary_sensitivity(long_coef_res, long_coef_res_sens, c("exposure", "feature", "term")) %>% mutate(component = "Q4 longitudinal")
)

if (nrow(stability_comp) > 0) {
  stability_tbl <- stability_comp %>%
    group_by(component) %>%
    summarise(
      pairs_compared = n(),
      median_abs_delta = median(abs_delta, na.rm = TRUE),
      q_reclassification_fraction = mean(q_reclassified, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      median_abs_delta = round(median_abs_delta, 3),
      q_reclassification_fraction = round(q_reclassification_fraction, 3)
    )

  stability_tbl %>%
    pivot_longer(cols = c(median_abs_delta, q_reclassification_fraction), names_to = "metric", values_to = "value") %>%
    mutate(metric = recode(metric,
      median_abs_delta = "Median absolute beta change",
      q_reclassification_fraction = "Fraction reclassified at q < 0.10"
    )) %>%
    ggplot(aes(x = component, y = value, fill = component)) +
    geom_col(width = 0.65, show.legend = FALSE) +
    facet_wrap(~ metric, scales = "free_y") +
    labs(x = "Component", y = "Value") +
    theme_minimal(base_size = 15) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Volcano plots (representative early and late exposures)

```{r}
#| label: fig-volcano
#| classes: fig-wide
#| fig-width: 14
#| fig-height: 10
#| fig-cap: "Illustrative volcano plots for one post hoc exemplar exposure from the early and late pregnancy model families. Exemplar exposures are selected by the smallest observed q-value within each family and are not pre-specified headline exposures."
rep_early <- pick_representative_exposure(early_res)
rep_late <- pick_representative_exposure(late_res)

volcano_df <- bind_rows(
  early_res %>% filter(exposure == rep_early) %>% mutate(window = "Early", sig = q.value < 0.1),
  late_res %>% filter(exposure == rep_late) %>% mutate(window = "Late", sig = q.value < 0.1)
) %>%
  mutate(
    logp = safe_logp(p.value),
    visit = factor(visit_label(visit), levels = c("6 Months", "12 Months", "24 Months", "5-6 Years"))
  )

volcano_df %>%
  ggplot(aes(x = estimate, y = logp, color = sig)) +
  geom_point(alpha = 0.65, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey45") +
  facet_grid(window ~ visit, scales = "free") +
  scale_color_manual(values = c("FALSE" = "grey65", "TRUE" = "firebrick")) +
  labs(
    x = "Beta",
    y = "-log10(p)",
    color = "q < 0.10"
  ) +
  theme_minimal(base_size = 16)
```

## Age-specific heatmaps

```{r}
#| label: fig-heatmaps
#| classes: fig-wide
#| fig-cap: "Heatmaps of effect sizes (Beta) across childhood visits for the top diet-metabolite associations identified in cross-sectional models."
#| fig-width: 14
#| fig-height: 10
all_age_meta <- all_age %>%
  filter(is.finite(q.value)) %>%
  add_feature_meta()

top_exposures <- all_age_meta %>%
  group_by(timing, exposure) %>%
  summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
  group_by(timing) %>%
  slice_min(best_q, n = 4, with_ties = FALSE) %>%
  ungroup()

heat_pool <- all_age_meta %>%
  inner_join(top_exposures, by = c("timing", "exposure"))

top_features <- heat_pool %>%
  group_by(timing, feature, feature_group) %>%
  summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
  group_by(timing) %>%
  slice_min(best_q, n = 20, with_ties = FALSE) %>%
  ungroup()

heat_df <- heat_pool %>%
  inner_join(top_features, by = c("timing", "feature", "feature_group")) %>%
  mutate(
    panel = paste0(timing, " | ", label_exposure(exposure)),
    feature_lab = paste0(feature_group, " | ", feature)
  )

panel_order <- heat_df %>%
  group_by(panel) %>%
  summarise(best_q = min(q.value, na.rm = TRUE), .groups = "drop") %>%
  arrange(best_q) %>%
  pull(panel)

for (panel_i in panel_order) {
  panel_df <- heat_df %>%
    filter(panel == panel_i) %>%
    mutate(feature_lab = fct_reorder(feature_lab, q.value, .desc = FALSE))

  # Extract the clean exposure and timing names for the title/subtitle
  current_exposure <- label_exposure(unique(panel_df$exposure))
  current_timing <- unique(panel_df$timing)
  current_timing_label <- ifelse(tolower(current_timing) == "early", "Early Pregnancy", "Late Pregnancy")

  p <- panel_df %>%
    ggplot(aes(x = factor(visit), y = feature_lab, fill = estimate)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
    labs(
      title = paste("Maternal Diet:", current_exposure),
      subtitle = paste("Pregnancy Window:", current_timing_label),
      x = "Child age",
      y = "Metabolite (Group | Feature)",
      fill = "Beta"
    ) +
    theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12, color = "grey35"),
      axis.text.y = element_text(size = 13)
    ) +
    scale_x_discrete(labels = visit_label)

  print(p)
}
```

## Timing comparison: early vs late window

```{r}
#| label: fig-timing-comp
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Direct comparison of early vs. late pregnancy diet effect sizes for top-ranked associations. Points on the diagonal indicate consistent effects across windows."
mutual_plot <- mutual_res %>%
  add_feature_meta() %>%
  mutate(
    best_p = pmin(p_early, p_late, na.rm = TRUE),
    visit = factor(visit_label(visit), levels = c("6 Months", "12 Months", "24 Months", "5-6 Years"))
  ) %>%
  group_by(visit) %>%
  slice_min(best_p, n = 60, with_ties = FALSE) %>%
  ungroup()

mutual_plot %>%
  ggplot(aes(x = estimate_early, y = estimate_late, color = feature_group)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
  geom_point(alpha = 0.7, size = 1.6) +
  facet_wrap(~ visit) +
  labs(
    x = "Early-window beta",
    y = "Late-window beta",
    color = "Metabolite group"
  ) +
  theme_minimal(base_size = 16)
```

## Q4 Longitudinal model status and descriptive support

The saved longitudinal output now covers all attempted longitudinal models, but these panels are still intended as cautionary context because the mixed-effects fits generated substantial warnings.

```{r}
#| label: fig-q7-yield
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Q4 longitudinal output yield by timing window. Attempted exposures come from the saved execution log; returned coefficient rows come from the saved longitudinal result object."
q7_attempt_tbl <- model_status %>%
  filter(question == "Q7") %>%
  count(timing, name = "attempted_exposures")

q7_return_tbl <- if (nrow(long_coef_res) > 0) {
  long_coef_res %>%
    count(timing, name = "returned_rows")
} else {
  q7_attempt_tbl %>% transmute(timing, returned_rows = 0L)
}

q7_yield_tbl <- full_join(q7_attempt_tbl, q7_return_tbl, by = "timing") %>%
  mutate(
    attempted_exposures = ifelse(is.na(attempted_exposures), 0L, attempted_exposures),
    returned_rows = ifelse(is.na(returned_rows), 0L, returned_rows)
  )

q7_yield_tbl %>%
  pivot_longer(cols = c(attempted_exposures, returned_rows), names_to = "quantity", values_to = "count") %>%
  mutate(quantity = recode(quantity,
    attempted_exposures = "Attempted exposures",
    returned_rows = "Returned coefficient rows"
  )) %>%
  ggplot(aes(x = timing, y = count, fill = quantity)) +
  geom_col(position = "dodge", width = 0.65) +
  geom_text(aes(label = count), position = position_dodge(width = 0.65), vjust = -0.2, size = 3.1) +
  labs(x = "Timing window", y = "Count", fill = NULL) +
  theme_minimal(base_size = 15)
```

```{r}
#| label: fig-long-status
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Status report for longitudinal mixed-effects models based on the saved execution log. Status reflects whether saved output rows were produced, not whether the fits were warning-free."
status_tbl <- model_status %>%
  filter(question == "Q7") %>%
  count(timing, status, name = "n")

status_tbl %>%
  ggplot(aes(x = timing, y = n, fill = status)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = n), vjust = -0.25, size = 3.3) +
  labs(
    x = "Timing window",
    y = "Exposure attempts",
    fill = "Status"
  ) +
  theme_minimal(base_size = 16)
```

```{r}
#| label: fig-trajectories
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Descriptive trajectories for the top age-specific association, shown by exposure tertile. This is a companion descriptive figure and is not itself a longitudinal-model result."
# Descriptive trajectories are shown as a companion view to the age-specific results, not as a replacement for longitudinal-model inference.
best_row <- all_age %>%
  filter(is.finite(q.value)) %>%
  arrange(q.value) %>%
  slice_head(n = 1)

if (nrow(best_row) > 0) {
  exp0 <- best_row$exposure[[1]]
  feat0 <- best_row$feature[[1]]

  traj_df <- df_long %>%
    filter(visit_num %in% c(4, 5, 6, 7)) %>%
    mutate(
      exposure_value = suppressWarnings(as.numeric(.data[[exp0]])),
      metabolite_value = suppressWarnings(as.numeric(.data[[feat0]])),
      exposure_tertile = make_exposure_tertiles(exposure_value)
    ) %>%
    filter(is.finite(metabolite_value), !is.na(exposure_tertile)) %>%
    group_by(visit_num, exposure_tertile) %>%
    summarise(
      mean_value = mean(metabolite_value, na.rm = TRUE),
      se = sd(metabolite_value, na.rm = TRUE) / sqrt(dplyr::n()),
      .groups = "drop"
    )

  traj_df %>%
    ggplot(aes(x = factor(visit_num), y = mean_value, color = exposure_tertile, group = exposure_tertile)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = mean_value - 1.96 * se, ymax = mean_value + 1.96 * se), width = 0.1) +
    labs(
      x = "Child age",
      y = "Mean standardized metabolite",
      color = "Exposure tertile"
    ) +
    theme_minimal(base_size = 16) +
    scale_x_discrete(labels = visit_label)
}
```
