---
title: "Exposure atlas: maternal diet"
---

Exposure-family reporting across the primary question set.

- Every maternal exposure family is shown across the early-pregnancy, late-pregnancy, and mutual-timing age-specific components when available.
- The longitudinal component is shown with coefficient-level output and execution status when those rows are available.
- Each age-specific or mutual component includes a concise summary table, an overview plot, a raw-`p` top-hits forest plot, and a searchable full result table.
- The full result tables are the exhaustive record. The overview plots are summaries.

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")
source("shared/circular_manhattan_helper.R")

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(knitr)
library(MultiAssayExperiment)
library(SummarizedExperiment)

results <- readRDS("../output/results/targeted/prenatal_diet_study2_results.rds")
age_res <- results$age_specific
long_res <- results$longitudinal
model_status <- results$model_status

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = c("visit_4", "visit_5", "visit_6", "visit_7")
)
mae <- analysis_obj$mae

feature_meta <- as.data.frame(
  SummarizedExperiment::rowData(experiments(mae)[["visit_all"]]),
  stringsAsFactors = FALSE
)
feature_meta$feature <- rownames(feature_meta)
feature_meta$feature_group <- if ("Group" %in% names(feature_meta)) as.character(feature_meta$Group) else "Unknown"

add_feature_meta <- function(df) {
  left_join(df, feature_meta[, c("feature", "feature_group")], by = "feature")
}

base_exposure_prenatal_diet <- function(x) sub("[12]$", "", x)

pretty_exposure <- function(x) {
  x <- base_exposure_prenatal_diet(x)
  x <- gsub("^M", "", x)
  x <- gsub("DietaryPatterns", "Dietary Patterns ", x)
  x <- gsub("DIINormalDiet", "Dietary Inflammatory Index ", x)
  x <- gsub("DIIDensityDiet", "Dietary Inflammatory Index Density ", x)
  x <- gsub("^IDQ$", "IDQ", x)
  x <- gsub("EPercent", "Percent ", x)
  x <- gsub("Fiball", "Fibre ", x)
  x <- gsub("CHO", "Carbohydrate", x)
  x <- gsub("Pro", "Protein", x)
  x <- gsub("Fat", "Fat", x)
  x <- gsub("SFA", "SFA", x)
  x <- gsub("MUFA", "MUFA", x)
  x <- gsub("PUFA", "PUFA", x)
  x <- gsub("FAn3", "n-3 fatty acids", x)
  x <- gsub("FAn6", "n-6 fatty acids", x)
  x <- gsub("_", " ", x, fixed = TRUE)
  x <- gsub("\\s+", " ", trimws(x))
  tools::toTitleCase(tolower(x))
}

component_label <- function(x) dplyr::case_when(
  x == "Early age-specific" ~ "Q1: Early-Pregnancy Age-Specific Models",
  x == "Late age-specific" ~ "Q2: Late-Pregnancy Age-Specific Models",
  x == "Mutual early term" ~ "Q3: Mutual Timing: Early-Pregnancy Term",
  x == "Mutual late term" ~ "Q3: Mutual Timing: Late-Pregnancy Term",
  x == "Longitudinal mixed-effects models" ~ "Q4: Longitudinal Mixed-Effects Models",
  x == "Longitudinal status" ~ "Q4: Longitudinal Output Status",
  TRUE ~ x
)

longitudinal_term_label <- function(term, exposure) dplyr::case_when(
  term == exposure ~ "Main exposure effect",
  term == paste0(exposure, ":visit_num") ~ "Exposure-by-age interaction",
  term == paste0("visit_num:", exposure) ~ "Exposure-by-age interaction",
  TRUE ~ as.character(term)
)

summary_tbl <- function(df, n_col = NULL) {
  out <- tibble(
    result_rows = nrow(df),
    significant_rows = sum(df$q.value < 0.10, na.rm = TRUE),
    significant_fraction = if (nrow(df) > 0) mean(df$q.value < 0.10, na.rm = TRUE) else NA_real_
  )
  if (!is.null(n_col) && n_col %in% names(df)) {
    vals <- suppressWarnings(as.numeric(df[[n_col]]))
    vals <- vals[is.finite(vals)]
    out$n_range <- if (length(vals) == 0) NA_character_ else paste0(min(vals), "-", max(vals))
  }
  out
}

emit_static_table <- function(tbl, caption = NULL, digits = 3) {
  if (is.null(tbl) || nrow(tbl) == 0) {
    cat("<p>No rows available.</p>\n")
    return(invisible(NULL))
  }

  tbl_out <- tbl
  num_cols <- vapply(tbl_out, is.numeric, logical(1))
  if (any(num_cols)) {
    tbl_out[num_cols] <- lapply(tbl_out[num_cols], function(x) {
      if (all(is.na(x))) return(x)
      if (all(abs(x[is.finite(x)] - round(x[is.finite(x)])) < 1e-8)) {
        round(x, 0)
      } else {
        round(x, digits)
      }
    })
  }

  html <- knitr::kable(
    tbl_out,
    format = "html",
    row.names = FALSE,
    caption = caption,
    table.attr = 'class="table table-sm table-striped table-hover atlas-table"'
  )
  cat(as.character(html), "\n")
  invisible(NULL)
}

forest_plot <- function(df, title_text) {
  top_df <- df %>%
    slice_min(p.value, n = min(20L, nrow(df)), with_ties = FALSE) %>%
    mutate(feature_lab = paste0(feature_group, " | ", feature))

  if (nrow(top_df) == 0) {
    return(NULL)
  }

  ggplot(top_df, aes(x = estimate, y = fct_reorder(feature_lab, estimate), color = feature_group)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey45") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "grey55") +
    geom_point(size = 2) +
    labs(
      x = "Standardized beta (95% CI)",
      y = NULL,
      title = title_text,
      subtitle = "Top 20 rows ranked by raw p-value",
      color = "Metabolite group"
    ) +
    theme_minimal(base_size = 14)
}

flatten_age_specific <- function(age_res, timing_label) {
  bind_rows(lapply(names(age_res), function(v) {
    age_res[[v]][[timing_label]] %>%
      mutate(
        visit = as.integer(v),
        component = paste(timing_label, "age-specific"),
        base_exposure = base_exposure_prenatal_diet(exposure)
      )
  }))
}

flatten_mutual <- function(age_res, term = c("early", "late")) {
  term <- match.arg(term)
  bind_rows(lapply(names(age_res), function(v) {
    dat <- age_res[[v]]$mutual %>% mutate(visit = as.integer(v), base_exposure = exposure_pair)
    if (term == "early") {
      dat %>%
        transmute(
          feature,
          estimate = estimate_early,
          conf.low = conf.low_early,
          conf.high = conf.high_early,
          p.value = p_early,
          q.value = q_early,
          visit,
          base_exposure,
          component = "Mutual early term",
          n = n
        )
    } else {
      dat %>%
        transmute(
          feature,
          estimate = estimate_late,
          conf.low = conf.low_late,
          conf.high = conf.high_late,
          p.value = p_late,
          q.value = q_late,
          visit,
          base_exposure,
          component = "Mutual late term",
          n = n
        )
    }
  }))
}

early_all <- flatten_age_specific(age_res, "early") %>% add_feature_meta()
late_all <- flatten_age_specific(age_res, "late") %>% add_feature_meta()
mutual_early_all <- flatten_mutual(age_res, "early") %>% add_feature_meta()
mutual_late_all <- flatten_mutual(age_res, "late") %>% add_feature_meta()
long_all <- long_res %>%
  add_ci() %>%
  mutate(
    base_exposure = base_exposure_prenatal_diet(exposure),
    component = "Longitudinal mixed-effects models",
    term_label = longitudinal_term_label(term, exposure)
  ) %>%
  add_feature_meta()

all_primary <- bind_rows(early_all, late_all, mutual_early_all, mutual_late_all) %>%
  mutate(
    conf.low = if ("conf.low" %in% names(.)) conf.low else NA_real_,
    conf.high = if ("conf.high" %in% names(.)) conf.high else NA_real_
  )

all_exposure_families <- unique(c(
  early_all$base_exposure,
  late_all$base_exposure,
  mutual_early_all$base_exposure,
  mutual_late_all$base_exposure,
  long_all$base_exposure
))
all_exposure_families <- all_exposure_families[order(pretty_exposure(all_exposure_families))]
component_order <- c("Early age-specific", "Late age-specific", "Mutual early term", "Mutual late term", "Longitudinal mixed-effects models")
```

## Exposure-by-exposure atlas

```{r}
#| label: atlas-loop
#| results: asis
#| echo: false
#| fig-width: 14
#| fig-height: 14

for (exp_base in all_exposure_families) {
  cat("\n\n### Exposure Family: ", pretty_exposure(exp_base), "{.unnumbered}\n\n", sep = "")
  cat("Raw variable family: `", exp_base, "`\n\n", sep = "")

  exp_primary <- all_primary %>% filter(base_exposure == exp_base)
  long_coef <- long_all %>% filter(base_exposure == exp_base)
  long_model_status <- model_status %>%
    mutate(base_exposure = base_exposure_prenatal_diet(exposure)) %>%
    filter(question == "Q7", base_exposure == exp_base)

  available_components <- component_order[component_order %in% unique(c(exp_primary$component, if (nrow(long_coef) > 0 || nrow(long_model_status) > 0) "Longitudinal mixed-effects models"))]

  cat("::: {.panel-tabset}\n")

  for (component_i in available_components) {
    cat("\n#### ", component_label(component_i), "\n\n", sep = "")

    if (component_i == "Longitudinal mixed-effects models") {
      yield_tbl <- tibble(
        timing = c("Early", "Late"),
        attempted_exposures = c(
          sum(long_model_status$timing == "Early", na.rm = TRUE),
          sum(long_model_status$timing == "Late", na.rm = TRUE)
        ),
        returned_rows = c(
          sum(long_coef$timing == "Early", na.rm = TRUE),
          sum(long_coef$timing == "Late", na.rm = TRUE)
        ),
        returned_term_families = c(
          dplyr::n_distinct(long_coef$term_label[long_coef$timing == "Early"]),
          dplyr::n_distinct(long_coef$term_label[long_coef$timing == "Late"])
        )
      )
      emit_static_table(yield_tbl, caption = "Q4 longitudinal output yield")

      if (nrow(long_model_status) > 0) {
        detail_tbl <- long_model_status %>%
          transmute(
            timing,
            status,
            reason,
            n_complete_min,
            n_complete_max,
            n_features_tested
          )
        cat("\n**Model status details**\n\n")
        emit_static_table(detail_tbl, caption = "Q4 longitudinal model-status details")
      }

      if (nrow(long_coef) == 0) {
        cat("\nNo coefficient-level longitudinal rows are available for this exposure family.\n\n")
        next
      }

      cat("\n**Coefficient-level longitudinal results**\n\n")
      cat("These rows come from the mixed-effects panel and are split below by pregnancy timing and coefficient family.\n\n")
      cat("::: {.panel-tabset}\n")

      long_view_tbl <- tibble(
        timing = rep(c("Early", "Late"), each = 2),
        term_label = rep(c("Main exposure effect", "Exposure-by-age interaction"), times = 2)
      ) %>%
        filter(paste(timing, term_label) %in% paste(long_coef$timing, long_coef$term_label))

      for (view_i in seq_len(nrow(long_view_tbl))) {
        timing_i <- long_view_tbl$timing[[view_i]]
        term_i <- long_view_tbl$term_label[[view_i]]
        cat("\n##### ", timing_i, " | ", term_i, "\n\n", sep = "")

        view_df <- long_coef %>% filter(timing == timing_i, term_label == term_i)
        if (nrow(view_df) == 0) {
          cat("No rows available for this longitudinal view.\n\n")
          next
        }

        summary_out <- tibble(
          result_rows = nrow(view_df),
          significant_rows = sum(view_df$q.value < 0.10, na.rm = TRUE),
          significant_fraction = if (nrow(view_df) > 0) mean(view_df$q.value < 0.10, na.rm = TRUE) else NA_real_,
          n_obs_range = paste0(min(view_df$n_obs, na.rm = TRUE), "-", max(view_df$n_obs, na.rm = TRUE)),
          n_ind_range = paste0(min(view_df$n_ind, na.rm = TRUE), "-", max(view_df$n_ind, na.rm = TRUE))
        )
        emit_static_table(summary_out, caption = paste0(component_label(component_i), " | ", timing_i, " | ", term_i, " summary"))

        cat("\n**Metabolome-wide overview plot**\n\n")
        cat("::: {.column-screen-inset-right}\n")
        plot_circular_manhattan(view_df, paste0(pretty_exposure(exp_base), " | ", component_label(component_i), " | ", timing_i, " | ", term_i))
        cat("\n:::\n\n")

        cat("**Top 20 rows by raw p-value**\n\n")
        p_forest <- forest_plot(view_df, paste0(pretty_exposure(exp_base), " | ", component_label(component_i), " | ", timing_i, " | ", term_i))
        if (!is.null(p_forest)) {
          print(p_forest)
        } else {
          cat("No rows available for the top-ranked forest plot.\n\n")
        }

        cat("**Full result table**\n\n")
        cat("This table is the exhaustive longitudinal result set for this exposure family, timing window, and coefficient family.\n\n")
        display_df <- view_df %>%
          arrange(p.value, feature) %>%
          transmute(
            feature_group,
            feature,
            estimate,
            conf.low,
            conf.high,
            p.value,
            q.value,
            n_obs,
            n_ind
          )
        emit_static_table(display_df, caption = paste0(component_label(component_i), " | ", timing_i, " | ", term_i, " full results"))
        cat("\n\n")
      }

      cat(":::\n\n")
      next
    }

    comp_df <- exp_primary %>% filter(component == component_i)
    visit_levels <- c(4L, 5L, 6L, 7L)
    visit_levels <- visit_levels[visit_levels %in% unique(comp_df$visit)]

    if (nrow(comp_df) == 0) {
      cat("No rows available for this component.\n\n")
      next
    }

    cat("::: {.panel-tabset}\n")

    for (visit_i in visit_levels) {
      cat("\n##### ", visit_label(visit_i), "\n\n", sep = "")

      visit_df <- comp_df %>% filter(visit == visit_i)
      if (nrow(visit_df) == 0) {
        cat("No rows available for this visit.\n\n")
        next
      }

      summary_out <- summary_tbl(visit_df, n_col = "n")
      emit_static_table(summary_out, caption = paste0(component_label(component_i), " | ", visit_label(visit_i), " summary"))

      cat("\n**Metabolome-wide overview plot**\n\n")
      cat("::: {.column-screen-inset-right}\n")
      plot_circular_manhattan(visit_df, paste0(pretty_exposure(exp_base), " | ", component_label(component_i), " | ", visit_label(visit_i)))
      cat("\n:::\n\n")

      cat("**Top 20 rows by raw p-value**\n\n")
      if (!all(c("conf.low", "conf.high") %in% names(visit_df)) || all(is.na(visit_df$conf.low)) || all(is.na(visit_df$conf.high))) {
        cat("Confidence intervals are not stored for this component, so the forest plot is omitted. Use the full table below for exhaustive estimates and q-values.\n\n")
      } else {
        p_forest <- forest_plot(visit_df, paste0(pretty_exposure(exp_base), " | ", component_label(component_i), " | ", visit_label(visit_i)))
        if (!is.null(p_forest)) {
          print(p_forest)
        } else {
          cat("No rows available for the top-ranked forest plot.\n\n")
        }
      }

      cat("**Full result table**\n\n")
      cat("This table is the exhaustive result set for this exposure family, component, and child age.\n\n")
      display_df <- visit_df %>%
        arrange(p.value, feature) %>%
        transmute(
          feature_group,
          feature,
          estimate,
          conf.low,
          conf.high,
          p.value,
          q.value,
          n
        )
      emit_static_table(display_df, caption = paste0(component_label(component_i), " | ", visit_label(visit_i), " full results"))
      cat("\n\n")
    }

    cat(":::\n\n")
  }

  cat(":::\n\n")
  cat("\n\n---\n\n")
}
```
