---
title: "Beta diversity"
---

## Community composition

```{r data}
# Set Cairo to enable transparency in all devices
knitr::opts_chunk$set(dev = "png", 
                      dev.args = list(type = "cairo"))
source("../funct.R")
# devtools::load_all("~/Rpackages/microbiome/miaverse/miaViz")
tse <- readRDS("../../data/tse.Rds")
```

### Phylum

```{r community-composition-phyla}
#| label: fig-figure-of-composition-phyla
#| fig-width: 10
#| fig-height: 2
#| out.width: "100%"
#| fig.align: "center"

 top_taxa <- getTop(altExp(tse, "phylum"), top = 5, assay.type = "relabundance")
 phylum_renamed <- lapply(rowData(altExp(tse, "phylum"))$phylum, function(x){
     if (x %in% top_taxa) {x} else {"Other"}})
 rowData(altExp(tse, "phylum"))$phylum_sub <- as.character(phylum_renamed)
 altExp(tse, "phylum") <- agglomerateByVariable(altExp(tse, "phylum"), 
                                    by = "rows", 
                                    f = "phylum_sub")

tse_p <- altExp(tse, "phylum")

plots <- lapply(unique(tse_p$time), function(time_point) {
  tse_filtered <- tse_p[, tse_p$time == time_point]
    
    time_plots <- lapply(sort(unique(tse_filtered$Intervention)), function(t_id) {
        tse_t <- tse_filtered[, tse_filtered$Intervention == t_id]

        # Plot abundance for the filtered tse
        plotAbundance(
            tse_t,
            assay.type = "relabundance",
            color.by = "phylum_sub",
            group = "phylum_sub",
            order.row.by = "name",
            order.col.by = "Firmicutes"
        ) +
            theme(
              axis.text.y = element_text(face = "bold", size = 12),
              axis.title.x = element_text(face = "bold", size = 14),
              axis.title.y = element_text(face = "bold", size = 14),
              legend.text = element_text(face = "bold", size = 14),
              legend.title = element_text(face = "bold", size = 14),
              plot.title = element_text(face = "bold", size = 14),
               strip.text = element_text(size = 12, face = "bold")
            ) + scale_y_continuous(label=scales::percent) +
          labs(ylab = "Abundance(%)") + facet_wrap( Intervention ~ time, nrow = 1, drop = TRUE )  })
     return(time_plots)
})

plots_flat <- unlist(plots, recursive = FALSE)

# Create the combined plot
bar_plot <- wrap_plots(plots_flat, nrow = 1, guides = "collect", axes = "collect")
bar_plot

```

### Genus

```{r community-composition-genus}
#| label: fig-figure-of-composition-genus
#| fig-width: 14
#| fig-height: 3
#| out.width: "100%"
#| fig.align: "center"
#| 
top_taxa <- getTop(altExp(tse, "genus"), top = 5, assay.type = "relabundance")
genus_renamed <- lapply(rowData(altExp(tse, "genus"))$genus, function(x){
    if (x %in% top_taxa) {x} else {"Other"}})
rowData(altExp(tse, "genus"))$genus_sub <- as.character(genus_renamed)
altExp(tse, "genus") <- agglomerateByVariable(altExp(tse, "genus"), 
                                   by = "rows", 
                                   f = "genus_sub")

tse_g <- altExp(tse, "genus")

plots <- lapply(unique(tse_g$time), function(time_point) {
  tse_filtered <- tse_g[, tse_g$time == time_point]
    
    time_plots <- lapply(sort(unique(tse_filtered$Intervention)), function(t_id) {
        tse_t <- tse_filtered[, tse_filtered$Intervention == t_id]

        # Plot abundance for the filtered tse
        plotAbundance(
            tse_t,
            assay.type = "relabundance",
            color.by = "genus_sub",
            group = "genus_sub",
            order.row.by = "name",
            order.col.by = "Blautia"
        ) +
            theme(
              axis.text.y = element_text(face = "bold", size = 12),
              axis.title.x = element_text(face = "bold", size = 14),
              axis.title.y = element_text(face = "bold", size = 14),
              legend.text = element_text(face = "bold", size = 14),
              legend.title = element_text(face = "bold", size = 14),
              plot.title = element_text(face = "bold", size = 14),
               strip.text = element_text(size = 14, face = "bold")
            ) + scale_y_continuous(label=scales::percent) +
          labs(ylab = "Abundance(%)") + facet_wrap( Intervention ~ time, nrow = 1, drop = TRUE )  })
     return(time_plots)
})

plots_flat <- unlist(plots, recursive = FALSE)

# Create the combined plot
bar_plot <- wrap_plots(plots_flat, nrow = 1, guides = "collect", axes = "collect") 
bar_plot

```

### Heatmap

```{r heatmap}
#| label: fig-figure-of-composition-heat
#| fig.width: 14
#| fig.height: 2
#| out.width: "100%"
#| fig.align: "center"
tse_heat <-  altExp(tse, "genus")
tse_heat <- transformAssay(tse_heat, assay.type = "relabundance", method = "clr", pseudocount = TRUE)

tse_heat <- transformAssay(
    tse_heat, assay.type = "clr", MARGIN = "rows", method = "standardize",
    name = "clr_z")

order_indices <- order(tse_heat$Intervention, tse_heat$time)
tse_heat <- tse_heat[, order_indices]

sechm(
            tse_heat, assayName = "clr_z", features = rownames(tse_heat),
            show_rownames = TRUE, show_colnames = FALSE, do.scale = TRUE, name = "clr_z", top_annotation="time", gaps_at=c("Intervention", "time"), show_annotation_legend = F, hmcols = c("blue", "white", "red"), cluster_rows = T, cluster_cols = F, na_col = "black", column_gap = unit(4, "mm")
)
```

## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta diversity analysis. Specifically, in this section, we employed Principal Coordinates Analysis (PCoA) to assess community dissimilarity among microbiota samples.

```{r setup, include=FALSE}
# #STEP2: PCoA
# Define variables
tax.level <- "species"
dissimilarity.index <- "bray"
variable <- "group"  
adjust.method <- "fdr"
test <- "Wilcoxon"
index <- "divergence"
```

-   **Taxonomic level:** `r tax.level`
-   **Dissimilarity index:** `r dissimilarity.index`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`

### PCoA plot

```{r pcoa-plot, fig.show="hold"}
#| label: fig-figure-of-pcoa-plot-b
#| fig-width: 8 
#| fig-height: 3 
#| fig.align: "center"
##| layout-ncol: 3
tse <- runMDS(
        tse,
        FUN = getDissimilarity,
        method = "bray",
        assay.type = "relabundance",
        name = "PCoA_BC"
    )
tse <- transformAssay(tse, assay.type = "relabundance", method = "clr", pseudocount = TRUE)
# Calculate explained variance
e <- attr(reducedDim(tse, "PCoA_BC"), "eig")
rel_eig <- e / sum(e[e > 0])
# Create the PCoA plot
p <- plotReducedDim(tse, "PCoA_BC", 
                    colour_by = "time", 
                    other_fields = c("Intervention", "group", "timepoint")) + facet_wrap(~Intervention)
# Add explained variance for axes
p1 <- p + labs(
  x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
  y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
) 
# Add ellipses for groups
p_ellipse <- p1 + 
  theme(
      legend.position = "right",
      legend.box.margin = margin(-10, 0, 0, 0)
  ) 

p_ellipse
```

### Permanova analysis

```{r}
tse <- altExp(tse, "species_prevalent")
taxa_level <- "species_prevalent"
assay <- "relabundance"

prevalence.th <- 10/100
detection.th <- 0.1/100
variable <- "Intervention" 
formula_used_within <- 'X ~ group'
```


-   Dissimilarity index: `r dissimilarity.index`
-   Group variable: `r variable`
-   Formula: `r formula_used_within`
-   assay: `r assay``
-   taxonomic level: `r taxa_level`
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

**Interpretation**

-   Homogeneity \< 0.05: The assumption of homogeneity of variance is violated (i.e., the variances are significantly different across groups).

-   Homogeneity ≥ 0.05: The assumption of homogeneity of variance is not violated (i.e., the variances are similar across groups).

```{r run_rdba_tests}
# Function to perform dbRDA and permanova analysis
perform_dbrda_permanova <- function(tse, variable, comparison) {
    tse_subset <- tse[, colData(tse)$group %in% comparison]
    valid_indices <- complete.cases(colData(tse_subset)[, variable])
    tse_rda <- tse_subset[, valid_indices]
    
    # Restrict permutations within subject blocks for paired/repeated measures
    if (any(duplicated(tse_rda$subject))) {
      perm_ctrl <- permute::how(plots = permute::Plots(strata = factor(tse_rda$subject), type = "free"))
    } else {
      perm_ctrl <- 999
    }
    
    perm_res <- getPERMANOVA(tse_rda, assay.type = "relabundance", 
                          formula = X ~ group, 
                          permutations = perm_ctrl)
    
    # Extract only the 'group' row and relevant columns
    perm_df <- as.data.frame(perm_res)
    group_row <- perm_df[rownames(perm_df) == "group", ]
    
    result <- data.frame(
        Group1 = comparison[1],
        Group2 = comparison[2],
        P_value = group_row$"permanova.Pr..F.",
        Homogeneity = signif(tail(perm_df$homogeneity.Pr..F., 1), 3)
        #P_value_adjusted = p.adjust(group_row$"permanova.Pr..F.", method = "fdr")
    )
    
    return(result)
}

# Function to apply the test over all comparisons and combine results
run_rdba_tests <- function(tse, comparisons, variable) {
    all_results <- list()
    
    for (comp in comparisons) {
        result <- perform_dbrda_permanova(tse, variable, comp)
        all_results[[length(all_results) + 1]] <- result
    }
    
    # Combine all results into a single data frame
    combined_results <- do.call(rbind, all_results)
    return(combined_results)
}
```

```{r permanova_before-after}
# For the before treatment comparisons
final_results_permanova <- run_rdba_tests(tse, comparisons_before, variable)
final_results_permanova$p.adj <- p.adjust(final_results_permanova$P_value, method = "fdr")
# For the after treatment comparisons
final_results_permanova_after <- run_rdba_tests(tse, comparisons_after, variable)
final_results_permanova_after$p.adj <- p.adjust(final_results_permanova_after$P_value, method = "fdr")

final_results_permanova[] <- lapply(final_results_permanova, function(x) gsub(".before|.after", "", x))
final_results_permanova_after[] <- lapply(final_results_permanova_after, function(x) gsub(".before|.after", "", x))

final_results_permanova$group1 <- ifelse(final_results_permanova$Group1  == "4", "Control", 
                    ifelse(final_results_permanova$Group1  == "B", "HWP", "HWP + LGG"))
final_results_permanova_after$Group2 <- ifelse(final_results_permanova_after$Group2  == "4", "Control", 
                    ifelse(final_results_permanova_after$Group2  == "B", "HWP", "HWP + LGG"))
```

```{r permanova_paired}
# For the paired treatment comparisons
final_results_permanova_paired <- run_rdba_tests(tse, comparisons_paired, variable)
final_results_permanova_paired$p.adj <- p.adjust(final_results_permanova_paired$P_value, method = "fdr")
```

#### Across-group

##### Before treatment

```{r permanova_res_before}
#| label: tbl-permanova_res_before

if (knitr::is_html_output()) {
  datatable(final_results_permanova,
            extensions = c("Buttons"),
            options = list(
              pageLength = 3,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "PERMANOVA Results - berfore treatment")
} else {
  library(kableExtra)
  final_results_permanova %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - before treatment",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

##### After treatment

```{r permanova_res_after}
#| label: tbl-permanova_res_after

if (knitr::is_html_output()) {
  datatable(final_results_permanova_after,
            extensions = c("Buttons"),
            options = list(
              pageLength = 3,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "PERMANOVA Results - after treatment")
} else {
  library(kableExtra)
  final_results_permanova_after %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - after treatment",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_after)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

#### Within-group

##### Paired treatment between timepoint(1 and 2)

```{r permanova_paired_tab}
#| label: tbl-permanova_paired_tab
Intervention <- c("Control", "HWP", "HWP + LGG")
final_results_permanova_paired <- data.frame(Intervention = Intervention, final_results_permanova_paired)
final_results_permanova_paired$Group1 <- NULL
final_results_permanova_paired$Group2 <- NULL
if (knitr::is_html_output()) {
  datatable(final_results_permanova_paired,
            extensions = c("Buttons"),
            options = list(
              pageLength = 3,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "PERMANOVA Results - Paired between timepoint(1 and 2)")
} else {
  library(kableExtra)
  final_results_permanova_paired %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - Paired",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_paired)-1))) %>%
    row_spec(0, bold = TRUE)
}

prevalence.th <- 10/100
detection.th <- 0.1/100
```

### Average abundance and 95 percentile of prevalent species

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r}
tse <- readRDS("../../data/tse.Rds")
tabs_species <- NULL 
tse_g <- altExp(tse, "species_prevalent")
for (g in unique(tse_g$Intervention)) {
    tseg <- tse_g[, tse_g$Intervention==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in na.omit(rowData(tseg)$species)) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df_before <- df[df$timepoint == "before", ]
      df_after <- df[df$timepoint == "after", ]
      
      # Sort by subject to align vectors perfectly
      df_before <- df_before[order(df_before$subject), ]
      df_after <- df_after[order(df_after$subject), ]

      tabs_species <-
        rbind(tabs_species,
        c(
          Intervention = g,
          Taxon = tax,
          mean1 = mean(df_before$signal, na.rm = TRUE),
          mean2 = mean(df_after$signal, na.rm = TRUE),
          iqr95_1 = quantile(df_before$signal, 0.95, na.rm = TRUE),
          iqr95_2 = quantile(df_after$signal, 0.95, na.rm = TRUE),
          logFC.rel = mean(df_after$signal - df_before$signal, na.rm = TRUE)
        ))
    }
}
tabs_species <- as.data.frame(tabs_species)
tabs_species$mean.rel1 <- round(100 * as.numeric(tabs_species$mean1), 3); tabs_species$mean1 <- NULL
tabs_species$mean.rel2 <- round(100 * as.numeric(tabs_species$mean2), 3); tabs_species$mean2 <- NULL
tabs_species$iqr95_1 <- round(100 * as.numeric(tabs_species$iqr95_1), 3); tabs_species$`iqr95_1.95%` <- NULL
tabs_species$iqr95_2 <- round(100 * as.numeric(tabs_species$iqr95_2), 3); tabs_species$`iqr95_2.95%` <- NULL
tabs_species$logFC.rel <- round(100 * as.numeric(tabs_species$logFC.rel), 3); 
```


```{r}
tabs_genus <- NULL 
tse_g <- altExp(tse, "genus_prevalent")
for (g in unique(tse_g$Intervention)) {
    tseg <- tse_g[, tse_g$Intervention==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in na.omit(rowData(tseg)$genus)) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df_before <- df[df$timepoint == "before", ]
      df_after <- df[df$timepoint == "after", ]
      
      # Sort by subject to align vectors perfectly
      df_before <- df_before[order(df_before$subject), ]
      df_after <- df_after[order(df_after$subject), ]

      tabs_genus <-
        rbind(tabs_genus,
        c(
          Intervention = g,
          Taxon = tax,
          mean1 = mean(df_before$signal, na.rm = TRUE),
          mean2 = mean(df_after$signal, na.rm = TRUE),
          iqr95_1 = quantile(df_before$signal, 0.95, na.rm = TRUE),
          iqr95_2 = quantile(df_after$signal, 0.95, na.rm = TRUE),
          logFC.rel = mean(df_after$signal - df_before$signal, na.rm = TRUE)
        ))
    }
}
tabs_genus <- as.data.frame(tabs_genus)
tabs_genus$mean.rel1 <- round(100 * as.numeric(tabs_genus$mean1), 3); tabs_genus$mean1 <- NULL
tabs_genus$mean.rel2 <- round(100 * as.numeric(tabs_genus$mean2), 3); tabs_genus$mean2 <- NULL
tabs_genus$iqr95_1 <- round(100 * as.numeric(tabs_genus$iqr95_1), 3); tabs_genus$`iqr95_1.95%` <- NULL
tabs_genus$iqr95_2 <- round(100 * as.numeric(tabs_genus$iqr95_2), 3); tabs_genus$`iqr95_2.95%` <- NULL
tabs_genus$logFC.rel <- round(100 * as.numeric(tabs_genus$logFC.rel), 3); 
```

```{r}
#| label: tbl-tablespecies
if (knitr::is_html_output()) {
  datatable(tabs_species,
            extensions = c("Buttons"),
            options = list(
              pageLength = 10,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "Table of average relative abundance and 95th percentile - Species")
} else {
  kable(tabs_species)
}
```

```{r}
#| label: tbl-tablegenus
if (knitr::is_html_output()) {
  datatable(tabs_genus,
            extensions = c("Buttons"),
            options = list(
              pageLength = 10,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "Table of average relative abundance and 95th percentile - Genus")
} else {
  kable(tabs_genus)
}
```

## Supervised analysis

### Significant variables

-   **Formula** assay ~ Treatment + timepoint

```{r significant variables}
tse <- readRDS("../../data/tse.Rds")
# Perform RDA
tse_4 <- tse[, tse$Treatment == "4"]
tse_B <- tse[, tse$Treatment == "B"]
tse_C <- tse[, tse$Treatment == "C"]

tse <- addRDA(
    tse,
    assay.type = "relabundance",
    formula = assay ~ Intervention + time,
    distance = dissimilarity.index,
    na.action = na.exclude)

tse_4 <- addRDA(
    tse_4,
    assay.type = "relabundance",
    formula = assay ~ time,
    distance = dissimilarity.index,
    na.action = na.exclude)
tse_B <- addRDA(
    tse_B,
    assay.type = "relabundance",
    formula = assay ~ time,
    distance = dissimilarity.index,
    na.action = na.exclude)
tse_C <- addRDA(
    tse_C,
    assay.type = "relabundance",
    formula = assay ~ time,
    distance = dissimilarity.index,
    na.action = na.exclude)

tse_list <- list(tse_4, tse_B, tse_C)

# Calculate explained variance
e_rda <- attr(reducedDim(tse, "RDA"), "eig")
rel_eig_rda <- e_rda / sum(e_rda[e_rda > 0])

```

```{r}
#| label: fig-figure-of-pRDA
plotRDA(tse, "RDA", color.by = "timepoint", add.ellipse = "fill") + 
  labs(
    x = paste("RDA 1 (", round(100 * rel_eig_rda[[1]], 1), "%)", sep = ""),
    y = paste("RDA 2 (", round(100 * rel_eig_rda[[2]], 1), "%)", sep = "")
  ) +
  labs(color = "Treatment group") +
  theme(legend.position = "top", legend.title = element_text(size = 12), legend.text = element_text(size = 10))
```



### Significant taxa



-   **RDA - Estimates for species that exhibit the largest differences between the Treatments**

```{r significant-taxa-4}
#| label: fig-figure-of-RDA-dignificant
#| fig-width: 10
#| fig-height: 3
lapply(tse_list, function(tse) {
loadings_matrix <- attr(reducedDim(tse, "RDA"), "species")
#rownames(loadings_matrix) <- rowData(tse)$genus
rownames(loadings_matrix) <- apply(rowData(tse)[, c("genus", "species")], 1, function (x) {paste(x, collapse=":")})
plotLoadings(loadings_matrix, ncomponents = 2, n = 10) + 
  labs(
    title = unique(tse$Treatment)
  )
})
```


## Data Summary

```{r data-summary}
tse
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::
