---
title: "Clinical characteristics"
---

## Setup

```{r}
#| label: clinical-characteristics-01
source("analysis_data_utils.R")
source("shared/methods_common.R")
library(dplyr)
library(tidyr)

visits <- c(4L, 5L, 6L, 7L)

visit_label <- function(v) dplyr::case_when(
  v == 4L ~ "6 Months",
  v == 5L ~ "12 Months",
  v == 6L ~ "24 Months",
  v == 7L ~ "5-6 Years",
  TRUE ~ paste("Visit", v)
)

pretty_var <- function(x) {
  x <- sub("^M", "", x)
  x <- gsub("DietaryPatterns", "Dietary Patterns ", x)
  x <- gsub("DIINormalDiet", "Dietary Inflammatory Index ", x)
  x <- gsub("DIIDensityDiet", "Dietary Inflammatory Index Density ", x)
  x <- gsub("^IDQ1$", "IDQ", x)
  x <- gsub("^IDQ2$", "IDQ", x)
  x <- gsub("EPercent", "Percent ", x)
  x <- gsub("Fiball", "Fibre ", x)
  x <- gsub("CHO", "Carbohydrate", x)
  x <- gsub("Pro", "Protein", x)
  x <- gsub("FAn3", "n-3 fatty acids", x)
  x <- gsub("FAn6", "n-6 fatty acids", x)
  x <- gsub("SFA", "SFA", x)
  x <- gsub("MUFA", "MUFA", x)
  x <- gsub("PUFA", "PUFA", x)
  x <- sub("[12]$", "", x)
  x <- gsub("_", " ", x, fixed = TRUE)
  x <- gsub("\\s+", " ", trimws(x))
  tools::toTitleCase(tolower(x))
}

exposures_early <- c(
  "MDietaryPatterns_1", "MDIINormalDiet1", "MDIIDensityDiet1", "MIDQ1",
  "MEPercentProDiet1", "MEPercentCHODiet1", "MEPercentFatDiet1",
  "MEPercentPUFADiet1", "MEPercentMUFADiet1", "MEPercentSFADiet1",
  "MEPercentFAn3Diet1", "MEPercentFAn6Diet1", "MEPercentFiballDiet1"
)
exposures_late <- c(
  "MDietaryPatterns_2", "MDIINormalDiet2", "MDIIDensityDiet2", "MIDQ2",
  "MEPercentProDiet2", "MEPercentCHODiet2", "MEPercentFatDiet2",
  "MEPercentPUFADiet2", "MEPercentMUFADiet2", "MEPercentSFADiet2",
  "MEPercentFAn3Diet2", "MEPercentFAn6Diet2", "MEPercentFiballDiet2"
)
exposure_map <- list(
  "4" = c(exposures_early, exposures_late),
  "5" = c(exposures_early, exposures_late),
  "6" = c(exposures_early, exposures_late),
  "7" = c(exposures_early, exposures_late)
)

core_covars <- c(
  "Intervention", "CGender", "MEthnicity", "MDise"
)
sensitivity_covars <- c(
  "MprepBMI", "MAge1", "MprevGDM_new", "Bfdurationm",
  "MPRSmoke", "MUniEdu", "CWeightBirth"
)

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = paste0("visit_", visits)
)

mae <- analysis_obj$mae

first_non_missing <- function(x) {
  idx <- which(!is.na(x))
  if (length(idx) == 0) return(NA)
  x[[idx[1]]]
}
```

## Objective-specific context

Maternal baseline characteristics, prenatal diet exposures, and child-age analytic panel characteristics are summarized here.

Early- and late-pregnancy diet exposures are not re-measured across child ages; age-specific rows refer to the child outcome panels linked to those baseline pregnancy measures.

## Sample counts by visit and intervention

```{r}
#| label: clinical-characteristics-02
counts_tbl <- bind_rows(lapply(visits, function(v) {
  vn <- paste0("visit_", v)
  if (!vn %in% names(experiments(mae))) return(NULL)

  cd <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[[vn]]), stringsAsFactors = FALSE)
  if (!"Intervention" %in% names(cd)) {
    return(data.frame(
      visit = v,
      Intervention = "missing",
      n = nrow(cd),
      stringsAsFactors = FALSE
    ))
  }

  cd %>%
    mutate(Intervention = as.character(Intervention)) %>%
    group_by(child_age = visit_label(v), Intervention) %>%
    summarise(n = n(), .groups = "drop")
}))

make_table(counts_tbl, caption = "Child-age panel counts by intervention")
```

## Continuous variables by visit

```{r}
#| label: clinical-characteristics-03
continuous_summary <- function(df, vars, group_var = "Intervention") {
  vars <- intersect(vars, names(df))
  if (length(vars) == 0) return(data.frame())

  num_vars <- vars[vapply(df[vars], function(x) {
    x_num <- suppressWarnings(as.numeric(x))
    sum(is.finite(x_num)) >= 20L && length(unique(x_num[is.finite(x_num)])) >= 8L
  }, logical(1))]

  if (length(num_vars) == 0) return(data.frame())

  out <- bind_rows(lapply(num_vars, function(vn) {
    x <- suppressWarnings(as.numeric(df[[vn]]))

    overall <- data.frame(
      variable = pretty_var(vn),
      variable_code = vn,
      group = "Overall",
      n = sum(is.finite(x)),
      mean = mean(x, na.rm = TRUE),
      sd = stats::sd(x, na.rm = TRUE),
      median = stats::median(x, na.rm = TRUE),
      q1 = stats::quantile(x, 0.25, na.rm = TRUE),
      q3 = stats::quantile(x, 0.75, na.rm = TRUE),
      stringsAsFactors = FALSE
    )

    by_group <- data.frame()
    if (group_var %in% names(df)) {
      by_group <- bind_rows(lapply(sort(unique(as.character(df[[group_var]]))), function(g) {
        idx <- as.character(df[[group_var]]) == g
        gx <- x[idx]
        data.frame(
          variable = pretty_var(vn),
          variable_code = vn,
          group = g,
          n = sum(is.finite(gx)),
          mean = mean(gx, na.rm = TRUE),
          sd = stats::sd(gx, na.rm = TRUE),
          median = stats::median(gx, na.rm = TRUE),
          q1 = stats::quantile(gx, 0.25, na.rm = TRUE),
          q3 = stats::quantile(gx, 0.75, na.rm = TRUE),
          stringsAsFactors = FALSE
        )
      }))
    }

    bind_rows(overall, by_group)
  }))

  out
}

cont_tbl <- bind_rows(lapply(visits, function(v) {
  vn <- paste0("visit_", v)
  if (!vn %in% names(experiments(mae))) return(NULL)
  cd <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[[vn]]), stringsAsFactors = FALSE)
  vars <- unique(c(exposure_map[[as.character(v)]], core_covars, sensitivity_covars))
  tab <- continuous_summary(cd, vars, group_var = "Intervention")
  if (nrow(tab) == 0) return(NULL)
  tab$child_age <- visit_label(v)
  tab
})) %>%
  select(child_age, variable, variable_code, group, n, mean, sd, median, q1, q3)

make_table(round_numeric(cont_tbl, 3), caption = "Continuous clinical and panel-linked exposure summaries by child age")
```

## Categorical variables by visit

```{r}
#| label: clinical-characteristics-04
categorical_summary <- function(df, vars, group_var = "Intervention") {
  vars <- intersect(vars, names(df))
  if (length(vars) == 0) return(data.frame())

  cat_vars <- vars[vapply(df[vars], function(x) {
    x_chr <- as.character(x)
    n_non_na <- sum(!is.na(x_chr) & nzchar(x_chr))
    n_non_na >= 20L && length(unique(x_chr[!is.na(x_chr) & nzchar(x_chr)])) <= 10L
  }, logical(1))]

  if (length(cat_vars) == 0) return(data.frame())

  bind_rows(lapply(cat_vars, function(vn) {
    xv <- as.character(df[[vn]])
    xv[is.na(xv) | xv == ""] <- "<missing>"

    overall <- as.data.frame(table(xv), stringsAsFactors = FALSE) %>%
      transmute(
        variable = pretty_var(vn),
        variable_code = vn,
        group = "Overall",
        level = xv,
        n = Freq,
        pct = 100 * Freq / sum(Freq)
      )

    by_group <- data.frame()
    if (group_var %in% names(df)) {
      g <- as.character(df[[group_var]])
      g[is.na(g) | g == ""] <- "<missing_group>"
      by_group <- as.data.frame(table(g, xv), stringsAsFactors = FALSE) %>%
        group_by(g) %>%
        mutate(pct = 100 * Freq / sum(Freq)) %>%
        ungroup() %>%
        transmute(
          variable = pretty_var(vn),
          variable_code = vn,
          group = g,
          level = xv,
          n = Freq,
          pct = pct
        )
    }

    bind_rows(overall, by_group)
  }))
}

cat_tbl <- bind_rows(lapply(visits, function(v) {
  vn <- paste0("visit_", v)
  if (!vn %in% names(experiments(mae))) return(NULL)
  cd <- as.data.frame(SummarizedExperiment::colData(experiments(mae)[[vn]]), stringsAsFactors = FALSE)
  vars <- unique(c(exposure_map[[as.character(v)]], core_covars, sensitivity_covars))
  tab <- categorical_summary(cd, vars, group_var = "Intervention")
  if (nrow(tab) == 0) return(NULL)
  tab$child_age <- visit_label(v)
  tab
})) %>%
  select(child_age, variable, variable_code, group, level, n, pct)

make_table(round_numeric(cat_tbl, 3), caption = "Categorical clinical and panel-linked exposure summaries by child age")
```

## Maternal baseline characteristics

```{r}
#| label: clinical-characteristics-02b
maternal_baseline_vars <- unique(c(core_covars, sensitivity_covars))
maternal_baseline_vars <- maternal_baseline_vars[maternal_baseline_vars %in% names(analysis_obj$data)]

maternal_df <- analysis_obj$data %>%
  arrange(StudyID, visit_num) %>%
  group_by(StudyID) %>%
  summarise(across(all_of(c("Intervention", maternal_baseline_vars)), first_non_missing), .groups = "drop")

maternal_cont_tbl <- continuous_summary(
  maternal_df,
  vars = maternal_baseline_vars,
  group_var = "Intervention"
) %>%
  select(variable, variable_code, group, n, mean, sd, median, q1, q3)

maternal_cat_tbl <- categorical_summary(
  maternal_df,
  vars = maternal_baseline_vars,
  group_var = "Intervention"
) %>%
  select(variable, variable_code, group, level, n, pct)

if (nrow(maternal_cont_tbl) > 0) {
  make_table(round_numeric(maternal_cont_tbl, 3), caption = "Maternal baseline continuous characteristics")
}

if (nrow(maternal_cat_tbl) > 0) {
  make_table(round_numeric(maternal_cat_tbl, 3), caption = "Maternal baseline categorical characteristics")
}
```

## Requested table of characteristics

Table 1 summarizes baseline maternal and child characteristics using one record per child/mother pair, taking the first non-missing value when a variable appears across visits. Continuous variables are mean (SD); categorical variables are n (%).

Table 2 summarizes maternal diet variables in early and late pregnancy. Continuous early-late comparisons use paired t-tests; categorical early-late comparisons use McNemar tests where applicable.

```{r}
#| label: clinical-characteristics-04b
prenatal_diet_table1_path <- "../data/prenatal_diet_characteristics_table1.csv"
prenatal_diet_table2_path <- "../data/prenatal_diet_characteristics_table2.csv"

if (file.exists(prenatal_diet_table1_path)) {
  make_table(
    read.csv(prenatal_diet_table1_path, check.names = FALSE, colClasses = "character", na.strings = character()),
    caption = "Table of characteristics: maternal and child baseline characteristics"
  )
}

if (file.exists(prenatal_diet_table2_path)) {
  make_table(
    read.csv(prenatal_diet_table2_path, check.names = FALSE, colClasses = "character", na.strings = character()),
    caption = "Table of characteristics: maternal diet characteristics, early versus late pregnancy"
  )
}
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r}
#| label: clinical-characteristics-05
#| echo: false
sessionInfo()
```
:::
