---
title: "Species-Level LimROTS Analysis"
author: ""
---

This report performs differential abundance testing of species using **LimROTS**, which combines the linear modeling/empirical Bayes framework of `limma` with the Reproducibility-Optimized Test Statistic (`ROTS`) framework.

```{r setup}
# Source helper functions and load packages
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
taxa_name <- "species"
prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Intervention" 
correction <- "ROTS permutation-based FDR"
niter_val <- 200 # For faster rendering while preserving statistical validity
```

Analysis information:

-   **Method:** LimROTS (limma + ROTS)
-   **Taxonomic level:** **`r taxa_name`**
-   **Number of Bootstrap Iterations:** `r niter_val`
-   **Detection threshold:** `r 100 * detection.th`%
-   **Prevalence threshold:** `r 100 * prevalence.th`%

---

## Across-Group Comparisons (Week 8)

We evaluate differences in species abundance between treatment groups at the post-intervention timepoint (Week 8).

```{r across-group-run}
# Get species altExp
se_species <- altExp(tse, "species_prevalent")

# Filter to Week 8 (after treatment)
se_after <- se_species[, se_species$timepoint == "after"]

# Helper to run pairwise comparison
run_pairwise_limrots <- function(se_obj, gp1, gp2) {
  # Subset to group pair
  sub_se <- se_obj[, se_obj$Intervention %in% c(gp1, gp2)]
  sub_se$Intervention <- factor(sub_se$Intervention, levels = c(gp1, gp2))
  
  # Subset the original tse to match sub_se columns
  temp_tse <- tse[, colnames(sub_se)]
  temp_tse$Intervention <- factor(temp_tse$Intervention, levels = c(gp1, gp2))
  
  res <- run_limrots_analysis(
    tse = temp_tse, 
    assay_name = "species_prevalent", 
    group_col = "Intervention", 
    formula_str = "~ 0 + Intervention",
    niter = niter_val
  )
  
  # Filter rows matching sub_se features
  res <- res[res$feature %in% rownames(sub_se), ]
  res$comparison <- paste(gp2, "vs", gp1)
  return(res)
}

# Run comparisons
res_hwp_vs_ctrl <- run_pairwise_limrots(se_after, "Control", "HWP")
res_hwp_lgg_vs_ctrl <- run_pairwise_limrots(se_after, "Control", "HWP + LGG")
res_hwp_lgg_vs_hwp <- run_pairwise_limrots(se_after, "HWP", "HWP + LGG")

# Combine results
across_results <- rbind(res_hwp_vs_ctrl, res_hwp_lgg_vs_ctrl, res_hwp_lgg_vs_hwp)
```

### Table of Findings (Week 8 Differences)

```{r across-table}
#| label: tbl-limrots-across-species
if (knitr::is_html_output()) {
  datatable(
    across_results %>% select(feature, comparison, logFC, p.value, FDR, BH.pvalue),
    extensions = c("Buttons"),
    options = list(
      pageLength = 10,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = "LimROTS Pairwise Differences at Week 8 (Species)",
    rownames = FALSE
  ) %>%
    formatRound(columns = c("logFC"), digits = 3) %>%
    formatSignif(columns = c("p.value", "FDR", "BH.pvalue"), digits = 3)
} else {
  kable(across_results)
}
```

### Volcano Plot of Week 8 Comparisons

```{r across-volcano}
#| label: fig-limrots-volcano-across-species
#| fig-width: 10
#| fig-height: 5
ggplot(across_results, aes(x = logFC, y = -log10(p.value), color = FDR < 0.25)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ comparison) +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = c("grey60", "tomato")) +
  geom_text_repel(data = subset(across_results, FDR < 0.25), aes(label = feature), size = 3, max.overlaps = 15) +
  labs(title = "Volcano Plot of Treatment Differences at Week 8", x = "Log2 Fold Change", y = "-Log10 P-value")
```

---

## Within-Group Comparisons (Baseline vs. Week 8)

We evaluate changes in species abundance over time within each of the three intervention arms.

```{r within-group-run}
# Helper to run longitudinal time comparison
run_time_limrots <- function(se_obj, group_name) {
  # Subset to group
  sub_se <- se_obj[, se_obj$Intervention == group_name]
  sub_se$time <- factor(sub_se$time, levels = c("Baseline", "Week 8"))
  
  # Build temporary TSE to run analysis on subset
  temp_tse <- tse[, tse$Intervention == group_name]
  
  res <- run_limrots_analysis(
    tse = temp_tse, 
    assay_name = "species_prevalent", 
    group_col = "time", 
    formula_str = "~ 0 + time",
    niter = niter_val
  )
  
  res$group <- group_name
  return(res)
}

# Run comparisons
res_ctrl_time <- run_time_limrots(se_species, "Control")
res_hwp_time <- run_time_limrots(se_species, "HWP")
res_hwp_lgg_time <- run_time_limrots(se_species, "HWP + LGG")

# Combine results
within_results <- rbind(res_ctrl_time, res_hwp_time, res_hwp_lgg_time)
```

### Table of Findings (Longitudinal Shifts)

```{r within-table}
#| label: tbl-limrots-within-species
if (knitr::is_html_output()) {
  datatable(
    within_results %>% select(feature, group, logFC, p.value, FDR, BH.pvalue),
    extensions = c("Buttons"),
    options = list(
      pageLength = 10,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = "LimROTS Longitudinal Changes (Week 8 vs. Baseline) within Arms (Species)",
    rownames = FALSE
  ) %>%
    formatRound(columns = c("logFC"), digits = 3) %>%
    formatSignif(columns = c("p.value", "FDR", "BH.pvalue"), digits = 3)
} else {
  kable(within_results)
}
```

### Volcano Plot of Longitudinal Changes

```{r within-volcano}
#| label: fig-limrots-volcano-within-species
#| fig-width: 10
#| fig-height: 5
ggplot(within_results, aes(x = logFC, y = -log10(p.value), color = FDR < 0.25)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ group) +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = c("grey60", "tomato")) +
  geom_text_repel(data = subset(within_results, FDR < 0.25), aes(label = feature), size = 3, max.overlaps = 15) +
  labs(title = "Volcano Plot of Longitudinal Changes (Week 8 vs Baseline)", x = "Log2 Fold Change", y = "-Log10 P-value")
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::
