---
title: "Differential abundance analysis"
author: ""
params: 
  tax.level: default
---

# Wilcoxon Tests

```{r setup}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

# Define variables and formulas
variable <- "diet" 
prevalence.th <- 10/100
detection.th <- 0.1/100
tax.level <- params$tax.level
tax.level <- "species_prevalent"

# Add CLR transformation
tse <- transformAssay(
  altExp(tse, tax.level), assay.type = "relabundance",
  method = "clr",
  pseudocount=TRUE
)

tse$diet <- factor(tse$diet, levels=c("oat", "rice"))
```

Analysis information:

-   Taxonomic level: `r tax.level`

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%
-   Number of taxonomic groups: `r nrow(tse)`

```{r tsesub}
# pick tse for time point 1 and 2 separately
tse <- tse[, tse$paired=="yes"]
tse1 <- tse[, tse$timepoint=="before"]
tse2 <- tse[, tse$timepoint=="after"]

```

```{r within_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (tp in unique(tse$duration)) {
  # Pick data for the given time point
  tse2 <- tse[, tse$duration==tp]
  pvs <- apply(assay(tse2, "clr"), 1, function (x) {kruskal.test(x ~ tse2$diet)$p.value})
  padj <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$duration)
pvals.timepoints <- pvals
```

Number of significant findings within time points (Kruskal test, Benjamini-Hochberg FDR adjustment).

```{r between_timepoint_analysis2}
print(colSums(pvals.timepoints<0.25))
```

```{r between_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (g in unique(tse$diet)) {
  # Pick data for the given Treatment for time points 1 and 2
  tse1  <- tse[, tse$diet==g & tse$timepoint=="before"]
  tse2  <- tse[, tse$diet==g & tse$timepoint=="after"]
  # Ensure that subject match (pairing ok)
  if (!all(tse1$id==tse2$id)) {stop("Subjects do not match!")}
  pvs   <- sapply(rownames(tse), function (tax) {
               wilcox.test(assay(tse1, "clr")[tax,], assay(tse2, "clr")[tax,], paired=TRUE)$p.value
	     }
	   )
  padj  <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$diet)
pvals.treatments <- pvals
```

Number of significant findings within treatments (Paired Wilcoxon test, Benjamini-Hochberg FDR adjustment).

-   **Multiple testing correction:** BH
-   **Statistical test:** Paired Wilcoxon

```{r between_timepoint_analysis3}
print(colSums(pvals.treatments<0.25))
```

Let us look more closely to the significant hits. The logFC corresponds to logFC in CLR (tp2-tp1).

```{r tsefix}
file_path1 <- file.path("../../output/daa/daa_taxa/species/all_results.tsv")
file_path2 <- file.path("../../output/daa/daa_taxa/genus/all_results.tsv")


signif.features <- names(which(pvals.treatments[, 1] < 0.25 | pvals.treatments[, 2] < 0.25 #| pvals.treatments[, 3] < 0.25 | pvals.treatments[, 4] < 0.25
                               ))

taxa_of_interest <- c(signif.features)

daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% taxa_of_interest) %>%
      select(feature,metadata, value, qval_joint) 
daa_tab2 <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% taxa_of_interest) %>%
      select(feature,metadata, value, qval_joint)

daa_tab <- rbind(daa_tab1, daa_tab2) %>%
  mutate(qval_joint = signif(qval_joint, digits = 3)) 
```

```{r hits}
tabs <- NULL 

for (g in unique(tse$diet)) {
    
  if (length(signif.features)>0) {
    tseg <- tse[, tse$diet==g] 
    dfd <- as.data.frame(colData(tseg)) 
    # for (tax in unique(rownames(tse))) {
    for (tax in signif.features) {
      dfd$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      dfd$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(dfd[, vars])
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]

      tabs <-
        rbind(tabs,
        c(
          Diet = g,
          Taxon = tax,
          mean1 = mean(subset(dfd, timepoint=="before")$signal),
          mean2 = mean(subset(dfd, timepoint=="after")$signal),
	  logFC.clr = mean(subset(dfd, timepoint=="after")$clr - subset(dfd, timepoint=="before")$clr),
          FDR   = pv 
        ))
    }
  }
}
```


```{r}
# Clean up the table
tabs <- as.data.frame(tabs)
tabs$mean.rel1 <- round(100 * as.numeric(tabs$mean1), 3); tabs$mean1 <- NULL
tabs$mean.rel2 <- round(100 * as.numeric(tabs$mean2), 3); tabs$mean2 <- NULL
tabs$FDR <- round(as.numeric(tabs$FDR), 4)
tabs$logFC.clr <- round(as.numeric(tabs$logFC.clr), 4)
tabs <- tabs %>%
  arrange(FDR, desc(logFC.clr))
```


For a complete table of results, see [complete results](https://seafile.utu.fi/f/a5d57b061c3249e193da/)

```{r}
#| label: table
if (knitr::is_html_output()) {
  datatable(tabs,
            options = list(pageLength = 10),
            caption = "Table of significant hits within groups")
} else {
  kable(tabs)
}
```

```{r}
df <- mia::meltSE(tse, add.col = T, assay.type = "relabundance")

df_not_zer0 <- df %>%
    group_by(FeatureID, duration, diet) %>%
    summarise(
        nonzero_n = sum(relabundance > 0, na.rm = TRUE),
        total_n = sum(!is.na(relabundance)),
        pct_nonzero = round(100 * nonzero_n / total_n, 1),
        .groups = "drop"
    ) %>%
    filter(nonzero_n > 0) %>%
    pivot_wider(
        names_from = c(duration, diet),
        values_from = c(nonzero_n, total_n, pct_nonzero),
        names_glue = "{diet}_{duration}_{.value}"
    )

plots_list <- list()
plots_list <- setNames(lapply(taxa_of_interest, function(taxon_name) {
    
    p1 <- plotBoxplot( tse, assay.type = "relabundance", x = "Diet",  fill.by = "Duration", pair.by = "id", features = c(taxon_name), facet.by = "rownames", add.change = T, add.significance = F, ncol = 1, label.size = 5) +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  ) + scale_y_log10(labels = scales::percent_format(scale = 1)) + 
        labs(
      x = "Diet",
      y = "Relative Abundance (%)\n(log10 scale)"
    )

    p2 <- plotBoxplot( tse, assay.type = "relabundance", x = "Duration",  fill.by = "Diet", features = c(taxon_name), facet.by = "rownames", add.significance = F, ncol = 1, label.size = 5) +
        scale_fill_viridis_d(option = "D") +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  ) + scale_y_log10(labels = scales::percent_format(scale = 1)) + 
        labs(
      x = "Duration",
      y = "Relative Abundance (%)\n(log10 scale)"
    )
    
    label_text <- generate_prevalence_label(df_not_zer0, taxon_name)
    
    p12 <- wrap_plots(p1, p2, axes = "collect") + 
      labs(caption = label_text) +
      theme(
    plot.caption = element_text(size = 10, 
                                face = "bold.italic", 
                                hjust = 1, color = "darkblue") )
    
    return(p12)
    
}
    ),
 taxa_of_interest)
```

```{r significant-hit-taxa}
#| layout-ncol: 1
#| fig-width: 10
#| fig-height: 6

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}
```

```{r volcano-tab}
tabs_all <- NULL
for (g in unique(tse$diet)) {
  # P-values
  all_features <-  names(pvals.treatments[, g])

  if (length(all_features)>0) {
    tseg <- tse[, tse$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in all_features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars], verbose = FALSE)
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]

      tabs_all <-
        rbind(tabs_all,
        c(
          Diet = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          # Standard Deviations
        # sd1 = sd(subset(df, timepoint == "before")$signal),
        # sd2 = sd(subset(df, timepoint == "after")$signal),
        # 
        # # Medians
        # median1 = median(subset(df, timepoint == "before")$signal),
        # median2 = median(subset(df, timepoint == "after")$signal),
        # 
        # # IQR (25th and 75th percentiles)
        # iqr25_1 = quantile(subset(df, timepoint == "before")$signal, 0.25),
        # iqr75_1 = quantile(subset(df, timepoint == "before")$signal, 0.75),
        # 
        # iqr25_2 = quantile(subset(df, timepoint == "after")$signal, 0.25),
        # iqr75_2 = quantile(subset(df, timepoint == "after")$signal, 0.75),
	  logFC.clr = mean(subset(df, timepoint=="after")$clr - subset(df, timepoint=="before")$clr),
          FDR   = pv 
        ))
    }
  }
}
```


```{r volcano-tab-fix}
tabs_all <- as.data.frame(tabs_all)
tabs_all$mean.rel1 <- round(100 * as.numeric(tabs_all$mean1), 1); tabs_all$mean1 <- NULL
tabs_all$mean.rel2 <- round(100 * as.numeric(tabs_all$mean2), 1); tabs_all$mean2 <- NULL
tabs_all$FDR <- round(as.numeric(tabs_all$FDR), 3)
tabs_all$logFC.clr <- round(as.numeric(tabs_all$logFC.clr), 4)
```

```{r}
tabs_all_tp <- NULL
for ( bigG in unique(tse$diet)){
    for (g in unique(tse$duration)) {
      # P-values
      all_features <-  names(pvals.timepoints[, g])
    
      if (length(all_features)>0) {
        tseg <- tse[, tse$duration==g] 
        df <- as.data.frame(colData(tseg)) 
        for (tax in all_features) {
          df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
          df$clr <- as.vector(assay(tseg, "clr")[tax,])      
          vars <- c("id", "duration", "signal")
          dfm <- melt(df[, vars], verbose = FALSE)
          colnames(dfm) <- vars
          pv <- pvals.timepoints[tax, g]
    
          tabs_all_tp <-
            rbind(tabs_all_tp,
            c(
              Diet = bigG,
              Timepoint = g,
              Taxon = tax,
              mean = mean(subset(df, duration==g)$signal),
    
            # sd = sd(subset(df, duration == g)$signal),
            # # Medians
            # median = median(subset(df, duration == g)$signal),
            # # IQR (25th and 75th percentiles)
            # iqr25 = quantile(subset(df, duration == g)$signal, 0.25),
            # iqr75 = quantile(subset(df, duration == g)$signal, 0.75),
            logFC.clr = mean(subset(df, duration==g)$clr),
              FDR   = pv 
            ))
        }
      }
    }
}
```

```{r}
tabs_all_tp <- as.data.frame(tabs_all_tp)
tabs_all_tp$mean.rel <- round(100 * as.numeric(tabs_all_tp$mean), 1); tabs_all_tp$mean <- NULL

tabs_all_tp$FDR <- round(as.numeric(tabs_all_tp$FDR), 3)
tabs_all_tp$logFC.clr <- round(as.numeric(tabs_all_tp$logFC.clr), 4)

# For a complete table of results, see [complete timepoint results]()
```



```{r}
#| label: table_tp
if (knitr::is_html_output()) {
  datatable(tabs_all_tp,
            options = list(pageLength = 10),
            caption = "Table of features within timepoints between diet groups")
} else {
  kable(tabs_all_tp)
}
```

```{r volcano-plot}
#| fig.width: 12
#| fig.height: 6

volcano_plot <- ggplot(tabs_all, aes(x = logFC.clr, y = FDR, size = mean.rel2)) +
  geom_point(aes(color = FDR < 0.25), alpha = 0.7) +  
  labs(
    title = "Volcano Plot: Effect Size vs Significance",
    x = "Log Fold Change (CLR)",
    y = "FDR"
  ) +
  scale_color_manual(values = c("blue", "red")) +  
  theme_minimal(base_size = 14) +
  facet_wrap(~ Diet) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
volcano_plot + 
  geom_text_repel(data = subset(tabs_all, FDR < 0.25),
                  aes(label = Taxon), size = 3, box.padding = 0.3)
```

```{r setup_m}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
tse <- tse[, tse$paired=="yes"]
taxa_levels <- c("species")

prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Diet" 
formula_used <- "~ diet * timepoint + (1 | id)"
correction <- "Benjamini-Hochberg FDR adjustment"

```

## Significant species from the maaslin model.

This section highlights significant taxa from maaslin3.

Analysis information:

-   Method: maaslin3
-   Taxonomic level: `r taxa_levels`
-   Group variable: `r variable`
-   Formula used: **`r formula_used`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

### N.B pvalues used in plot are from Wilcoxon tests

```{r tsesub_m}
file_path1 <- file.path("../../output/daa/daa_taxa/species/significant_results.tsv")
file_path2 <- file.path("../../output/daa/daa_taxa/species/all_results.tsv")
file_path3 <- file.path("../../output/daa/daa_taxa/oat/species/significant_results.tsv")
file_path4 <- file.path("../../output/daa/daa_taxa/rice/species/significant_results.tsv")
daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE)
daa_tab3 <- read_tsv(file_path3, col_names = TRUE, show_col_types = FALSE)
daa_tab4 <- read_tsv(file_path4, col_names = TRUE, show_col_types = FALSE)
significant_species <- unique(c(daa_tab1$feature, daa_tab3$feature, daa_tab4$feature))

daa_tab <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% significant_species) %>%
      select(feature,metadata, value, qval_joint) %>%
      mutate(qval_joint = signif(qval_joint, digits = 3)) 
```

```{r before-after_m}
tse$diet <- factor(tse$diet, levels=c("oat", "rice"))

```

```{r}

taxa_tse <- altExp(tse, "species_prevalent")
df <- mia::meltSE(taxa_tse, add.col = T, assay.type = "relabundance")

df_not_zer0 <- df %>%
    group_by(FeatureID, duration, diet) %>%
    summarise(
        nonzero_n = sum(relabundance > 0, na.rm = TRUE),
        total_n = sum(!is.na(relabundance)),
        pct_nonzero = round(100 * nonzero_n / total_n, 1),
        .groups = "drop"
    ) %>%
    filter(nonzero_n > 0) %>%
    pivot_wider(
        names_from = c(duration, diet),
        values_from = c(nonzero_n, total_n, pct_nonzero),
        names_glue = "{diet}_{duration}_{.value}"
    )

plots_list <- list()
plots_list <- setNames(lapply(significant_species, function(taxon_name) {
    
    p1 <- plotBoxplot(taxa_tse, assay.type = "relabundance", x = "diet",  fill.by = "duration", pair.by = "id", features = c(taxon_name), facet.by = "rownames", add.change = T, add.significance = T, ncol = 1, label.size = 5) +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )

    p2 <- plotBoxplot(taxa_tse, assay.type = "relabundance", x = "duration",  fill.by = "diet", features = c(taxon_name), facet.by = "rownames", add.significance = T, ncol = 1, label.size = 5) +
    scale_fill_viridis_d(option = "D") +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )
    
    label_text <- generate_prevalence_label(df_not_zer0, taxon_name)
    
    p12 <- wrap_plots(p1, p2, axes = "collect") + 
      labs(caption = label_text) +
      theme(
    plot.caption = element_text(size = 10, 
                                face = "bold.italic", 
                                hjust = 1, color = "darkblue") ) 
    
    return(p12)
    
}
    ),
 significant_species)

```



```{r}
#| label: fig-figure-of-highlights
#| layout-ncol: 1
#| fig-width: 10 
#| fig-height: 6 

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}

```
