---
title: "Beta diversity"
---

## Community composition

In the report, a suffix of 1~ Baseline and 2~ Week 6

```{r data}
# Set Cairo to enable transparency in all devices
knitr::opts_chunk$set(dev = "png", 
                      dev.args = list(type = "cairo"), fig.pos = "H")
# Source fxns
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
tse$diet <- factor(tse$diet, levels=c("oat", "rice"))
colnames(colData(tse)) <- str_replace_all(colnames(colData(tse)), 
                                          c("diet" = "Diet", "duration" = "Duration"))

colData(tse)$Time <- colData(tse)$Duration
colData(tse)$Diet <- str_to_title(colData(tse)$Diet)
colnames(colData(tse)) <- str_replace_all(colnames(colData(tse)), "Duration", "Time")
```

### Phylum

```{r}
#| label: fig-figure-of-composition-phyla
#| fig-width: 18
#| fig-height: 6
#| out.width: "100%"
#| fig.align: "center"
#| layout-nrow: 1

 top_taxa <- getTop(altExp(tse, "phylum"), top = 10, assay.type = "relabundance")
 phylum_renamed <- lapply(rowData(altExp(tse, "phylum"))$phylum, function(x){
     if (x %in% top_taxa) {x} else {"Other"}})
 rowData(altExp(tse, "phylum"))$phylum_sub <- as.character(phylum_renamed)
 altExp(tse, "phylum") <- agglomerateByVariable(altExp(tse, "phylum"), 
                                    by = "rows", 
                                    f = "phylum_sub")
colData(altExp(tse, "phylum")) <- colData(tse)
altExp(tse, "phylum")$Diet <- factor(altExp(tse, "phylum")$Diet, levels=c("Oat", "Rice"))
tse_p <- altExp(tse, "phylum")

plots <- lapply(sort(unique(tse_p$Diet)), function(diet_id) {
    tse_diet <- tse_p[, tse_p$Diet == diet_id]
    time_plots <- lapply(unique(tse_diet$Time), function(time_point) {
        tse_filtered <- tse_diet[, tse_diet$Time == time_point]
        #lab <- paste0(unique(tse_filtered$diet), " at time ", time_point)

        # Plot abundance for the filtered tse
        plotAbundance(
            tse_filtered,
            assay.type = "relabundance",
            color.by = "phylum_sub",
            group = "phylum_sub",
            order.row.by = "name",
            order.col.by = "Other"
        ) + scale_fill_discrete(name = "Phylum") +
            theme(
              axis.text.y = element_text(face = "bold", size = 12),
              axis.title.x = element_text(face = "bold", size = 14),
              axis.title.y = element_text(face = "bold", size = 14),
              legend.text = element_text(face = "bold", size = 14),
              legend.title = element_text(face = "bold", size = 14),
              plot.title = element_text(face = "bold", size = 14),
              strip.text = element_text(size = 14, face = "bold")
            ) + scale_y_continuous(label=scales::percent) +
          labs(y = "Relative abundance (%)") + facet_wrap( Diet ~ Time, nrow = 1, drop = TRUE )  })
     return(time_plots)
})

plots_flat <- unlist(plots, recursive = FALSE)

# Create the combined plot
bar_plot <- wrap_plots(plots_flat, nrow = 1, guides = "collect", axes = "collect")
bar_plot

```

```{r}
#| label: tbl-table-of-percentages-phylum
avg <- assay(tse_p, "relabundance")
phylum_names <- rowData(tse_p)$phylum_sub
sample_totals <- colSums(avg)
phylum_percent <- sweep(avg, 2, sample_totals, "/") * 100
results <- as.data.frame(phylum_percent) %>%
  mutate(Phylum = phylum_names) %>%
  pivot_longer(cols = -Phylum, names_to = "Sample", values_to = "Percentage")

summary_results <- results %>%
  group_by(Phylum) %>%
  summarize(Mean_Percentage = mean(Percentage)) %>%
  arrange(desc(Mean_Percentage)) %>% 
  mutate(Mean_Percentage = round(Mean_Percentage, 3)) 

kable(summary_results)
```

```{r}
#| label: fig-figure-of-percentages-phylum
ggplot(summary_results, aes(x = reorder(Phylum, Mean_Percentage), y = Mean_Percentage)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Horizontal bars
  labs(x = "Phylum", y = "Mean Relative Abundance (%)", 
       title = "Phylum-Level Composition") +
  theme_minimal()
```


### Genus

```{r}
#| label: fig-figure-of-composition-genus
#| fig-width: 18
#| fig-height: 6
#| out.width: "100%"
#| fig.align: "center"
#| layout-ncol: 1
# Visualization - composition bar plot
 top_taxa <- getTop(altExp(tse, "genus"), top = 10, assay.type = "relabundance")
 genus_renamed <- lapply(rowData(altExp(tse, "genus"))$genus, function(x){
     if (x %in% top_taxa) {x} else {"Other"}})
 rowData(altExp(tse, "genus"))$genus_sub <- as.character(genus_renamed)
 altExp(tse, "genus") <- agglomerateByVariable(altExp(tse, "genus"), 
                                    by = "rows", 
                                    f = "genus_sub")
colData(altExp(tse, "genus")) <- colData(tse)
altExp(tse, "genus")$Diet <- factor(altExp(tse, "genus")$Diet, levels=c("Oat", "Rice"))

tse_g <- altExp(tse, "genus")

plots <- lapply(sort(unique(tse_g$Diet)), function(diet_id) {
    tse_diet <- tse_g[, tse_g$Diet == diet_id]
    time_plots <- lapply(unique(tse_diet$Time), function(time_point) {
        tse_filtered <- tse_diet[, tse_diet$Time == time_point]
        #lab <- paste0(unique(tse_filtered$diet), " at time ", time_point)

        # Plot abundance for the filtered tse
        plotAbundance(
            tse_filtered,
            assay.type = "relabundance",
            color.by = "genus_sub",
            group = "genus_sub",
            order.row.by = "name",
            order.col.by = "Other"
        ) +  scale_fill_discrete(name = "Genus") +
            theme(
              axis.text.y = element_text(face = "bold", size = 12),
              axis.title.x = element_text(face = "bold", size = 14),
              axis.title.y = element_text(face = "bold", size = 14),
              legend.text = element_text(face = "bold", size = 14),
              legend.title = element_text(face = "bold", size = 14),
              plot.title = element_text(face = "bold", size = 14),
              strip.text = element_text(size = 14, face = "bold")
            ) + scale_y_continuous(label=scales::percent) +
          labs(y = "Relative abundance (%)") + facet_wrap( Diet ~ Time, nrow = 1, drop = TRUE )  })
     return(time_plots)
})

plots_flat <- unlist(plots, recursive = FALSE)

# Create the combined plot
bar_plot <- wrap_plots(plots_flat, nrow = 1, guides = "collect", axes = "collect") 
bar_plot
```

### Heatmap

```{r}
#| fig-width: 18
#| fig-height: 6
#| out.width: "100%"
#| fig.align: "center"
tse_heat <-  altExp(tse, "genus")
tse_heat <- transformAssay(tse_heat, assay.type = "relabundance", method = "clr", pseudocount = TRUE)

tse_heat <- transformAssay(
    tse_heat, assay.type = "clr", MARGIN = "rows", method = "standardize",
    name = "clr_z")

order_indices <- order(tse_heat$Diet_in, tse_heat$Time)
tse_heat <- tse_heat[, order_indices]

sechm(
            tse_heat, assayName = "clr_z", features = rownames(tse_heat),
            show_rownames = TRUE, show_colnames = FALSE, do.scale = TRUE, name = "clr_z", top_annotation="Time", gaps_at=c("Diet", "Time"), show_annotation_legend = F, hmcols = c("blue", "white", "red"), cluster_rows = T, cluster_cols = F, na_col = "black", column_gap = unit(4, "mm")
        )
```


## Unsupervised analysis

### Community dissimilarity

This document presents the results of a microbiota project, focusing on beta diversity analysis. Specifically, in this section, we employed Principal Coordinates Analysis (PCoA) to assess community dissimilarity among microbiota samples.

```{r setup, include=FALSE}
# #STEP2: PCoA
# Define variables
tax.level <- "species"
dissimilarity.index <- "bray"
variable <- "Diet"  
adjust.method <- "fdr"
test <- "Wilcoxon"
index <- "divergence" 
```

-   **Taxonomic level:** `r tax.level`
-   **Dissimilarity index:** `r dissimilarity.index`
-   **Group variable:** `r variable`

### PCoA plots

```{r, fig.show="hold"}
#| label: fig-figure-of-pcoa-plot-b
#| layout-ncol: 3
#| fig.width: 5
#| fig.height: 3

tse <- runMDS(
        tse,
        FUN = getDissimilarity,
        method = "bray",
        assay.type = "relabundance",
        name = "PCoA_BC"
    )
tse <- transformAssay(tse, assay.type = "relabundance", method = "clr", pseudocount = TRUE)
# Calculate explained variance
e <- attr(reducedDim(tse, "PCoA_BC"), "eig")
rel_eig <- e / sum(e[e > 0])
# Create the PCoA plot
p <- plotReducedDim(tse, "PCoA_BC",  colour_by = "group", shape_by = "Diet",
                    other_fields = c("Diet", "group", "Time", "time")) +
    scale_color_manual(
        name = "Diet × Time",
        values = c(
            "diet_1_visit_1"  = "#EEB093",
            "diet_1_visit_2"    = "#DB6022",
            "diet_2_visit_1" = "#88EDFF",
            "diet_2_visit_2"   = "#0098B2"
        ),
        labels = c(
            "diet_1_visit_1"  = "Oat Baseline",
            "diet_1_visit_2"    = "Oat Week 6",
            "diet_2_visit_1" = "Rice Baseline",
            "diet_2_visit_2"   = "Rice Week 6"
        )
    ) + theme(legend.position = "none") +
    scale_shape_manual(
        values = c(Oat = 16, Rice = 15)
    ) +
    guides(shape = "none") +
    labs(
        title = "(A) Oat vs Rice : Baseline vs Week 6",
        x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
        y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
    )

p_ellipse <- p

tse_bb <- tse[, tse$timepoint == "before"]
tse_w6 <- tse[, tse$timepoint == "after"]
tse_obw <- tse[, tse$Diet == "Oat"]
tse_rbw <- tse[, tse$Diet == "Rice"]

p1 <- plotReducedDim(tse_bb, "PCoA_BC",  colour_by = "Diet", shape_by = "Diet",
                    other_fields = c("Diet", "group", "timepoint", "time")) +
  scale_color_manual(
    values = c(Oat = "#EEB093", Rice = "#88EDFF")
  ) +
  scale_shape_manual(
    values = c(Oat = 16, Rice = 15)
  ) +
  theme(legend.position = "none") +
  labs(
    title = "(B) Baseline: Oat vs Rice",
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
  )

p2 <- plotReducedDim(tse_w6, "PCoA_BC",  colour_by = "Diet", shape_by = "Diet",
                    other_fields = c("Diet", "group", "timepoint", "time")) +
  scale_color_manual(
    values = c(Oat = "#DB6022", Rice = "#0098B2")
  ) +
  scale_shape_manual(
    values = c(Oat = 16, Rice = 15)
  ) +
  theme(legend.position = "none") +
  labs(
    title = "(C) Week 6: Oat vs Rice",
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
  )

p3 <- plotReducedDim(tse_obw, "PCoA_BC",  colour_by = "Time", shape_by = "Diet",
                    other_fields = c("Diet", "group", "timepoint", "time")) +
  scale_color_manual(
    values = c(Baseline = "#EEB093", "Week 6" = "#DB6022")
  ) +
  scale_shape_manual(
    values = c(Oat = 16, Rice = 15)
  ) +
  theme(legend.position = "none") +
  labs(
    title = "(D) Oat: Baseline vs Week 6",
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
  )

p4 <- plotReducedDim(tse_rbw, "PCoA_BC",  colour_by = "Time", shape_by = "Diet",
                    other_fields = c("Diet", "group", "timepoint", "time")) +
  scale_color_manual(
    values = c(Baseline = "#88EDFF", "Week 6" = "#0098B2")
  ) +
  scale_shape_manual(
    values = c(Oat = 16, Rice = 15)
  ) +
  theme(legend.position = "none") +
  labs(
    title = "(E) Rice: Baseline vs Week 6",
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%)", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%)", sep = "")
  )

# Define color mapping
color_mapping <- c(
  "Oat Baseline"  = "#EEB093",
  "Oat Week 6"    = "#DB6022",
  "Rice Baseline" = "#88EDFF",
  "Rice Week 6"   = "#0098B2"
)

# Create custom legend
grp_levels <- names(color_mapping)
legend_four <- ggplot(
    data.frame(x = 1, y = 1, grp = factor(grp_levels, levels = grp_levels)),
    aes(x, y, fill = grp)
) +
    geom_point(shape = 22, size = 8) +
    scale_fill_manual(
        name = "Diet × Time",
        values = color_mapping,
        breaks = grp_levels,
        labels = c("Oat Baseline", "Oat Week 6", "Rice Baseline", "Rice Week 6")
    ) +
    theme_void() + 
    theme(
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 10)
    )
legend_four_grob <- cowplot::get_legend(legend_four)

# Patch all 4 plots with legend
p_patched <- wrap_plots(p1, p2, p3, p4, nrow = 2, ncol = 2) +
    plot_layout(guides = "collect")

# Combine with custom legend using cowplot
final_plot <- cowplot::plot_grid(p_patched, legend_four_grob, rel_widths = c(4, 0.5))
p_grid <- wrap_plots(p1, p2, p3, p4, nrow = 2, ncol = 2)
final_plot <- wrap_plots(
    p_ellipse,
    p_grid,
    legend_four_grob,
    nrow = 3,
    ncol = 3,
    widths = c(2, 1),
    heights = c(1, 2)
) +
    plot_layout(design = "
    14
    23
  ")

```

```{r trajectories}
# Plot 2: Trajectories
tse <- addStepwiseDivergence(tse, group = "id",
                            time.col = "time",
                            assay.type = "relabundance")
tse_sub <- tse[, !is.na(colData(tse)$divergence)]


```

```{r run_diversity_tests}
# Calculate p-values
run_divergence_tests <- function(tse, index = "divergence") {
  # Extract values and timepoint
  values <- colData(tse)[[index]]
  timepoint <- tse$timepoint
  
  # Subset data for "after" timepoint only
  after_values <- values[timepoint == "after"]
  after_treatment <- tse$Diet[timepoint == "after"]
  
  # Create an empty results dataframe
  results <- data.frame(
    Diet = character(),
    mean = numeric(),
    p.value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through unique treatment groups
  for (g_id in unique(after_treatment)) {
    # Subset the data for current treatment group
    subset_values <- after_values[after_treatment == g_id]
    
    # Calculate the mean for "after" timepoint
    mean_after <- mean(subset_values, na.rm = TRUE)
    
    # Perform Wilcoxon test (test if the median of "after" diverges from zero)
    test_result <- wilcox.test(
      subset_values,
      mu = 0,
      exact = FALSE,
      na.action = na.omit
    )
    
    # Add results to dataframe
    results <- rbind(results, data.frame(
      Diet = g_id,
      mean = mean_after,
      p.adj = p.adjust(test_result$p.value, method = "fdr"),
      stringsAsFactors = FALSE
    )) %>%
      arrange(Diet)
  }
  
  return(results)
}

```

```{r pvalues-paired}
# Calculate statistical results
results_unpaired <- run_divergence_tests(tse_sub, index)
```

### Table of the top findings

##### Paired treatment between timepoint(1 and 2)

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **paired `r test`**

```{r}
#| label: tbl-table-of-findings-paired
if (knitr::is_html_output()) {
  datatable(
    results_unpaired, 
    options = list(
      pageLength = 2,
      dom = 'Bfrtip'
    ),
    caption = "Table of divergence paired between timepoint(1 and 2)",
    rownames = FALSE
) %>% 
    formatRound(columns = c("mean"), digits = 2) %>%
  formatSignif(columns = c("p.adj"), digits = 3)
} else {
  library(kableExtra)
  results_unpaired %>% 
    kbl(caption = "Table of divergence paired between timepoint(1 and 2)",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

## Supervised analysis

### Permanova analysis

-   Dissimilarity index: **`r dissimilarity.index`**

-   Group variable: **`r variable`**

**Interpretation**

-   Homogeneity \< 0.05: The assumption of homogeneity of variance is violated (i.e., the variances are significantly different across groups).

-   Homogeneity ≥ 0.05: The assumption of homogeneity of variance is not violated (i.e., the variances are similar across groups).

-   R2 is the proportion of variance in the distance matrix explained by diet or time.

```{r run_rdba_tests}
# Function to perform dbRDA and permanova analysis
perform_dbrda_permanova <- function(tse, variable, comparison, var, times) {
    tse_subset <- tse[, colData(tse)[, variable] %in% comparison]
    valid_indices <- complete.cases(colData(tse_subset)[, variable])
    tse_rda <- tse_subset[, valid_indices]
    tse_rda <- tse_rda[, tse_rda$timepoint == times]
    
    formula <- as.formula(paste("X ~",variable))
    
    perm_res <- getPERMANOVA(tse_rda, assay.type = "relabundance", 
                          formula = formula, 
                          permutations = 999)
    
    # Extract only the 'group' row and relevant columns
    perm_df <- as.data.frame(perm_res$permanova)
    hom_df <- as.data.frame(perm_res$homogeneity)
    # group_row <- perm_df[rownames(perm_df) == var, ]
    
    result <- data.frame(
        Term = "Diet",
        P_value = perm_df$`Pr(>F)`[1],
        R2 = signif(tail(perm_df$`R2`[1], ,), 3),
        Homogeneity = hom_df$`Pr(>F)`[1]
        # Homogeneity = signif(tail(perm_df$homogeneity.Pr..F., 1), 3)
        #P_value_adjusted = p.adjust(group_row$"permanova.Pr..F.", method = "fdr")
    )
    
    return(result)
}

# Function to apply the test over all comparisons and combine results
run_rdba_tests <- function(tse, comparisons, variable, var, times) {
    all_results <- list()
    
    for (comp in comparisons) {
        result <- perform_dbrda_permanova(tse, variable, comp, var, times)
        all_results[[length(all_results) + 1]] <- result
    }
    
    # Combine all results into a single data frame
    combined_results <- do.call(rbind, all_results)
    return(combined_results)
}
```

```{r permanova_before-after}
# For the before treatment comparisons
# final_results_permanova <- run_rdba_tests(tse, comparisons_before, variable, var = variable)
# final_results_permanova$p.adj <- p.adjust(final_results_permanova$P_value, method = "fdr")
# # For the after treatment comparisons
# final_results_permanova_after <- run_rdba_tests(tse, comparisons_after, variable, var = variable)
# final_results_permanova_after$p.adj <- p.adjust(final_results_permanova_after$P_value, method = "fdr")

final_results_permanova_tp <- run_rdba_tests(tse, comparisons_before, variable, var = "time", times = "before")
final_results_permanova_tp$p.adj <- p.adjust(final_results_permanova_tp$P_value, method = "fdr")
# For the after treatment comparisons
final_results_permanova_after_tp <- run_rdba_tests(tse, comparisons_after, variable, var = "time", times = "after")
final_results_permanova_after_tp$p.adj <- p.adjust(final_results_permanova_after_tp$P_value, method = "fdr")

final_results_permanova_tp <- final_results_permanova_tp %>%
                                    select(-P_value)

final_results_permanova_after_tp <- final_results_permanova_after_tp %>%
                                    select(-P_value)

#final_results_permanova[] <- lapply(final_results_permanova, function(x) gsub(".before|.after", "", x))
#final_results_permanova_after[] <- lapply(final_results_permanova_after, function(x) gsub(".before|.after", "", x))
```

```{r permanova_paired}
# Function to perform dbRDA and permanova analysis
perform_dbrda_permanova_pair <- function(tse, variable, comparison, var) {
    tse_subset <- tse[, colData(tse)[, variable] %in% comparison]
    valid_indices <- complete.cases(colData(tse_subset)[, variable])
    tse_rda <- tse_subset[, valid_indices]
    
    formula <- as.formula(paste("X ~", var))
    
    perm_res <- getPERMANOVA(tse_rda, assay.type = "relabundance", 
                          formula = formula, 
                          permutations = 999)
    
    # Extract only the 'group' row and relevant columns
    perm_df <- as.data.frame(perm_res)
    group_row <- perm_df[rownames(perm_df) == var, ]
    
    result <- data.frame(
        Diet = comparison,
        P_value = group_row$"permanova.Pr..F.",
        R2 = signif(tail(group_row$"permanova.R2", 1), 3),
        Homogeneity = signif(tail(perm_df$homogeneity.Pr..F., 1), 3)
        #P_value_adjusted = p.adjust(group_row$"permanova.Pr..F.", method = "fdr")
    )
    
    return(result)
}

# Function to apply the test over all comparisons and combine results
run_rdba_tests_pair <- function(tse, comparisons, variable, var) {
    all_results <- list()
    
    for (comp in comparisons) {
        result <- perform_dbrda_permanova_pair(tse, variable, comp, var)
        all_results[[length(all_results) + 1]] <- result
    }
    
    # Combine all results into a single data frame
    combined_results <- do.call(rbind, all_results)
    return(combined_results)
}
```

```{r permanova_paired_res}
# For the paired treatment comparisons
diet_groups <- sort(unique(tse$Diet))
# Calculate statistical results
  
# Run the diversity tests
 # for (comp in meal_groups) {
 #    respair <- run_rdba_tests_pair(tse, comp, variable, var= "time")
 # }

final_results_permanova_paired<- lapply(diet_groups, function(diet_id) {
    respair <- run_rdba_tests_pair(tse, comp = diet_id, variable = "Diet", var = "time")
    return(respair)
}) %>% 
  bind_rows() %>%
  arrange(Diet)
final_results_permanova_paired$p.adj <- p.adjust(final_results_permanova_paired$P_value, method = "fdr")
final_results_permanova_paired <- final_results_permanova_paired %>%
                                    select(-P_value)

#final_results_permanova_paired <- final_results_permanova_paired %>% arrange(desc(Diet))
```

#### Across-group

```{r}
#| label: tbl-permanova_res_beforetp

if (knitr::is_html_output()) {
  datatable(final_results_permanova_tp,
            options = list(pageLength = 1),
            caption = "PERMANOVA Results - variance across diets oat and rice at 
Baseline: diet explained 1.89% of the variance in microbial community composition 
(R² = 0.0189), with no evidence of dispersion differences between groups 
(homogeneity = 0.914). The effect was not statistically significant after 
correction (adjusted p = 0.099).")
} else {
  library(kableExtra)
  final_results_permanova_tp %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - before treatment - X ~ time",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_tp)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

```{r}
#| label: tbl-permanova_res_aftertp

if (knitr::is_html_output()) {
  datatable(final_results_permanova_after_tp,
            options = list(pageLength = 6),
            caption = "PERMANOVA Results - variance across diets oat and rice at 
Week 6: diet accounted for 2.03% of the variation in microbial community 
composition (R² = 0.0203), with no significant differences in dispersion 
between groups (homogeneity = 0.927). The effect was statistically significant 
after correction (adjusted p = 0.048).")
} else {
  library(kableExtra)
  final_results_permanova_after_tp %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - after treatment - X ~ time",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_after_tp)-1))) %>%
    row_spec(0, bold = TRUE)
}
```

#### Within-group

```{r}
#| label: tbl-permanova_paired_tab
if (knitr::is_html_output()) {
  datatable(final_results_permanova_paired,
            options = list(pageLength = 4),
            caption = "PERMANOVA Results - Variance of diets oat and rice between 
  Baseline and Week 6: Within-group comparisons over time showed that temporal 
  changes in microbial composition from baseline to Week 6 were minimal and not 
  statistically significant for either diet. For oat, time explained 0.63% of 
  the variance (R² = 0.0063, homogeneity = 0.669, p.adj = 1); for rice, 
  it explained 1.09% (R² = 0.0109, homogeneity = 0.835, p.adj = 1).")
} else {
  library(kableExtra)
  final_results_permanova_paired %>%
    mutate(across(where(is.numeric), ~round(., 3))) %>%
    kbl(caption = "PERMANOVA Results - Paired after intervention - X ~ time",
        booktabs = TRUE,
        align = c('l', rep('r', ncol(final_results_permanova_paired)-1))) %>%
    row_spec(0, bold = TRUE)
}

prevalence.th <- 10/100
detection.th <- 0.1/100
   
```

### Average abundance and 95 percentile of prevalent species and genus

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r}
tse <- readRDS("../../data/tse.Rds")
tabs_species <- NULL 
tse_g <- altExp(tse, "species_prevalent")
for (g in unique(tse_g$diet)) {
    tseg <- tse_g[, tse_g$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in na.omit(rowData(tseg)$species)) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars

      tabs_species <-
        rbind(tabs_species,
        c(
          Diet = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          iqr95_1 = quantile(subset(df, timepoint == "before")$signal, 0.95),
          iqr95_2 = quantile(subset(df, timepoint == "after")$signal, 0.95),
	  logFC.rel = mean(subset(df, timepoint=="after")$signal - subset(df, timepoint=="before")$signal)
        ))
    }
}
tabs_species <- as.data.frame(tabs_species)
tabs_species$mean.rel1 <- round(100 * as.numeric(tabs_species$mean1), 3); tabs_species$mean1 <- NULL
tabs_species$mean.rel2 <- round(100 * as.numeric(tabs_species$mean2), 3); tabs_species$mean2 <- NULL
tabs_species$iqr95_1 <- round(100 * as.numeric(tabs_species$iqr95_1), 3); tabs_species$`iqr95_1.95%` <- NULL
tabs_species$iqr95_2 <- round(100 * as.numeric(tabs_species$iqr95_2), 3); tabs_species$`iqr95_2.95%` <- NULL
tabs_species$logFC.rel <- round(100 * as.numeric(tabs_species$logFC.rel), 3); 
```


```{r}
tabs_genus <- NULL 
 tse_g <- altExp(tse, "genus_prevalent")
for (g in unique(tse_g$diet)) {
  tseg <- tse_g[, tse_g$diet==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in na.omit(rowData(tseg)$genus)) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      vars <- c("id", "timepoint", "signal")
      dfm <- melt(df[, vars])
      colnames(dfm) <- vars

      tabs_genus <-
        rbind(tabs_genus,
        c(
          Diet = g,
          Taxon = tax,
          mean1 = mean(subset(df, timepoint=="before")$signal),
          mean2 = mean(subset(df, timepoint=="after")$signal),
          iqr95_1 = quantile(subset(df, timepoint == "before")$signal, 0.95),
          iqr95_2 = quantile(subset(df, timepoint == "after")$signal, 0.95),
	  logFC.rel = mean(subset(df, timepoint=="after")$signal - subset(df, timepoint=="before")$signal)
        ))
    }
}
tabs_genus <- as.data.frame(tabs_genus)
tabs_genus$mean.rel1 <- round(100 * as.numeric(tabs_genus$mean1), 3); tabs_genus$mean1 <- NULL
tabs_genus$mean.rel2 <- round(100 * as.numeric(tabs_genus$mean2), 3); tabs_genus$mean2 <- NULL
tabs_genus$iqr95_1 <- round(100 * as.numeric(tabs_genus$iqr95_1), 3); tabs_genus$`iqr95_1.95%` <- NULL
tabs_genus$iqr95_2 <- round(100 * as.numeric(tabs_genus$iqr95_2), 3); tabs_genus$`iqr95_2.95%` <- NULL
tabs_genus$logFC.rel <- round(100 * as.numeric(tabs_genus$logFC.rel), 3); 
```

```{r}
#| label: tbl-tablespecies
if (knitr::is_html_output()) {
  datatable(tabs_species,
            options = list(pageLength = 10),
            caption = "Table of average relative abundance and 95th percentile - Species")
} else {
  kable(tabs_species)
}
```

```{r}
#| label: tbl-tablegenus
if (knitr::is_html_output()) {
  datatable(tabs_genus,
            options = list(pageLength = 10),
            caption = "Table of average relative abundance and 95th percentile - Genus")
} else {
  kable(tabs_genus)
}
```


### Data Summary

```{r data-summary}
tse
```
