---
title: "Data quality and assumptions"
---

## Setup

```{r}
#| label: data-quality-assumptions-01
source("analysis_data_utils.R")
library(dplyr)
library(ggplot2)

visits <- c(6L, 7L)

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = c("visit_6", "visit_7")
)

df <- analysis_obj$data
metabolite_cols <- analysis_obj$metabolite_cols

raw_df <- extract_long_from_mae(
  analysis_obj$mae,
  visits = c("visit_6", "visit_7")
)

results <- readRDS("../output/results/targeted_v2/child_diet_study1_results_v2.rds")
model_status <- results$model_status
if (!"visit_exposure" %in% names(model_status)) {
  model_status$visit_exposure <- NA_integer_
}
if (!"visit_outcome" %in% names(model_status)) {
  model_status$visit_outcome <- NA_integer_
}

visit_label <- function(v) ifelse(v == 6L, "2 Years", "5-6 Years")

pretty_var <- function(x) {
  x <- gsub("CEPercent", "Percent ", x)
  x <- gsub("CFiball", "Fibre ", x)
  x <- gsub("C_", "", x)
  x <- gsub("DietQuality", "Diet Quality ", x)
  x <- gsub("CHO", "Carbohydrate", x)
  x <- gsub("Pro", "Protein", x)
  x <- gsub("FAn3", "n-3 fatty acids", x)
  x <- gsub("FAn6", "n-6 fatty acids", x)
  x <- gsub("SFA", "SFA", x)
  x <- gsub("MUFA", "MUFA", x)
  x <- gsub("PUFA", "PUFA", x)
  x <- gsub("_", " ", x, fixed = TRUE)
  x <- gsub("\\s+", " ", trimws(x))
  tools::toTitleCase(tolower(x))
}
```

## Missingness in candidate variables

```{r}
#| label: data-quality-assumptions-02
candidate_vars <- c(
  "CEPercentProDiet6", "CEPercentCHODiet6", "CEPercentFatDiet6", "CEPercentSFADiet6",
  "CEPercentMUFADiet6", "CEPercentPUFADiet6", "CEPercentFAn3Diet6", "CEPercentFAn6Diet6",
  "CFiballDiet6", "C_DietQuality_points_6", "C_DietQuality_classes_6",
  "CEPercentProDiet7", "CEPercentCHODiet7", "CEPercentFatDiet7", "CEPercentSFADiet7",
  "CEPercentMUFADiet7", "CEPercentPUFADiet7", "CEPercentFAn3Diet7", "CEPercentFAn6Diet7",
  "CFiballDiet7", "C_DietQuality_points_7", "C_DietQuality_classes_7",
  "Intervention", "MAge1", "MprepBMI", "MPrimipara", "MUniEdu",
  "MPRSmoke", "MGDMOGTT1OR2Fi", "Premature", "MModeOfDelivery_cat",
  "CGender", "CWeightBirth", "Bfdurationm",
  "CFastingHoursBloodSample6", "CFastingHoursBloodSample7",
  "CMed6", "CMed7", "CAb6", "CAbQ7",
  "CSGA", "CMacrosomy", "CHealthy6", "CHealthy7"
)

missingness_tbl <- bind_rows(lapply(visits, function(v) {
  dv <- df %>% filter(visit_num == v)

  bind_rows(lapply(candidate_vars, function(var_nm) {
    miss_fraction <- mean(is.na(dv[[var_nm]]))
    data.frame(
      child_age = visit_label(v),
      variable = pretty_var(var_nm),
      variable_code = var_nm,
      n_missing = sum(is.na(dv[[var_nm]])),
      missing_fraction = miss_fraction,
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  arrange(desc(missing_fraction), child_age, variable)

make_table(round_numeric(missingness_tbl, 3), caption = "Variable missingness by child age")
```

## Metabolite sparsity by visit

```{r}
#| label: data-quality-assumptions-03
sparsity_summary <- bind_rows(lapply(visits, function(v) {
  dv <- raw_df %>% filter(visit_num == v)
  mat <- as.matrix(dv[, metabolite_cols, drop = FALSE])
  observed_fraction <- colMeans(!is.na(mat))

  data.frame(
    child_age = visit_label(v),
    n_features = length(observed_fraction),
    median_observed_fraction = median(observed_fraction, na.rm = TRUE),
    p10_observed_fraction = as.numeric(quantile(observed_fraction, 0.10, na.rm = TRUE, names = FALSE)),
    features_obs_lt_0_50 = sum(observed_fraction < 0.50, na.rm = TRUE),
    features_obs_lt_0_70 = sum(observed_fraction < 0.70, na.rm = TRUE),
    stringsAsFactors = FALSE
  )
}))

make_table(round_numeric(sparsity_summary, 3), caption = "Metabolite observation coverage by child age")
```

```{r}
#| label: data-quality-assumptions-04
sparse_features <- bind_rows(lapply(visits, function(v) {
  dv <- raw_df %>% filter(visit_num == v)
  mat <- as.matrix(dv[, metabolite_cols, drop = FALSE])
  observed_fraction <- colMeans(!is.na(mat))

  data.frame(
    child_age = visit_label(v),
    metabolite = names(observed_fraction),
    observed_fraction = as.numeric(observed_fraction),
    stringsAsFactors = FALSE
  )
})) %>%
  arrange(observed_fraction, child_age, metabolite) %>%
  slice_head(n = 20)

make_table(round_numeric(sparse_features, 3), caption = "Lowest-coverage metabolites across visits")
```

## Model coverage by analysis question

```{r}
#| label: data-quality-assumptions-05
coverage_tbl <- model_status %>%
  mutate(
    question = dplyr::case_when(
      question == "Q1" ~ "Concurrent at 2 Years",
      question == "Q2" ~ "Concurrent at 5-6 Years",
      question == "Q3" ~ "Prospective: 2-Year Diet to 5-6-Year Metabolites",
      question == "Q4" ~ "Repeated-Measures Across 2 Years and 5-6 Years",
      TRUE ~ question
    ),
    exposure = pretty_var(exposure),
    n_complete = ifelse(
      is.na(n_complete_min) | !is.finite(n_complete_min),
      NA_character_,
      ifelse(n_complete_min == n_complete_max, as.character(n_complete_min), paste0(n_complete_min, "-", n_complete_max))
    )
  ) %>%
  select(question, analysis, exposure, status, reason, n_complete, n_features_tested, term_family, visit_exposure, visit_outcome)

make_table(coverage_tbl, caption = "Observed model coverage and execution status from the saved analyses")
```

## Log-transform feasibility

```{r}
#| label: data-quality-assumptions-06
non_positive_tbl <- bind_rows(lapply(visits, function(v) {
  dv <- raw_df %>% filter(visit_num == v)
  mat <- as.matrix(dv[, metabolite_cols, drop = FALSE])

  non_positive_fraction <- colMeans(mat <= 0, na.rm = TRUE)
  has_non_positive <- non_positive_fraction > 0

  data.frame(
    child_age = visit_label(v),
    metabolites_with_non_positive = sum(has_non_positive, na.rm = TRUE),
    fraction_with_non_positive = mean(has_non_positive, na.rm = TRUE),
    median_non_positive_fraction = median(non_positive_fraction, na.rm = TRUE),
    p90_non_positive_fraction = as.numeric(quantile(non_positive_fraction, 0.90, na.rm = TRUE, names = FALSE)),
    stringsAsFactors = FALSE
  )
}))

make_table(round_numeric(non_positive_tbl, 3), caption = "Non-positive values before log transformation")
```

## Sample-level signal quality

```{r}
#| label: data-quality-assumptions-07
sample_qc <- df %>%
  transmute(
    StudyID,
    child_age = factor(visit_label(visit_num), levels = visit_label(visits)),
    sample_missing_fraction = rowMeans(is.na(across(all_of(metabolite_cols)))),
    median_abs_z = apply(abs(as.matrix(select(., all_of(metabolite_cols)))), 1, median, na.rm = TRUE)
  )

qc_summary <- sample_qc %>%
  group_by(child_age) %>%
  summarise(
    samples = n(),
    median_missing_fraction = median(sample_missing_fraction, na.rm = TRUE),
    p90_missing_fraction = as.numeric(quantile(sample_missing_fraction, 0.90, na.rm = TRUE, names = FALSE)),
    median_abs_z = median(median_abs_z, na.rm = TRUE),
    p90_abs_z = as.numeric(quantile(median_abs_z, 0.90, na.rm = TRUE, names = FALSE)),
    .groups = "drop"
  )

make_table(round_numeric(qc_summary, 3), caption = "Sample-level quality summary")
```

```{r}
#| label: fig-sample-missingness
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 5
ggplot(sample_qc, aes(x = sample_missing_fraction, fill = child_age)) +
  geom_histogram(binwidth = 0.02, alpha = 0.6, position = "identity") +
  theme_minimal(base_size = 11) +
  labs(
    title = "Sample missingness across child ages",
    x = "Fraction of missing metabolites per sample",
    y = "Count",
    fill = "Child age"
  )
```

```{r}
#| label: fig-sample-zscore
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 5
ggplot(sample_qc, aes(x = child_age, y = median_abs_z, color = child_age)) +
  geom_boxplot(outlier.alpha = 0.3) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Distribution of sample-level median |z| by child age",
    x = "Child age",
    y = "Median absolute z-score"
  )
```

## Visit participation profile

```{r}
#| label: data-quality-assumptions-08
visit_profile <- df %>%
  distinct(StudyID, visit_num) %>%
  count(StudyID, name = "n_visits")

profile_tbl <- visit_profile %>%
  count(n_visits, name = "n_children") %>%
  mutate(pct_children = 100 * n_children / sum(n_children)) %>%
  arrange(n_visits)

make_table(round_numeric(profile_tbl, 3), caption = "Number of visits contributed per child")
```

```{r}
#| label: fig-followup-coverage
#| classes: fig-small
#| fig-width: 7
#| fig-height: 5
ggplot(profile_tbl, aes(x = factor(n_visits), y = n_children)) +
  geom_col(fill = "#2C3E50") +
  theme_minimal(base_size = 11) +
  labs(
    title = "Child follow-up coverage",
    x = "Visits contributed",
    y = "Number of children"
  )
```

## Feature-level missingness pattern

```{r}
#| label: data-quality-assumptions-09
missing_long <- bind_rows(lapply(visits, function(v) {
  dv <- df %>% filter(visit_num == v)
  miss <- colMeans(is.na(dv[, metabolite_cols, drop = FALSE]))

  data.frame(
    child_age = visit_label(v),
    metabolite = names(miss),
    missing_fraction = as.numeric(miss),
    stringsAsFactors = FALSE
  )
}))

top_missing_features <- missing_long %>%
  group_by(metabolite) %>%
  summarise(mean_missing = mean(missing_fraction, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_missing)) %>%
  slice_head(n = 25)

make_table(round_numeric(top_missing_features, 3), caption = "Top 25 metabolites by average missingness")
```

```{r}
#| label: fig-missingness-heatmap
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 8
missing_plot_df <- missing_long %>%
  filter(metabolite %in% top_missing_features$metabolite) %>%
  mutate(
    metabolite = factor(metabolite, levels = rev(top_missing_features$metabolite))
  )

ggplot(missing_plot_df, aes(x = child_age, y = metabolite, fill = missing_fraction)) +
  geom_tile() +
  scale_fill_gradient(low = "#F7FBFF", high = "#08306B") +
  theme_minimal(base_size = 10) +
  labs(
    title = "Missingness heatmap for highest-missing metabolites",
    x = "Child age",
    y = "Metabolite",
    fill = "Missing fraction"
  )
```

## Raw vs transformed distribution check

```{r}
#| label: data-quality-assumptions-10
sample_values <- function(dat, cols, label, max_points = 50000L) {
  vals <- as.numeric(as.matrix(dat[, cols, drop = FALSE]))
  vals <- vals[is.finite(vals)]
  if (length(vals) > max_points) {
    set.seed(123)
    vals <- sample(vals, max_points)
  }
  data.frame(scale = label, value = vals, stringsAsFactors = FALSE)
}

dist_raw <- sample_values(raw_df, metabolite_cols, "raw")
dist_transformed <- sample_values(df, metabolite_cols, "log10+z")

dist_df <- bind_rows(dist_raw, dist_transformed)

dist_summary <- dist_df %>%
  group_by(scale) %>%
  summarise(
    n_values = n(),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    p01 = as.numeric(quantile(value, 0.01, na.rm = TRUE, names = FALSE)),
    median = as.numeric(quantile(value, 0.50, na.rm = TRUE, names = FALSE)),
    p99 = as.numeric(quantile(value, 0.99, na.rm = TRUE, names = FALSE)),
    .groups = "drop"
  )

make_table(round_numeric(dist_summary, 3), caption = "Distribution summary before and after transformation")
```

```{r}
#| label: fig-dist-check
#| classes: fig-medium
#| fig-width: 10
#| fig-height: 5
plot_limits <- quantile(dist_df$value, probs = c(0.01, 0.99), na.rm = TRUE)

ggplot(
  dist_df %>% filter(value >= plot_limits[[1]], value <= plot_limits[[2]]),
  aes(x = value, color = scale, fill = scale)
) +
  geom_density(alpha = 0.2) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Distribution check: raw vs log10+z",
    x = "Value (trimmed to 1st-99th percentile)",
    y = "Density",
    color = "Scale",
    fill = "Scale"
  )
```

## Global structure sanity check (PCA)

The full PCA visualization and loadings diagnostics are reported in the Data Summary page to avoid duplicated computation and repeated figures.

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r}
#| label: data-quality-assumptions-11
#| echo: false
sessionInfo()
```
:::
