---
title: "Primary results: maternal diet and child metabolites"
---

Maternal prenatal diet is related to child metabolomics from 6 months to 5-6 years. The age-specific and mutual-timing models carry the main inferential weight; the longitudinal component now returns coefficient-level output across all attempted exposures but remains more warning-prone and assumption-sensitive.

```{r}
#| label: setup
#| include: false
source("report_utils.R")
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)

# Override datatable locally to add Excel/CSV export buttons by default
datatable <- function(data, ..., options = list(), extensions = character(0)) {
  options$dom <- "Blfrtip"
  options$buttons <- c("copy", "csv", "excel")
  DT::datatable(data, ..., extensions = c(extensions, "Buttons"), options = options)
}

results <- readRDS("../output/results/targeted/prenatal_diet_study2_results.rds")
age_res <- results$age_specific
long_res <- results$longitudinal
age_res_sens <- results$age_specific_sensitivity
long_res_sens <- results$longitudinal_sensitivity
model_status <- results$model_status

early_res <- flatten_age(age_res, "early")
late_res <- flatten_age(age_res, "late")
mutual_res <- flatten_age(age_res, "mutual")
early_res_sens <- if (!is.null(age_res_sens)) flatten_age(age_res_sens, "early") else data.frame()
late_res_sens <- if (!is.null(age_res_sens)) flatten_age(age_res_sens, "late") else data.frame()
mutual_res_sens <- if (!is.null(age_res_sens)) flatten_age(age_res_sens, "mutual") else data.frame()
mutual_early_res <- flatten_mutual_term(mutual_res, "early")
mutual_late_res <- flatten_mutual_term(mutual_res, "late")
mutual_early_res_sens <- flatten_mutual_term(mutual_res_sens, "early")
mutual_late_res_sens <- flatten_mutual_term(mutual_res_sens, "late")

show_table_prenatal_diet <- function(df, title = NULL) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  
  df_display <- df %>%
    mutate(
      estimate = round(estimate, 3),
      Conf_Interval = if (all(c("conf.low", "conf.high") %in% names(.))) {
        paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]")
      } else {
        NA_character_
      },
      p.value = formatC(p.value, format = "e", digits = 2),
      q.value = formatC(q.value, format = "e", digits = 2)
    ) %>%
    mutate(age = visit_label(visit)) %>%
    select(exposure, feature, age, n, Beta = estimate, `95% CI` = Conf_Interval, P = p.value, Q = q.value)
  
  datatable(
    df_display,
    caption = title,
    filter = "top",
    options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
  )
}

clean_label <- function(x) {
  gsub("_", " ", as.character(x), fixed = TRUE)
}

signal_summary_tbl <- build_signal_summary_prenatal_diet(results)

q7_yield_tbl <- model_status %>%
  filter(question == "Q7") %>%
  count(timing, name = "attempted_exposures") %>%
  mutate(returned_rows = if (!is.null(long_res) && "timing" %in% names(long_res)) {
    vapply(timing, function(t0) sum(long_res$timing == t0, na.rm = TRUE), integer(1))
  } else {
    0L
  })
```

## Signal summary

This table shows how much result volume and FDR-significant signal the main model components contribute before the detailed tables below.

```{r}
#| label: tbl-signal-summary
datatable(
  signal_summary_tbl,
  caption = "Question-level signal summary",
  options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
)
```

## Q1 Early-pregnancy diet associations across child visits

Associations between early-pregnancy diet and child metabolites across 4 ages. Complete-case `n` across fitted models: `r sample_n_label(early_res)`.

### Nutrients {.unnumbered}

```{r}
#| label: tbl-early-nutrients
show_table_prenatal_diet(early_res %>% filter(!grepl("DietaryPatterns|DII|IDQ", exposure)), "Early Pregnancy Diet Associations - Nutrients")
```

### Diet Quality Score (Continuous) {.unnumbered}

```{r}
#| label: tbl-early-diet-quality-continuous
show_table_prenatal_diet(early_res %>% filter(grepl("IDQ|DII", exposure) & !grepl("categorized", exposure)), "Early Pregnancy Diet Associations - Diet Quality Score (Continuous)")
```

### Diet Quality Classes / Patterns (Categorical) {.unnumbered}

```{r}
#| label: tbl-early-diet-quality-categorical
show_table_prenatal_diet(early_res %>% filter(grepl("DietaryPatterns|categorized", exposure)), "Early Pregnancy Diet Associations - Diet Quality Classes / Patterns (Categorical)")
```

## Q2 Late-pregnancy diet associations across child visits

Associations between late-pregnancy diet and child metabolites across 4 ages. Complete-case `n` across fitted models: `r sample_n_label(late_res)`.

### Nutrients {.unnumbered}

```{r}
#| label: tbl-late-nutrients
show_table_prenatal_diet(late_res %>% filter(!grepl("DietaryPatterns|DII|IDQ", exposure)), "Late Pregnancy Diet Associations - Nutrients")
```

### Diet Quality Score (Continuous) {.unnumbered}

```{r}
#| label: tbl-late-diet-quality-continuous
show_table_prenatal_diet(late_res %>% filter(grepl("IDQ|DII", exposure) & !grepl("categorized", exposure)), "Late Pregnancy Diet Associations - Diet Quality Score (Continuous)")
```

### Diet Quality Classes / Patterns (Categorical) {.unnumbered}

```{r}
#| label: tbl-late-diet-quality-categorical
show_table_prenatal_diet(late_res %>% filter(grepl("DietaryPatterns|categorized", exposure)), "Late Pregnancy Diet Associations - Diet Quality Classes / Patterns (Categorical)")
```

## Q3 Timing comparison: early vs late pregnancy

Direct evaluation of whether early and late pregnancy diet windows have independent effects. Complete-case `n` across fitted models: `r sample_n_label(mutual_res)`.

The saved mutual-adjustment output reports paired early and late coefficients, 95% confidence intervals, p-values, and q-values.

```{r}
#| label: tbl-mutual
if (!is.null(mutual_res)) {
  mutual_display <- mutual_res %>%
    mutate(
      estimate_early = round(estimate_early, 3),
      early_ci = paste0("(", round(conf.low_early, 3), ", ", round(conf.high_early, 3), ")"),
      p_early = formatC(p_early, format = "e", digits = 2),
      q_early = formatC(q_early, format = "e", digits = 2),
      estimate_late = round(estimate_late, 3),
      late_ci = paste0("(", round(conf.low_late, 3), ", ", round(conf.high_late, 3), ")"),
      p_late = formatC(p_late, format = "e", digits = 2),
      q_late = formatC(q_late, format = "e", digits = 2)
    ) %>%
    transmute(
      `Exposure Pair` = clean_label(exposure_pair),
      Metabolite = feature,
      Age = visit_label(visit),
      n = n,
      `Early Beta` = estimate_early,
      `Early 95% CI` = early_ci,
      `Early P` = p_early,
      `Early Q` = q_early,
      `Late Beta` = estimate_late,
      `Late 95% CI` = late_ci,
      `Late P` = p_late,
      `Late Q` = q_late
    )
  
  datatable(
    mutual_display,
    caption = "Mutually Adjusted Timing Models",
    filter = "top",
    options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
  )
}
```

## Q4 Longitudinal diet-by-age interaction models across childhood

Evaluation of whether prenatal diet is associated with metabolite **trajectories** across childhood (Diet $\times$ Age interaction). Complete-case `n` across fitted models: `r sample_n_label(long_res, 'n_obs')`.

This is still the most warning-prone component. The saved longitudinal result object contains `r nrow(long_res)` coefficient rows from `r sum(model_status$question == "Q7", na.rm = TRUE)` attempted longitudinal exposure runs, and the executed models generated substantial warnings. This primary-results section intentionally displays the `Diet x Age` interaction coefficients, because those are the direct Q4 longitudinal estimand in this report. The full longitudinal coefficient set, including the main exposure terms from the same mixed-effects models, is available in the exposure atlas.

```{r}
#| label: tbl-q7-yield
datatable(
  q7_yield_tbl,
  caption = "Q4 longitudinal output coverage by timing window",
  options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
)
```

```{r}
#| label: tbl-interaction
if (!is.null(long_res) && "term" %in% names(long_res)) {
  interaction_display <- long_res %>%
    filter(grepl(":", term)) %>% # Focus on interaction terms
    mutate(
      estimate = round(estimate, 3),
      conf.low = round(estimate - 1.96 * std.error, 3),
      conf.high = round(estimate + 1.96 * std.error, 3),
      Conf_Interval = paste0("[", conf.low, ", ", conf.high, "]"),
      p.value = formatC(p.value, format = "e", digits = 2),
      q.value = formatC(q.value, format = "e", digits = 2)
    ) %>%
    select(exposure, feature, term, n_obs, n_ind, Beta = estimate, `95% CI` = Conf_Interval, P = p.value, Q = q.value)
  
  datatable(
    interaction_display,
    caption = "Q4 longitudinal interaction effects (primary estimand)",
    filter = "top",
    options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
  )
} else {
  message("No longitudinal interaction rows are available in the saved result object.")
}
```

## Sensitivity summary

```{r}
#| label: tbl-sensitivity-summary
age_comp <- bind_rows(
  compare_primary_sensitivity(early_res, early_res_sens, c("exposure", "feature", "visit")) %>% mutate(window = "Early"),
  compare_primary_sensitivity(late_res, late_res_sens, c("exposure", "feature", "visit")) %>% mutate(window = "Late"),
  compare_primary_sensitivity(mutual_early_res, mutual_early_res_sens, c("exposure_pair", "feature", "visit")) %>% mutate(window = "Q3 mutual timing: early term"),
  compare_primary_sensitivity(mutual_late_res, mutual_late_res_sens, c("exposure_pair", "feature", "visit")) %>% mutate(window = "Q3 mutual timing: late term")
)

if (nrow(age_comp) > 0) {
  age_summary <- age_comp %>%
    group_by(window, exposure, visit) %>%
    summarise(
      pairs_compared = n(),
      median_abs_delta = median(abs_delta, na.rm = TRUE),
      p90_abs_delta = as.numeric(quantile(abs_delta, 0.90, na.rm = TRUE, names = FALSE)),
      direction_flip_fraction = mean(direction_flip, na.rm = TRUE),
      q_reclassification_fraction = mean(q_reclassified, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      median_abs_delta = round(median_abs_delta, 3),
      p90_abs_delta = round(p90_abs_delta, 3),
      direction_flip_fraction = round(direction_flip_fraction, 3),
      q_reclassification_fraction = round(q_reclassification_fraction, 3)
    )

  datatable(
    age_summary,
    caption = "Primary-vs-sensitivity estimate stability",
    filter = "top",
    options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
  )
} else {
  message("Sensitivity comparison unavailable until sensitivity outputs are generated.")
}
```

```{r}
#| label: tbl-sensitivity-longitudinal
if (!is.null(long_res) && !is.null(long_res_sens) && nrow(long_res) > 0 && nrow(long_res_sens) > 0 && all(c("exposure", "feature", "term", "estimate") %in% names(long_res)) && all(c("exposure", "feature", "term", "estimate") %in% names(long_res_sens))) {
  long_comp <- compare_primary_sensitivity(long_res, long_res_sens, c("exposure", "feature", "term"))
  if (nrow(long_comp) > 0) {
    long_summary <- long_comp %>%
      group_by(exposure, term) %>%
      summarise(
        pairs_compared = n(),
        median_abs_delta = median(abs_delta, na.rm = TRUE),
        p90_abs_delta = as.numeric(quantile(abs_delta, 0.90, na.rm = TRUE, names = FALSE)),
        direction_flip_fraction = mean(direction_flip, na.rm = TRUE),
        q_reclassification_fraction = mean(q_reclassified, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        median_abs_delta = round(median_abs_delta, 3),
        p90_abs_delta = round(p90_abs_delta, 3),
        direction_flip_fraction = round(direction_flip_fraction, 3),
        q_reclassification_fraction = round(q_reclassification_fraction, 3)
      )

    datatable(
      long_summary,
      caption = "Longitudinal primary-vs-sensitivity estimate stability",
      filter = "top",
      options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
    )
  }
}
```

## Execution status

This status table records execution status only. `ok` means an exposure-question combination produced saved output rows; it does not mean the underlying model family was warning-free or numerically stable.

```{r}
#| label: tbl-model-status
if (!is.null(model_status) && nrow(model_status) > 0) {
  status_display <- model_status %>%
    mutate(
      n_complete = ifelse(
        is.na(n_complete_min),
        NA_character_,
        ifelse(n_complete_min == n_complete_max, as.character(n_complete_min), paste0(n_complete_min, "-", n_complete_max))
      )
    ) %>%
    select(question, analysis, timing, visit, exposure, status, reason, n_complete, n_features_tested)

  datatable(
    status_display,
    caption = "Executed model status summary",
    filter = "top",
    options = list(pageLength = 10, scrollX = FALSE, autoWidth = FALSE)
  )
}
```
