---
title: "Study overview"
---

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")
library(dplyr)
library(tidyr)

results <- readRDS("../output/results/targeted_v2/child_diet_study1_results_v2.rds")
model_status <- results$model_status

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = c("visit_6", "visit_7")
)

raw_df <- extract_long_from_mae(
  analysis_obj$mae,
  visits = c("visit_6", "visit_7")
)
```

## Research aim

Child diet at 2 years and 5-6 years is evaluated against the child plasma metabolome measured at the same ages, with an additional prospective analysis from 2-year diet to 5-6-year metabolites.

Questions:

1. concurrent associations at 2 years
2. concurrent associations at 5-6 years
3. prospective associations from 2-year diet to 5-6-year metabolites, conditioning on the same metabolite at 2 years
4. repeated-measures associations across 2 years and 5-6 years

## Analytic panels

```{r}
#| label: study-overview-01
panel_tbl <- raw_df %>%
  distinct(StudyID, visit_num) %>%
  count(visit_num, name = "n_children") %>%
  mutate(
    child_age = visit_label(visit_num),
    metabolite_features = length(analysis_obj$metabolite_cols)
  ) %>%
  select(child_age, n_children, metabolite_features)

make_table(panel_tbl, caption = "Analytic child-age panels", table_mode = "compact")
```

## Exposure timing and families

```{r}
#| label: study-overview-02
exposure_tbl <- tibble(
  component = c(
    COMPONENT_LABELS$concurrent_2y,
    COMPONENT_LABELS$concurrent_56y,
    COMPONENT_LABELS$prospective,
    COMPONENT_LABELS$rm_both
  ),
  exposure_timing = c("2-year child diet", "5-6-year child diet", "2-year child diet", "2-year child diet paired across both child ages"),
  exposure_families = c(
    paste(pretty_var(unique(model_status$exposure[model_status$question == "Q1"])), collapse = "; "),
    paste(pretty_var(unique(model_status$exposure[model_status$question == "Q2"])), collapse = "; "),
    paste(pretty_var(unique(model_status$exposure[model_status$question == "Q3"])), collapse = "; "),
    paste(pretty_var(unique(model_status$exposure[model_status$question == "Q4"])), collapse = "; ")
  )
)

make_table(exposure_tbl, caption = "Exposure timing and families", table_mode = "wide")
```

## Covariate strategy

```{r}
#| label: study-overview-03
covar_tbl <- tibble(
  analysis_component = c(
    COMPONENT_LABELS$concurrent_2y,
    COMPONENT_LABELS$concurrent_56y,
    COMPONENT_LABELS$prospective,
    COMPONENT_LABELS$rm_both
  ),
  primary_covariates = c(
    "Intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration",
    "Intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration",
    "2-year exposure; baseline metabolite at 2 years; intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration",
    "Time-varying paired diet exposure across 2 years and 5-6 years; intervention; primiparity; university education; GDM status; child sex; birthweight; breastfeeding duration"
  ),
  sensitivity_additions = c(
    "Small-for-gestational-age; macrosomy; healthy-diet score at 2 years",
    "Maternal smoking; prematurity; small-for-gestational-age; macrosomy; healthy-diet score at 5-6 years",
    "Primary prospective block plus small-for-gestational-age; macrosomy; healthy-diet score at 2 years",
    "Primary repeated-measures block plus small-for-gestational-age; macrosomy; time-varying healthy-diet score"
  )
)

make_table(covar_tbl, caption = "Primary and sensitivity covariate strategy", table_mode = "wide")
```

## Model families

```{r}
#| label: study-overview-04
q4_split <- split_q4_terms(results$q4)
model_tbl <- tibble(
  component = c(
    COMPONENT_LABELS$concurrent_2y,
    COMPONENT_LABELS$concurrent_56y,
    COMPONENT_LABELS$prospective,
    COMPONENT_LABELS$rm_main,
    COMPONENT_LABELS$rm_interaction
  ),
  model_family = c(
    "Feature-wise linear model",
    "Feature-wise linear model",
    "Feature-wise linear model with baseline metabolite adjustment",
    "Feature-wise mixed model with child-level repeated measures",
    "Feature-wise mixed model with child-level repeated measures"
  ),
  outcome = c(
    "2-year metabolite",
    "5-6-year metabolite",
    "5-6-year metabolite",
    "Metabolite measured at both child ages",
    "Metabolite measured at both child ages"
  ),
  result_rows = c(
    nrow(results$q1),
    nrow(results$q2),
    nrow(results$q3),
    nrow(q4_split$main),
    nrow(q4_split$interaction)
  )
)

make_table(model_tbl, caption = "Primary model families", table_mode = "wide")
```

## Multiple testing

False-discovery control is based on Benjamini-Hochberg adjustment across metabolites, but the adjustment scope differs by model family.

- Concurrent outputs apply BH within each exposure-specific result set.
- Prospective outputs apply BH within each exposure-specific result set.
- Repeated-measures outputs apply BH separately within each exposure and coefficient family.

## Result inventory

```{r}
#| label: study-overview-05
inventory_tbl <- build_result_inventory(results)

make_table(inventory_tbl, caption = "Returned result inventory", table_mode = "compact")
```

## Execution status

```{r}
#| label: study-overview-06
status_tbl <- model_status %>%
  mutate(
    question = dplyr::case_when(
      question == "Q1" ~ COMPONENT_LABELS$concurrent_2y,
      question == "Q2" ~ COMPONENT_LABELS$concurrent_56y,
      question == "Q3" ~ "Prospective",
      question == "Q4" ~ "Repeated-Measures",
      TRUE ~ question
    ),
    exposure = pretty_var(exposure),
    n_complete = ifelse(
      is.na(n_complete_min) | !is.finite(n_complete_min),
      NA_character_,
      ifelse(n_complete_min == n_complete_max, as.character(n_complete_min), paste0(n_complete_min, "-", n_complete_max))
    )
  ) %>%
  select(question, analysis, exposure, status, reason, n_complete, n_features_tested, term_family)

make_table(status_tbl, caption = "Saved execution-status inventory", table_mode = "wide")
```

## Interpretation notes

1. The concurrent and prospective models carry the clearest inferential weight.
2. The prospective model is conditional on the earlier metabolite value, so it should be interpreted as a follow-up association rather than a pure baseline-only prediction.
3. The repeated-measures component is included because it is part of the Study 1 question set, but it remains the most warning-prone analysis family and should be interpreted more cautiously than the linear components.
4. Detailed coefficient tables, sensitivity comparisons, and supporting figures are reported elsewhere in the site; this page is the study-level map.
