---
title: "Bacteroidetes/Firmicutes ratio"
---

```{r data}
# Set Cairo to enable transparency in all devices
knitr::opts_chunk$set(dev = "png", 
                      dev.args = list(type = "cairo"))
source("../funct.R")
# devtools::load_all("~/Rpackages/microbiome/miaverse/miaViz")
tse <- readRDS("../../data/tse.Rds")
```

```{r setup, include=FALSE}
# Define variables
tax.level     <- "Phylum"
variable      <- "group"  
adjust.method <- "FDR"
test          <- "Wilcoxon"
index         <- "bact_firm_ratio"
```

## Group-wise comparisons

-   **Taxonomic level:** `r tax.level`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`

```{r bact-firm}
# We only need phylum level tse in this report, so let's pick
# that as tse to keep things simple.
tse <- altExp(tse, "phylum")

# Calculate the Bacteroidetes/Firmicutes ratio and add to colData
# using the chosen name
ratio <- "bact_firm_ratio"
bact <- assay(tse, "relabundance")["Bacteroidetes", ]
firm <- assay(tse, "relabundance")["Firmicutes", ]
pseudo <- 1e-5
colData(tse)[[ratio]] <- (bact + pseudo) / (firm + pseudo)
```




```{r pvalues-before}
# Calculate statistical results
results_before <- run_diversity_tests(tse, comparisons_before, 
                                      variable, index = "bact_firm_ratio", 
                                      adjust.method)
results_before$p.adj <- p.adjust(results_before$p.value, method = "fdr")
# Format results for display
final_results_before <- results_before %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_before <- final_results_before %>%
  mutate(y.position = sapply(1:nrow(final_results_before), function(i) {
    vals1 <- colData(tse)[[ratio]][colData(tse)[[variable]] == group1[i]]
    vals2 <- colData(tse)[[ratio]][colData(tse)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_before)[colnames(final_results_before) == "logFC"] <- "log2FC"
```

```{r pvalues-after}
# Calculate statistical results
results_after <- run_diversity_tests(tse, comparisons_after, 
                                     variable, index = "bact_firm_ratio", 
                                     adjust.method)
results_after$p.adj <- p.adjust(results_after$p.value, method = "fdr")
# Format results for display
final_results_after <- results_after %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_after <- final_results_after %>%
  mutate(y.position = sapply(1:nrow(final_results_after), function(i) {
    vals1 <- colData(tse)[[ratio]][colData(tse)[[variable]] == group1[i]]
    vals2 <- colData(tse)[[ratio]][colData(tse)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_after)[colnames(final_results_after) == "logFC"] <- "log2FC"
```

```{r}
final_results_before$time <- rep("Baseline", nrow(final_results_before))
final_results_after$time <- rep("Week 8", nrow(final_results_after))
final_results_comb <- rbind(final_results_before, final_results_after)
final_results_comb$group1 <- gsub("\\.before$|\\.after$", "", final_results_comb$group1)
final_results_comb$group2 <- gsub("\\.before$|\\.after$", "", final_results_comb$group2)

final_results_comb$group1 <- ifelse(final_results_comb$group1  == "4", "Control", 
                    ifelse(final_results_comb$group1  == "B", "HWP", "HWP + LGG"))
final_results_comb$group2 <- ifelse(final_results_comb$group2  == "4", "Control", 
                    ifelse(final_results_comb$group2  == "B", "HWP", "HWP + LGG"))

final_results_comb <- final_results_comb %>%
  dplyr::mutate(
    y.position = case_when(
      time == "Baseline" ~ y.position + 0.1,  
      time == "Week 8" ~ y.position + 0.1
    )
  )

final_results_comb$time <- factor(final_results_comb$time, levels = c("Baseline", "Week 8"))
```


```{r pvalues-paired}
# Calculate statistical results
results_paired <- run_diversity_tests(tse, comparisons_paired, 
                                      variable, index = "bact_firm_ratio", 
                                      adjust.method, paired = TRUE)
results_paired$p.adj <- p.adjust(results_paired$p.value, method = "fdr")
# Format results for display
final_results_paired <- results_paired %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_paired <- final_results_paired %>%
  mutate(y.position = sapply(1:nrow(final_results_paired), function(i) {
    vals1 <- colData(tse)[[ratio]][colData(tse)[[variable]] == group1[i]]
    vals2 <- colData(tse)[[ratio]][colData(tse)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
Intervention <- c("Control", "HWP", "HWP + LGG")
final_results_paired <- data.frame(Intervention = Intervention, final_results_paired)
final_results_paired$group1 <- NULL
final_results_paired$group2 <- NULL
colnames(final_results_paired)[colnames(final_results_paired) == "logFC"] <- "log2FC"
```

```{r visualize-plot}
#| label: fig-figure-of-findings-before-after
#| fig-width: 6
#| fig-height: 5
#| fig.align: "center"
plotBoxplot(tse, col.var = "bact_firm_ratio", x = "time", 
            fill.by = "Intervention", add.significance = T,
            point.shape = 19) +
  scale_fill_manual(values = c("#F1F5F9", "#CBD5E1", "#64748B")) +
  scale_color_manual(values = c("black", "black", "black")) +
  theme_minimal(base_size = 15) +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12)
  )
```

## Table of the top findings

### Across-group

#### Week 0

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-before}
#| label: tbl-table-of-findings-before
final_results_before[] <- lapply(final_results_before, function(x) gsub(".before|.after", "", x))
final_results_before$group1 <- ifelse(final_results_before$group1  == "4", "Control", 
                    ifelse(final_results_before$group1  == "B", "HWP", "HWP + LGG"))
final_results_before$group2 <- ifelse(final_results_before$group2  == "4", "Control", 
                    ifelse(final_results_before$group2  == "B", "HWP", "HWP + LGG"))
  datatable(
      final_results_before %>% select(c(-y.position, -time)), 
      extensions = c("Buttons"),
      options = list(
        pageLength = 3,
        dom = 'Bfrtip',
        buttons = c("copy", "csv", "excel")
      ),
      caption = "Table of  Bacteroidetes/Firmicutes before Treatment",
      rownames = FALSE
  ) %>%
    formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
    formatSignif(columns = "p.adj", digits = 3)

```

#### Week 8

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-after}
#| label: tbl-table-of-findings-after

final_results_after[] <- lapply(final_results_after, function(x) gsub(".before|.after", "", x))
final_results_after$group1 <- ifelse(final_results_after$group1  == "4", "Control", 
                    ifelse(final_results_after$group1  == "B", "HWP", "HWP + LGG"))
final_results_after$group2 <- ifelse(final_results_after$group2  == "4", "Control", 
                    ifelse(final_results_after$group2  == "B", "HWP", "HWP + LGG"))
  datatable(
    final_results_after %>% select(c(-y.position, -time)), 
    extensions = c("Buttons"),
    options = list(
      pageLength = 3,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = "Table of Bacteroidetes/Firmicutes after Treatment",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)

```



### Within-group- Paired treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **paired `r test`**

```{r table-of-findings-paired}
#| label: tbl-table-of-findings-paired

  datatable(
    final_results_paired %>% select(-y.position), 
    extensions = c("Buttons"),
    options = list(
      pageLength = 3,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = "Table of Bacteroidetes/Firmicutes paired between timepoint (1 and 2)",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)

```

### Linear Mixed Model

Research Question 1: How does each treatment influence gut microbiota composition over time?

-   The interaction term (Treatment:Timepoint) tests whether the effect of time on B/F ratio depends on the treatment group.

-   Post-hoc analysis within each treatment group identifies specific changes in B/F ratio over time.

Research Question 2: Are there significant differences between treatments?

-   The main effect of Treatment tests for differences in B/F ratio between treatment groups at baseline.

-   Post-hoc analysis across treatment groups at each timepoint identifies specific differences between treatments.

Interpret the Results

-   Main Fixed Effects:
    -   Treatment: Tests for differences in B/F ratio between treatment groups at baseline.

    -   Timepoint: Tests for changes in B/F ratio over time across all treatments.

    -   Treatment:Timepoint: Tests whether the effect of Timepoint on B/F ratio depends on the Treatment group.

-   Random Effects:
    -   (1|SubjectID): Accounts for within-subject variability.

```{r lmer-stuff}
adjust.method <- "FDR"
test <- "paired Wilcoxon"
formula_used <- "~ Intervention * time + (1 | subject)"

tse$bfratio <- log2(tse$bact_firm_ratio)
res <- run_lmer(tse, "bfratio")

```

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **Linear Mixed-Effects Model (LMM)**
-   Formula used: **`r formula_used`**

```{r table-fixed-fx}
#| label: tbl-table-of-interaction
sjPlot::tab_model(res)
```

```{r}
#| label: fig-figure-of-interaction
sjPlot::plot_model(res)
```


#### Post-Hoc Analysis

**Within-Group Comparisons**
```{r}
#| label: tbl-table-of-interaction-within
ref_grid <- emmeans(res, ~ Intervention * time)
within_group_comps <- contrast(ref_grid, method = "revpairwise", by = "Intervention")
kable(as.data.frame(within_group_comps), caption = "Estimates for Treatments")
# Add confidence intervals
within_group_comps_ci <- confint(within_group_comps)
```


```{r}
#| label: fig-figure-of-interaction-within
#| fig-width: 12
#| fig-height: 5
#| layout-ncol: 2
#| fig-cap: "Within-group changes over time from Baseline to Week 8."
#| fig-subcap: 
#|   - "Individual subject trajectories."
#|   - "Linear mixed model post-hoc estimates (Week 8 - Baseline)."

# A. Raw trajectory boxplot
plotBoxplot(tse, col.var = "bact_firm_ratio", x = "Intervention", fill.by = "time", pair.by = "subject", add.change = T, add.significance = T,
            point.shape = 19) +
  scale_color_manual(values = c("black", "black")) +
  theme_minimal(base_size = 15) +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12)
  )

# B. LMM estimates plot
within_group_df <- as.data.frame(within_group_comps_ci)
ggplot(within_group_df, aes(x = Intervention, y = estimate)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL), color = "#4A5568", size = 1.0, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Intervention", y = paste("Difference in B/F ratio (Week 8 - Baseline)"), 
       title = "LMM Within-Group Comparisons") +
  theme_minimal(base_size = 15) +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold")
  )
```

**Across-Group Comparisons**

> [!NOTE]
> **Interpretation Guide:** The following section shows examples of how to read the statistical tables based on baseline values. These are illustrative examples and may not reflect the exact updated statistics in your tables below.

* For Marginal Means (emmeans):
  * Negative emmean: A negative value indicates Firmicutes dominance (e.g. log2(B/F ratio) < 0).
  * Positive emmean: A positive value indicates Bacteroidetes dominance (e.g. log2(B/F ratio) > 0).

* For Contrasts:
  * Negative estimate: Indicates that the first group has a lower ratio than the second group.
  * Positive estimate: Indicates that the first group has a higher ratio than the second group.

```{r}
#| label: tbl-table-of-interaction-across-emmeans
posthoc <- emmeans(res, pairwise ~ Intervention | time)
# sjPlot::tab_model(as.data.frame(posthoc$contrasts)) 
data_emmean <- as.data.frame(posthoc$emmeans)
data_emmean <- data_emmean %>%
  # rename(emmeans = "coef") %>% 
  mutate(
    `2 ^ coef` = 2 ^ emmean, 
    lower.CL_actual = 2 ^ lower.CL,     
    upper.CL_actual = 2 ^ upper.CL
  )
some_order <- c("Intervention", "time", 
             "emmean", "2 ^ coef",
             "SE", "df", "lower.CL", "lower.CL_actual", 
             "upper.CL", "upper.CL_actual")
data_emmean <- data_emmean %>%
  select(all_of(some_order))  %>%
  rename(emmean = "coef")
kable(data_emmean, caption = "Estimated Marginal Means for Treatments")

```

```{r}
#| label: tbl-table-of-interaction-across-contrasts
kable(as.data.frame(posthoc$contrasts), caption = "Estimated contrasts for Treatments")
```


```{r}
#| label: fig-figure-of-interaction-across
#| fig-width: 12
#| fig-height: 3
#| fig.align: "center"
posthoc_ci <- confint(posthoc$contrasts)

# Convert to a data frame
posthoc_df <- as.data.frame(posthoc_ci)
#posthoc_df <- as.data.frame(posthoc$contrasts)
ggplot(posthoc_df, aes(x = contrast, y = estimate, color = time)) +
 geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    facet_wrap(~ time) +
  labs(x = "Comparison", y =  paste("Differences B/F"), 
       title = "Comparisons of B/F ratios") +
  theme_minimal()
```

## Data Summary

```{r data-summary}
tse
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::

