---
title: "Alpha Diversity"
params: 
  index: default
---

```{r variables}
# Alpha diversity
index <- params$index
adjust.method <- "FDR"
test <- "Wilcoxon"
```

```{r data}
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
```

## Group-wise comparisons

-   **Index:** `r params$index`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`




```{r pvalues-before}
# Calculate statistical results
results_before <- run_diversity_tests(tse, comparisons_before, variable, index, adjust.method)
results_before$p.adj <- p.adjust(results_before$p.value, method = "fdr")
# Format results for display
final_results_before <- results_before %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_before <- final_results_before %>%
  mutate(y.position = sapply(1:nrow(final_results_before), function(i) {
    vals1 <- colData(tse)[[index]][colData(tse)[[variable]] == group1[i]]
    vals2 <- colData(tse)[[index]][colData(tse)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_before)[colnames(final_results_before) == "logFC"] <- "log2FC"
```

```{r pvalues-after}
# Calculate statistical results
results_after <- run_diversity_tests(tse, comparisons_after, variable, index, adjust.method)
results_after$p.adj <- p.adjust(results_after$p.value, method = "fdr")
# Format results for display
final_results_after <- results_after %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_after <- final_results_after %>%
  mutate(y.position = sapply(1:nrow(final_results_after), function(i) {
    vals1 <- colData(tse)[[index]][colData(tse)[[variable]] == group1[i]]
    vals2 <- colData(tse)[[index]][colData(tse)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_after)[colnames(final_results_after) == "logFC"] <- "log2FC"
```

```{r pvalues-paired}
# Calculate statistical results
results_paired <- run_diversity_tests(tse, comparisons_paired, 
                                     variable, index, 
                                     adjust.method, paired = TRUE)
results_paired$p.adj <- p.adjust(results_paired$p.value, method = "fdr")
# Format results for display
final_results_paired <- results_paired %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 3)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_paired <- final_results_paired %>%
  mutate(y.position = sapply(1:nrow(final_results_paired), function(i) {
    vals1 <- colData(tse)[[index]][colData(tse)[[variable]] == group1[i]]
    vals2 <- colData(tse)[[index]][colData(tse)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
Intervention <- c("Control", "HWP", "HWP + LGG")
final_results_paired <- data.frame(Intervention = Intervention, final_results_paired)
final_results_paired$group1 <- NULL
final_results_paired$group2 <- NULL
colnames(final_results_paired)[colnames(final_results_paired) == "logFC"] <- "log2FC"
```


```{r visualize-plot}
#| label: fig-figure-of-findings-before-after
#| fig-width: 6
#| fig-height: 5
#| fig.align: "center"
plotBoxplot(tse, col.var = index, x = "time", 
            fill.by = "Intervention", add.significance = T,
            point.shape = 19) +
  scale_fill_manual(values = c("#F1F5F9", "#CBD5E1", "#64748B")) +
  scale_color_manual(values = c("black", "black", "black")) +
  theme_minimal(base_size = 15) +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12)
  )
```

## Table of the top findings

### Across-group

#### Before treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-before}
#| label: tbl-table-of-findings-before
final_results_before[] <- lapply(final_results_before, function(x) gsub(".before|.after", "", x))
final_results_before$group1 <- ifelse(final_results_before$group1  == "4", "Control", 
                    ifelse(final_results_before$group1  == "B", "HWP", "HWP + LGG"))
final_results_before$group2 <- ifelse(final_results_before$group2  == "4", "Control", 
                    ifelse(final_results_before$group2  == "B", "HWP", "HWP + LGG"))

  datatable(
      final_results_before %>% select(-y.position), 
      extensions = c("Buttons"),
      options = list(
        pageLength = 3,
        dom = 'Bfrtip',
        buttons = c("copy", "csv", "excel")
      ),
      caption = "Table of alpha diversity comparisons before Treatment",
      rownames = FALSE
  ) %>%
    formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
    formatSignif(columns = "p.adj", digits = 3)
```

#### After treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r table-of-findings-after}
#| label: tbl-table-of-findings-after
final_results_after[] <- lapply(final_results_after, function(x) gsub(".before|.after", "", x))
final_results_after$group1 <- ifelse(final_results_after$group1  == "4", "Control", 
                    ifelse(final_results_after$group1  == "B", "HWP", "HWP + LGG"))
final_results_after$group2 <- ifelse(final_results_after$group2  == "4", "Control", 
                    ifelse(final_results_after$group2  == "B", "HWP", "HWP + LGG"))
  datatable(
    final_results_after %>% select(-y.position), 
    extensions = c("Buttons"),
    options = list(
      pageLength = 3,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = "Table of alpha diversity comparisons after Treatment",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)

```



### Within-group- Paired treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **paired `r test`**

```{r table-of-findings-paired}
#| label: tbl-table-of-findings-paired

  datatable(
    final_results_paired %>% select(-y.position), 
    extensions = c("Buttons"),
    options = list(
      pageLength = 3,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = "Table of alpha diversity paired between timepoint(1 and 2)",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)
```

### Linear Mixed Model

Research Question 1: How does each treatment influence gut microbiota composition over time?

-   The interaction term (Treatment:Timepoint) tests whether the effect of time on alpha diversity depends on the treatment group.

-   Post-hoc analysis within each treatment group identifies specific changes in alpha diversity over time.

Research Question 2: Are there significant differences between treatments?

-   The main effect of Treatment tests for differences in alpha diversity between treatment groups at baseline.

-   Post-hoc analysis across treatment groups at each timepoint identifies specific differences between treatments.

Interpret the Results

-   Main Fixed Effects:
    -   Treatment: Tests for differences in alpha diversity between treatment groups at baseline.

    -   Timepoint: Tests for changes in alpha diversity over time across all treatments.

    -   Treatment:Timepoint: Tests whether the effect of Timepoint on alpha diversity depends on the Treatment group.

-   Random Effects:
    -   (1|SubjectID): Accounts for within-subject variability.

```{r lmer-stuff}

adjust.method <- "FDR"
test <- "paired Wilcoxon"
formula_used <- "~ Intervention * time + (1 | subject)"

if (index == "observed") {
  tse$observed_log2 <- log2(tse$observed)
  res <- run_lmer(tse, "observed_log2")
} else {
  res <- run_lmer(tse, index)
}

```

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **Linear Mixed-Effects Model (LMM)**
-   Formula used: **`r formula_used`**


```{r table-fixed-fx}
#| label: tbl-table-of-interaction
sjPlot::tab_model(res)
```


```{r figure-fixed-fx}
#| label: fig-figure-of-interaction
sjPlot::plot_model(res)
```
#### Post-Hoc Analysis

**Within-Group Comparisons**
```{r}
#| label: tbl-table-of-interaction-within
ref_grid <- emmeans(res, ~ Intervention * time)
within_group_comps <- contrast(ref_grid, method = "revpairwise", by = "Intervention")
kable(as.data.frame(within_group_comps), caption = "Estimates for Treatments")
# Add confidence intervals
within_group_comps_ci <- confint(within_group_comps)
```


```{r}
#| label: fig-figure-of-interaction-within
#| fig-width: 12
#| fig-height: 5
#| layout-ncol: 2
#| fig-cap: "Within-group changes over time from Baseline to Week 8."
#| fig-subcap: 
#|   - "Individual subject trajectories."
#|   - "Linear mixed model post-hoc estimates (Week 8 - Baseline)."

# A. Raw trajectory boxplot
plotBoxplot(tse, col.var = index, x = "Intervention", fill.by = "time", pair.by = "subject", add.change = T, add.significance = T,
            point.shape = 19) +
  scale_color_manual(values = c("black", "black")) +
  theme_minimal(base_size = 15) +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 12)
  )

# B. LMM estimates plot
within_group_df <- as.data.frame(within_group_comps_ci)
ggplot(within_group_df, aes(x = Intervention, y = estimate)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL), color = "#4A5568", size = 1.0, linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Intervention", y = paste("Difference in", index, "(Week 8 - Baseline)"), 
       title = "LMM Within-Group Comparisons") +
  theme_minimal(base_size = 15) +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold")
  )
```

**Across-Group Comparisons**

How to read the results;
N.B This interpretation is an example of how to read.

-   For Marginal Means (emmeans):
    -   Low diversity (e.g., Shannon = 2.5): "The mean Shannon diversity for Treatment 4 was 2.5 (95% CI: 2.3, 2.7), indicating lower microbial diversity compared to typical healthy gut microbiomes (Shannon ~3-5)."

    -   High diversity (e.g., Shannon = 4.2):"The mean Shannon diversity for Treatment B was 4.2 (95% CI: 4.0, 4.4), indicating a rich and balanced microbial community."

-   For Contrasts:
    -   Negative estimate (e.g., -0.8):"Treatment 4 had significantly lower Shannon diversity than Treatment B (estimate = -0.8, p < 0.05), suggesting reduced microbial richness."

    -   Positive estimate (e.g., 0.8):"Treatment B had significantly higher Shannon diversity than Treatment C (estimate = 0.8, p < 0.05), suggesting enhanced microbial balance."

```{r}
#| label: tbl-table-of-interaction-across-emmeans
posthoc <- emmeans(res, pairwise ~ Intervention | time)
# sjPlot::tab_model(as.data.frame(posthoc$contrasts)) 
data_emmean <- as.data.frame(posthoc$emmeans)
if (index == "observed") {
  data_emmean <- data_emmean %>%
    # rename(emmeans = "coef") %>% 
    mutate(
      `2 ^ coef` = 2 ^ emmean, 
      lower.CL_actual = 2 ^ lower.CL,     
      upper.CL_actual = 2 ^ upper.CL
    )
  some_order <- c("Intervention", "time", 
               "emmean", "2 ^ coef",
               "SE", "df", "lower.CL", "lower.CL_actual", 
               "upper.CL", "upper.CL_actual")
  data_emmean <- data_emmean %>%
    select(all_of(some_order))  %>%
    rename(emmean = "coef")
} else {
  some_order <- c("Intervention", "time", 
               "emmean", "SE", "df", "lower.CL",  
               "upper.CL")
  data_emmean <- data_emmean %>%
    select(all_of(some_order))  %>%
    rename(emmean = "coef")
}
kable(data_emmean,  caption = "Estimated Marginal Means for Treatments")
```


```{r}
#| label: tbl-table-of-interaction-across-contrasts
kable(as.data.frame(posthoc$contrasts), caption = "Estimated contrasts for Treatments")
```


```{r}
#| label: fig-figure-of-interaction-across
#| fig-width: 12
#| fig-height: 3
#| fig.align: "center"
posthoc_ci <- confint(posthoc$contrasts)

# Convert to a data frame
posthoc_df <- as.data.frame(posthoc_ci)
#posthoc_df <- as.data.frame(posthoc$contrasts)
ggplot(posthoc_df, aes(x = contrast, y = estimate, color = time)) +
 geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ time) +
  labs(x = "Comparison", y =  paste("Differences Alpha Diversity -", index), 
       title = "Comparisons of Alpha Diversity") +
  theme_minimal()
```


## Data Summary

```{r data-summary}
tse
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::
