---
title: "Analysis plan"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
plan <- load_plan()
```


## Model specification

```{r}
model_spec <- read_optional_table(c("config/model_specification.csv", "../config/model_specification.csv", "../../config/model_specification.csv"))
if (is.null(model_spec)) {
  cat("No model specification is connected yet.")
} else {
  model_spec
}
```

## Variables and covariates

```{r}
if (is.null(plan)) {
  cat("Variable roles will be populated from config/analysis_plan.R.")
} else {
  vars <- plan$variables
  data.frame(section = names(vars), value = vapply(vars, function(x) paste(unlist(x), collapse = "; "), character(1)))
}
```

## Multiple testing

State the FDR method, correction family, and threshold for interpretation. The correction family must match the generated model specification and result object.

## Sensitivity analyses

Define each sensitivity analysis as a deliberate change to the primary analysis, not as a replacement for the primary estimand.
