---
title: "Covariate diagnostics"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
plan <- load_plan()
```


## Covariate block

```{r}
if (is.null(plan)) {
  cat("Covariate definitions are pending until config/analysis_plan.R is connected.")
} else if (is.null(plan$variables$covariates)) {
  cat("No covariate block is defined in the config.")
} else {
  data.frame(block = names(plan$variables$covariates), variables = vapply(plan$variables$covariates, paste, collapse = "; ", FUN.VALUE = character(1)))
}
```

## Complete-case support

Report complete-case counts for each model family once the preparation and modeling scripts write their validation artifacts.

## Model-specification support

Document collinearity, sparse factor levels, balance, or exposure-covariate screening only when these diagnostics are part of the analysis plan or generated validation outputs.
