---
title: "Study overview"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
plan <- load_plan()
```


## Study identity

```{r}
if (is.null(plan)) {
  cat("No project config is connected yet. Create or link config/analysis_plan.R before finalizing this page.")
} else {
  data.frame(
    field = c("title", "key", "domain", "status"),
    value = c(plan$study$title, plan$study$key, plan$study$domain, plan$study$status)
  )
}
```

## Research question

State the primary scientific question from the draft analysis plan, including population, exposure or group, outcome or feature panel, timing, and intended estimand.

## Data sources

```{r}
if (is.null(plan)) {
  cat("Data paths will be listed after config/analysis_plan.R is connected.")
} else {
  data.frame(path_name = names(plan$paths), path = unlist(plan$paths, use.names = FALSE))
}
```

## Planned outputs

Link each planned output to the script or result object that creates it. Do not describe a result as complete until the corresponding validation artifact exists.
