---
title: "Bacteroidota/Baccilota ratio"
author:
  - name: Muluh
---


```{r data}
# Set Cairo to enable transparency in all devices
knitr::opts_chunk$set(dev = "png", 
                      dev.args = list(type = "cairo"))
# devtools::load_all("~/Rpackages/microbiome/miaverse/miaViz")
# Source fxns
source("../funct.R")
tse <- readRDS("../../data/tse.Rds")
```

```{r setup, include=FALSE}
# Define variables
tax.level     <- "Phylum"
variable      <- "diet"  
adjust.method <- "fdr"
test          <- "Wilcoxon"
index         <- "bact_firm_ratio"
```

### Group-wise comparisons

-   **Taxonomic level:** `r tax.level`
-   **Group variable:** `r variable`
-   **Multiple testing correction:** `r adjust.method`
-   **Statistical test** `r test`

In the report, a suffix of 1~ Baseline and 2~ Week 6

```{r bact-firm}
# Calculate the Bacteroidetes/Firmicutes ratio and add to colData
# using the chosen name
tse <- altExp(tse, "phylum")
tse$diet <- str_to_title(tse$diet)
tse$diet <- factor(tse$diet, levels=c("Oat", "Rice"))
ratio <- "bact_firm_ratio"
colData(tse)[[ratio]] <- assay(tse, "relabundance")["Bacteroidetes", ] /
                         assay(tse, "relabundance")["Firmicutes", ]
```

```{r run_diversity_tests}
# Calculate p-values
run_diversity_tests <- function(tse, comparisons, 
                                variable, ratio, 
                                adjust.method = "fdr", paired = FALSE) {
  # Get all values and groups
  values <- colData(tse)[[ratio]]
  groups <- colData(tse)[[variable]]
  
  # Create an empty results dataframe
  results <- data.frame(
    group1 = character(),
    group2 = character(),
    mean1 = numeric(),
    mean2 = numeric(),
    logFC = numeric(),
    p.value = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Loop through requested comparisons
  for (comp in comparisons) {
    group1 <- comp[1]
    group2 <- comp[2]
    
    # Subset data for just these two groups
    subset_indices <- which(groups %in% c(group1, group2))
    subset_values <- values[subset_indices]
    subset_groups <- groups[subset_indices]
    
    # Ensure paired data is properly aligned (if paired is TRUE)
    if (paired) {
      # Assuming 'subject' column exists for pairing
      subset_subjects <- tse$id[subset_indices]
      paired_data <- data.frame(
        subject = subset_subjects,
        group = subset_groups,
        value = subset_values
      ) %>%
        tidyr::pivot_wider(names_from = group, values_from = value) %>%
        drop_na()  # Drop pairs with missing data
      
      # Extract paired values
      paired_values <- paired_data %>% select(all_of(c(group1, group2)))
    }
    
    # Perform Wilcoxon test
    if (paired) {
      test_result <- wilcox.test(
        paired_values[[1]], paired_values[[2]],
        paired = TRUE, exact = FALSE
      )
    } else {
      test_result <- wilcox.test(
        subset_values ~ subset_groups,
        exact = FALSE
      )
    }
    
    # Calculate means
    mean1 <- mean(values[groups == group1])
    mean2 <- mean(values[groups == group2])

    if (paired) {
       # Note: when the values are paired, we can calculate FC separately
       # for every individual and summarize afterwards. 
       FC <- mean(values[groups == group2] / values[groups == group1])
    } else {
       FC <- mean2/mean1
    }

    # Add results
    results <- rbind(results, data.frame(
      group1 = group1,
      group2 = group2,
       mean1 = mean1,
       mean2 = mean2,
       logFC = log2(FC),
      p.value = test_result$p.value,
      stringsAsFactors = FALSE
    ))

  }
  
  # Add adjusted p-values using specified method
  # results$p.adj <- p.adjust(results$p.value, method = adjust.method)
  
  return(results)
}

```

```{r pvalues-before}
tse_before <- tse[, tse$visit == 1]
# Calculate statistical results
results_before <- run_diversity_tests(tse_before, comparisons_before, 
                                      variable, ratio = "bact_firm_ratio", 
                                      adjust.method)
results_before$p.adj <- p.adjust(results_before$p.value, method = "fdr")
# Format results for display
final_results_before <- results_before %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 2)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_before <- final_results_before %>%
  mutate(y.position = sapply(1:nrow(final_results_before), function(i) {
    vals1 <- colData(tse_before)[[ratio]][colData(tse_before)[[variable]] == group1[i]]
    vals2 <- colData(tse_before)[[ratio]][colData(tse_before)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_before)[colnames(final_results_before) == "logFC"] <- "log2FC"
```

```{r pvalues-after}
tse_after <- tse[, tse$visit == 2]
# Calculate statistical results
results_after <- run_diversity_tests(tse_after, comparisons_after, 
                                     variable, ratio = "bact_firm_ratio", 
                                     adjust.method)
results_after$p.adj <- p.adjust(results_after$p.value, method = "fdr")
# Format results for display
final_results_after <- results_after %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.adj = round(p.adj, 2)
  ) %>%
  select(group1, group2, mean1, mean2, logFC, p.adj)

# Calculate y-positions for p-value annotations
final_results_after <- final_results_after %>%
  mutate(y.position = sapply(1:nrow(final_results_after), function(i) {
    vals1 <- colData(tse_after)[[ratio]][colData(tse_after)[[variable]] == group1[i]]
    vals2 <- colData(tse_after)[[ratio]][colData(tse_after)[[variable]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
colnames(final_results_after)[colnames(final_results_after) == "logFC"] <- "log2FC"
```

```{r pvalues-paired}
diet_groups <- unique(tse$diet)
# Calculate statistical results
results_paired <- lapply(diet_groups, function(group) {
  # Subset tse for the current meal group
  tse_subset <- tse[, tse$diet == group]
  
  # Run the diversity tests
  res <- run_diversity_tests(
    tse = tse_subset, 
    comparisons = comparisons_paired, 
    variable = "duration", 
    ratio = "bact_firm_ratio", 
    adjust.method = adjust.method, 
    paired = TRUE
  )
  
  # Add the meal group as a new column
  res$diet <- group
  
  return(res)
}) %>% 
  bind_rows()

results_paired <- results_paired[, c("diet", setdiff(names(results_paired), "diet"))]

# results_paired$p.adj <- p.adjust(results_paired$p.value, method = "fdr")
# Format results for display
final_results_paired <- results_paired %>%
  mutate(
    mean1 = round(mean1, 2),
    mean2 = round(mean2, 2),
    logFC = round(logFC, 2),
    p.value = round(p.value, 2)
  ) %>%
  select(diet, group1, group2, mean1, mean2, logFC, p.value) %>% 
  arrange(diet)

# Calculate y-positions for p-value annotations
final_results_paired <- final_results_paired %>%
  mutate(y.position = sapply(1:nrow(final_results_paired), function(i) {
    vals1 <- colData(tse)[[ratio]][colData(tse)[["duration"]] == group1[i]]
    vals2 <- colData(tse)[[ratio]][colData(tse)[["duration"]] == group2[i]]
    max(c(vals1, vals2)) + i * 0.2
  }))
final_results_paired$group1 <- NULL
final_results_paired$group2 <- NULL
colnames(final_results_paired)[colnames(final_results_paired) == "logFC"] <- "log2FC"
```

```{r}
#| label: fig-figure-of-findings-before-after
#| fig-width: 5
#| fig-height: 4
#| fig.align: "center"

plotBoxplot(tse, col.var = "bact_firm_ratio", x = "duration", fill.by = "diet", add.significance = T) +
  scale_fill_viridis_d(option = "D")
```

#### Table of the top findings

#### Across-group

##### Before treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r}
#| label: tbl-table-of-findings-before

if (knitr::is_html_output()) {
  datatable(
      final_results_before %>% select(-y.position), 
      options = list(
        pageLength = 6,
        dom = 'Bfrtip'
      ),
      caption = "Table of  Bacteroidota/Baccilota ratio at the Baseline",
      rownames = FALSE
  ) %>%
    formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
    formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_before %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of Bacteroidota/Baccilota ratio at the Baseline",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

##### After treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**

```{r}
#| label: tbl-table-of-findings-after
if (knitr::is_html_output()) {
  datatable(
    final_results_after %>% select(-y.position), 
    options = list(
      pageLength = 6,
      dom = 'Bfrtip'
    ),
    caption = "Table of Bacteroidota/Baccilot at Week 6",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.adj", digits = 3)
} else {
  library(kableExtra)
  final_results_after %>% 
    select(-y.position) %>%
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of Bacteroidota/Baccilot at Week 6",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}

```

```{r}
#| label: fig-figure-of-findings-within
#| fig-width: 5
#| fig-height: 4
#| fig.align: "center"
#div_change
plotBoxplot(tse, col.var = "bact_firm_ratio", x = "diet", fill.by = "duration", pair.by = "id", add.change = T, add.significance = T)

```

#### Within-group

##### Paired treatment

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **paired `r test`**

```{r}
#| label: tbl-table-of-findings-paired
if (knitr::is_html_output()) {
  datatable(
    final_results_paired %>% select(-y.position),  
    options = list(
      pageLength = 4,
      dom = 'Bfrtip'
    ),
    caption = "Table of Bacteroidota/Baccilot paired between Baseline and Week 6",
    rownames = FALSE
) %>%
  formatRound(columns = c("mean1", "mean2", "log2FC"), digits = 2) %>%
  formatSignif(columns = "p.value", digits = 3)
} else {
  library(kableExtra)
  final_results_paired %>% 
    mutate(
      across(where(is.numeric), ~round(., 3))
      #p.adj = signif(p.adj, 3)
    ) %>%
    kbl(caption = "Table of Bacteroidota/Baccilot paired between Baseline and Week 6",
        booktabs = TRUE,
        longtable = TRUE,
        align = c('l', 'r', 'r', 'r', 'r')) %>%
    kable_styling(latex_options = "repeat_header") %>%
    row_spec(0, bold = TRUE)
}
```

### LMM

Research Question 1: Did the diet change the microbiota?

-   The interaction term (Treatment:Timepoint) tests whether the effect of time on B/F ratio depends on the diet group.

-   Post-hoc analysis within each diet group identifies specific changes in B/F ratio over time.

Research Question 2: Was the microbiota equal between the diet groups before and after diet?

-   The main effect of diet tests for differences in B/F ratio between diet groups at baseline.

-   Post-hoc analysis across diet groups at each timepoint identifies specific differences between diets

Interpret the Results

-   Main Effects:
    -   diet: Tests for differences in microbiota between Diet A and Diet B at Baseline.

    -   duration: Tests for changes in microbiota over time across both diets.

    -   diet:duration: Tests whether the effect of Timepoint depends on the diet group.

-   Random Effects:
    -   (1|id): Accounts for within-subject variability.

- **SE (Standard Error)**: Measures the uncertainty in the estimate. Smaller SE means more confidence in the effect.
- **DF (Degrees of Freedom)**: Used to assess statistical significance; depends on the number of observations and model complexity.
- **T00**: The variance of the random intercept (or slope) for a grouping factor (subject ID). Reflects variability between groups.
- **Nid**: The number of levels in the grouping factor (number of subjects).
- **ICC (Intra-Class Correlation)** or `iccand`: The proportion of total variance attributable to the grouping structure. Higher ICC means more similarity within groups.

```{r lmer-stuff}
adjust.method <- "FDR"
test <- "paired Wilcoxon"
formula_used <- "~ diet * timepoint + (1 | id)"
tse$diet <- factor(tse$diet, levels=c("Rice", "Oat"))

tse$bfratio <- log2(tse$bact_firm_ratio)
res <- run_lmer(tse, "bfratio")
```

-   Multiple testing correction: **`r adjust.method`**
-   Statistical test: **`r test`**
-   Formula used: **`r formula_used`**

```{r}
#| label: tbl-table-of-interaction
sjPlot::tab_model(res)
```

```{r}
#| label: fig-figure-of-interaction
sjPlot::plot_model(res)
```

#### Post-Hoc Analysis

*Within-Group Comparisons*
```{r}
#| label: tbl-table-of-interaction-within
ref_grid <- emmeans(res, ~ diet * duration)
within_group_comps <- contrast(ref_grid, method = "pairwise", by = "diet")
kable(as.data.frame(within_group_comps), caption = "Estimates for Diets")
# Add confidence intervals
within_group_comps_ci <- confint(within_group_comps)
```


```{r}
#| label: fig-figure-of-interaction-within
# Convert to a data frame
within_group_df <- as.data.frame(within_group_comps_ci)

ggplot(within_group_df, aes(x = diet, y = estimate, color = diet)) +
  geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Diet", y = paste("Differences B/F ratio (Baseline - Post-Intervention)"), 
       title = "Within-Group Comparisons of B/F ratio") +
  theme_minimal()
```

*Across-Group Comparisons*
How to read the results;

-   For Marginal Means (emmeans):
    -   Negative emmean:"The mean log2(B/F ratio) for Diet rice at baseline was `-2.238227` (95% CI: -2.607414		, -1.869040), indicating Firmicutes dominance (B/F ≈ 0.213)."

    -   Positive emmean:"if the mean log2(B/F ratio) for Diet oat was `0.4` (95% CI: 0.2, 0.6), then Bacteroidetes dominance (B/F ≈ 1.32)."

-   For Contrasts:
    -   Negative estimate:"Diet oat had a significantly lower log2(B/F ratio) than diet rice at timepoint 1 (estimate = `-1.0272585`, borderline p < 0.05), reflecting greater Firmicutes dominance."

    -   Positive estimate:"Diet oat had a  higher log2(B/F ratio) than diet rice at timepoint 2 (estimate = `0.5024736	`, p = 0.51), reflecting a non significant Bacteroidetes dominance."
```{r}
#| label: tbl-table-of-interaction-across-emmeans
posthoc <- emmeans(res, pairwise ~ diet | duration)
# sjPlot::tab_model(as.data.frame(posthoc$contrasts)) 
data_emmean <- as.data.frame(posthoc$emmeans)
data_emmean <- data_emmean %>%
  # rename(emmeans = "coef") %>% 
  mutate(
    `2 ^ coef` = 2 ^ emmean, 
    lower.CL_actual = 2 ^ lower.CL,     
    upper.CL_actual = 2 ^ upper.CL
  )
some_order <- c("diet", "duration", 
             "emmean", "2 ^ coef",
             "SE", "df", "lower.CL", "lower.CL_actual", 
             "upper.CL", "upper.CL_actual")
data_emmean <- data_emmean %>%
  select(all_of(some_order))  %>%
  rename(emmean = "coef", duration = "timepoint")
kable(data_emmean, caption = "Estimated Marginal Means for Diets")
```


```{r}
#| label: tbl-table-of-interaction-across-contrasts
kable(as.data.frame(posthoc$contrasts), caption = "Estimated contrasts for Diets")
```

```{r}
#| label: fig-figure-of-interaction-across
posthoc_ci <- confint(posthoc$contrasts)

# Convert to a data frame
posthoc_df <- as.data.frame(posthoc_ci)
#posthoc_df <- as.data.frame(posthoc$contrasts)
ggplot(posthoc_df, aes(x = contrast, y = estimate, color = duration)) +
 geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
  facet_wrap(~ duration) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Comparison", y =  paste("Differences B/F"), 
       title = "Comparisons of B/F ratios") +
  theme_minimal()
```

## Data Summary

```{r data-summary}
tse
```

```{r sessioninfo}
session_info <- capture.output(sessionInfo())
output_path <- "../../output/sesinfo.txt"
writeLines(session_info, output_path)
```
