---
title: "Differential abundance analysis"
author: ""
params: 
  tax.level: default
---

## Wilcoxon Tests

```{r setup}
# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

# Define variables and formulas
variable <- "group" 
prevalence.th <- 10/100
detection.th <- 0.1/100
tax.level <- params$tax.level
#tax.level <- "species_prevalent"

# Add CLR transformation
tse <- transformAssay(
  altExp(tse, tax.level), assay.type = "relabundance",
  method = "clr",
  pseudocount=TRUE
)
```

Analysis information:

-   Taxonomic level: `r tax.level`
-   assay: relabundance
-   transformation: clr

Taxa that are less prevalent than this (in relabundance) have been agglomerated:

-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%
-   Number of taxonomic groups: `r nrow(tse)`

Findings within BMI. (Paired Wilcoxon test, Bonferroni FDR adjustment)

```{r bmicheck}
# Check whether BMI is significantly different between conditions:
# Between conditions within time point 1
# Between conditions within time point 2
# Within conditions between time points (paired analysis)

pvs <- c()
for (tp in unique(tse$time)) {
  tse.sub <- tse[, tse$time==tp]
  pvs[[tp]] <- kruskal.test(tse.sub$BMI ~ tse.sub$Intervention)$p.value
}

# pick tse for time point 1 and 2 separately 
tse1 <- tse[, tse$timepoint=="before"]
tse2 <- tse[, tse$timepoint=="after"]
# Paired wilcoxon test for BMI within each treatment
for (g in unique(tse$Intervention)) {
  # subject order matches
  if (!all(tse1[, tse1$Intervention==g]$subject==tse2[, tse2$Intervention==g]$subject)) {stop("Subjects do not match")}

  pvs[[g]] <- wilcox.test(tse1[, tse1$Intervention==g]$BMI,
                          tse2[, tse2$Intervention==g]$BMI,
	                  paired=TRUE)$p.value
}
padj <- p.adjust(pvs, method="bonferroni")
print(padj)
```

```{r within_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (tp in unique(tse$timepoint)) {
  # Pick data for the given time point
  tse2 <- tse[, tse$timepoint==tp]
  pvs <- apply(assay(tse2, "clr"), 1, function (x) {kruskal.test(x ~ tse2$Treatment)$p.value})
  padj <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$time)
pvals.timepoints <- pvals
```

Number of significant findings within time points. (Kruskal test, Benjamini-Hochberg FDR adjustment, FDR<0.25)

```{r between_timepoint_analysis2}
print(colSums(pvals.timepoints<0.25))
```

```{r between_timepoint_analysis}
# Calculate adjusted p-values within each time point
# to test if any taxonomic group differs between treatments (Kruskal test)
pvals <- NULL
for (g in unique(tse$Intervention)) {
  # Pick data for the given Treatment for time points 1 and 2
  tse1  <- tse[, tse$Intervention==g & tse$timepoint=="before"]
  tse2  <- tse[, tse$Intervention==g & tse$timepoint=="after"]
  # Ensure that subject match (pairing ok)
  if (!all(tse1$subject==tse2$subject)) {stop("Subjects do not match!")}
  pvs   <- sapply(rownames(tse), function (tax) {
               wilcox.test(assay(tse1, "clr")[tax,], assay(tse2, "clr")[tax,], paired=TRUE)$p.value
	     }
	   )
  padj  <- p.adjust(pvs, "BH")
  pvals <- cbind(pvals, padj)
}
colnames(pvals) <- unique(tse$Intervention)
pvals.treatments <- pvals
```

Number of significant findings within treatments (Paired Wilcoxon test, Benjamini-Hochberg FDR adjustment, FDR<0.25)

```{r between_timepoint_analysis3}
print(colSums(pvals.treatments<0.25))
```

Let us look more closely to the significant hits. The logFC corresponds to logFC in CLR (tp2-tp1).

```{r tsesub}
file_path1 <- file.path("../../output/daa/daa_taxa/species/all_results.tsv")
file_path2 <- file.path("../../output/daa/daa_taxa/genus/all_results.tsv")

# Aggregate significant features across all intervention groups
signif.features <- rownames(pvals.treatments)[rowSums(pvals.treatments < 0.25, na.rm = TRUE) > 0]

taxa_of_interest <- c(signif.features)

daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% taxa_of_interest) %>%
      select(feature,metadata, value, qval_joint) 
daa_tab2 <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% taxa_of_interest) %>%
      select(feature,metadata, value, qval_joint)

daa_tab <- rbind(daa_tab1, daa_tab2) %>%
  mutate(qval_joint = signif(qval_joint, digits = 3)) 

```

```{r hits}
tabs <- NULL # Collect results in a table
for (g in unique(tse$Intervention)) {

  if (length(signif.features)>0) {
    tseg <- tse[, tse$Intervention==g] 
    dfg <- as.data.frame(colData(tseg)) 
    
    # for (tax in unique(rownames(tse))) {
    for (tax in signif.features) {
      dfg$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      dfg$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("subject", "timepoint", "signal")
      dfm <- melt(dfg[, vars])
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]
      
      dfg_before <- dfg[dfg$timepoint == "before", ]
      dfg_after  <- dfg[dfg$timepoint == "after", ]
      
      # Sort by subject to align vectors perfectly
      dfg_before <- dfg_before[order(dfg_before$subject), ]
      dfg_after  <- dfg_after[order(dfg_after$subject), ]

      tabs <-
        rbind(tabs,
        c(
          Intervention = g,
          Taxon = tax,
          mean1 = mean(dfg_before$signal, na.rm = TRUE),
          mean2 = mean(dfg_after$signal, na.rm = TRUE),
          logFC.clr = mean(dfg_after$clr - dfg_before$clr, na.rm = TRUE),
          FDR   = pv 
        ))
      
      
    }
  }
}

# Clean up the table
tabs <- as.data.frame(tabs)
tabs$mean.rel1 <- round(100 * as.numeric(tabs$mean1), 1); tabs$mean1 <- NULL
tabs$mean.rel2 <- round(100 * as.numeric(tabs$mean2), 1); tabs$mean2 <- NULL
tabs$FDR <- round(as.numeric(tabs$FDR), 3)
tabs$logFC.clr <- round(as.numeric(tabs$logFC.clr), 4)
tabs <- tabs %>%
  arrange(FDR, desc(abs(logFC.clr)))
```

For a complete table of results, see complete Treatment results
-   [Seafile](https://seafile.utu.fi/f/69be410144714a849d72/)
-   [Local](../../data/taxa_all_cleaned.csv)

```{r table_tm}
#| label: table_tm
if (knitr::is_html_output()) {
  datatable(tabs,
            extensions = c("Buttons"),
            options = list(
              pageLength = 10,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "Table of significant hits across treatments between timepoints")
} else {
  kable(tabs)
}
```

```{r}
plots_list <- list()
plots_list <- setNames(lapply(taxa_of_interest, function(taxon_name) {
    
    p1 <- plotBoxplot(tse, assay.type = "relabundance", x = "Intervention",  fill.by = "time", pair.by = "subject", features = c(taxon_name), facet.by = "rownames", add.change = T, add.significance = T, ncol = 1, label.size = 5) +
        theme_minimal(base_size = 15) +
         theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )

    p2 <- plotBoxplot(tse, assay.type = "relabundance", x = "time",  fill.by = "Intervention", features = c(taxon_name), facet.by = "rownames", ncol = 1, add.significance = T, label.size = 5) +
         scale_fill_viridis_d(option = "D") +
        theme_minimal(base_size = 15) +
         theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )
    
    p12 <- wrap_plots(p1, p2, axes = "collect") 
    
    return(p12)
    
}
    ),
 taxa_of_interest)


```

```{r}
#| label: fig-figure-of-highlights
#| layout-ncol: 1
#| fig-width: 15
#| fig-height: 5
ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}
```


```{r volcano-tab}
tabs_all <- NULL
for (g in unique(tse$Intervention)) {
  # P-values
  all_features <-  names(pvals.treatments[, g])

  if (length(all_features)>0) {
    tseg <- tse[, tse$Intervention==g] 
    df <- as.data.frame(colData(tseg)) 
    for (tax in all_features) {
      df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
      df$clr <- as.vector(assay(tseg, "clr")[tax,])      
      vars <- c("subject", "timepoint", "signal")
      dfm <- melt(df[, vars], verbose = FALSE)
      colnames(dfm) <- vars
      pv <- pvals.treatments[tax, g]

      df_before <- df[df$timepoint == "before", ]
      df_after  <- df[df$timepoint == "after", ]
      
      # Sort by subject to align vectors perfectly
      df_before <- df_before[order(df_before$subject), ]
      df_after  <- df_after[order(df_after$subject), ]

      tabs_all <-
        rbind(tabs_all,
        c(
          Intervention = g,
          Taxon = tax,
          mean1 = mean(df_before$signal, na.rm = TRUE),
          mean2 = mean(df_after$signal, na.rm = TRUE),
          logFC.clr = mean(df_after$clr - df_before$clr, na.rm = TRUE),
          FDR   = pv 
        ))
    }
  }
}
```

```{r volcano-tab-fix}
tabs_all <- as.data.frame(tabs_all)
tabs_all$mean.rel1 <- round(100 * as.numeric(tabs_all$mean1), 1); tabs_all$mean1 <- NULL
tabs_all$mean.rel2 <- round(100 * as.numeric(tabs_all$mean2), 1); tabs_all$mean2 <- NULL
tabs_all$FDR <- round(as.numeric(tabs_all$FDR), 3)
tabs_all$logFC.clr <- round(as.numeric(tabs_all$logFC.clr), 4)
```

```{r}
tabs_all_tp <- NULL
for ( bigG in unique(tse$Intervention)){
    tseG <- tse[, tse$Intervention==bigG] 
    for (g in unique(tseG$time)) {
      # P-values
      all_features <-  names(pvals.timepoints[, g])
    
      if (length(all_features)>0) {
        tseg <- tseG[, tseG$time==g] 
        df <- as.data.frame(colData(tseg)) 
        for (tax in all_features) {
          df$signal <- as.vector(assay(tseg, "relabundance")[tax,])
          df$clr <- as.vector(assay(tseg, "clr")[tax,])      
          vars <- c("subject", "time", "signal")
          dfm <- melt(df[, vars], verbose = FALSE)
          colnames(dfm) <- vars
          pv <- pvals.timepoints[tax, g]
    
          tabs_all_tp <-
            rbind(tabs_all_tp,
            c(
              Intervention = bigG,
              Timepoint = g,
              Taxon = tax,
              mean = mean(subset(df, time==g)$signal),
    
            sd = sd(subset(df, time == g)$signal),
            # Medians
            median = median(subset(df, time == g)$signal),
            # IQR (25th and 75th percentiles)
            iqr25 = quantile(subset(df, time == g)$signal, 0.25),
            iqr75 = quantile(subset(df, time == g)$signal, 0.75),
            mean_clr = mean(subset(df, time==g)$clr),
              FDR   = pv 
            ))
        }
      }
    }
}
```

```{r}
tabs_all_tp <- as.data.frame(tabs_all_tp)
tabs_all_tp$mean.rel <- round(100 * as.numeric(tabs_all_tp$mean), 1); tabs_all_tp$mean <- NULL

tabs_all_tp$FDR <- round(as.numeric(tabs_all_tp$FDR), 3)
tabs_all_tp$mean_clr <- round(as.numeric(tabs_all_tp$mean_clr), 4)
```

For a complete table of results, see complete timepoint results
-   [Seafile](https://seafile.utu.fi/smart-link/cde88ea4-b881-4a89-a6f9-2799ec69b723/)
-   [Local](../../data/tabs_all_tp_clean.csv)

```{r table_tp}
#| label: table
if (knitr::is_html_output()) {
  datatable(tabs_all_tp,
            extensions = c("Buttons"),
            options = list(
              pageLength = 10,
              dom = 'Bfrtip',
              buttons = c("copy", "csv", "excel")
            ),
            caption = "Table of significant hits within treatments between timepoints")
} else {
  kable(tabs_all_tp)
}
```


```{r volcano-plot}
#| layout-ncol: 1
#| fig.width: 12
#| fig.height: 10

volcano_plot <- ggplot(tabs_all, aes(x = logFC.clr, y = FDR, size = mean.rel2)) +
  geom_point(aes(color = FDR < 0.25), alpha = 0.7) +  
  labs(
    title = "Volcano Plot: Effect Size vs Significance",
    x = "Log Fold Change (CLR)",
    y = "FDR"
  ) +
  scale_color_manual(values = c("red", "blue")) +  
  theme_minimal(base_size = 18) +
  facet_wrap(~ Intervention) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
volcano_plot + 
  geom_text_repel(data = subset(tabs_all, FDR < 0.25),
                  aes(label = Taxon), size = 3, box.padding = 0.3)
```

```{r volcano-plot_tp}
#| layout-ncol: 1
#| fig.width: 12
#| fig.height: 10

volcano_plot <- ggplot(tabs_all_tp, aes(x = mean_clr, y = FDR, size = mean.rel)) +
  geom_point(aes(color = FDR < 0.5), alpha = 0.7) +  
  labs(
    title = "Abundance vs Significance Plot",
    x = "Mean Abundance (CLR)",
    y = "FDR(<0.5)"
  ) +
  scale_color_manual(values = c("red", "blue")) +  
  theme_minimal(base_size = 18) +
  facet_wrap(Timepoint ~ Intervention) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
volcano_plot + 
  geom_text_repel(data = subset(tabs_all_tp, FDR < 0.5),
                  aes(label = Taxon), size = 3, box.padding = 0.3)
```

```{r}
if (tax.level == "species_prevalent"){
    taxa_name <- "species"} else {taxa_name <- "genus"}
file_path <- file.path("../../output/daa/daa_taxa", taxa_name, "significant_results.tsv")
daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) 

  taxa_of_interest <- unique(daa_tab$feature)
```


```{r}
plots_list <- list()
plots_list <- setNames(lapply(taxa_of_interest, function(taxon_name) {
    
    p1 <- plotBoxplot(tse, assay.type = "relabundance", x = "Intervention",  fill.by = "time", pair.by = "subject", features = c(taxon_name), facet.by = "rownames", add.change = T, add.significance = T, ncol = 1, label.size = 5) +
      theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )

    p2 <- plotBoxplot(tse, assay.type = "relabundance", x = "time",  fill.by = "Intervention", features = c(taxon_name), facet.by = "rownames", add.significance = T, ncol = 1, label.size = 5) +
        scale_fill_viridis_d(option = "D") +
      theme_minimal(base_size = 15) +
  theme(text = element_text(family = "Roboto", size = 15), 
        axis.text = element_text(face = "bold"),
  )
    
    p12 <- wrap_plots(p1, p2, axes = "collect") 
    
    return(p12)
    
}
    ),
 taxa_of_interest)
```

## Visualizationof species with significant effects

The following box plots highlight significant species (qval_joint < 0.25) from the interaction model.

```{r plots_taxa_level}
#| label: fig-figure-of-highlights-mas
#| layout-ncol: 1
#| fig-width: 15
#| fig-height: 5

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}
```


## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::

