---
title: "DAA Metacyc"
---

# Interaction Model

```{r setup}
library(maaslin3)

source("../funct.R")
# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

metacyc_tse <- altExp(tse, "metacyc")

variable <- "Intervention" 
formula_used <- "~ Intervention * time + (1 | sbject)"
```

-   Group variable: **`r variable`**
-   Formula used: **`r formula_used`**

```{r data_prep}
# Create output directory for results
output_base_dir <- "../../output/daa/daa_mtc"
dir.create(output_base_dir, showWarnings = FALSE)
# Keep paired cases that have both time points
paired.cases <- names(which(table(colData(metacyc_tse)$subject) == 2))
metacyc_tse <- metacyc_tse[, metacyc_tse$subject %in% paired.cases]
```

```{r maaslin all}
# Run Maaslin3
run_maaslin3 <- function(taxa_tse, formula_used, dir) {
  # Keep only paired cases (individuals with both timepoint points)
  # paired.cases <- names(which(table(taxa_tse$subject) == 2))
  # taxa_tse <- taxa_tse[, taxa_tse$subject %in% paired.cases]
  # taxa_tse$treat <- taxa_tse$group

  # Run MaAsLin3
  fit <- maaslin3(
    taxa_tse, 
    output = dir,
    formula = formula_used,
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,
    standardize = TRUE,
    max_significance = 0.25,
    plot_associations = FALSE,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "WARN"
  )
  
  return(fit)
}
```

```{r}
file_path <- file.path("../../output/daa/daa_mtc/interaction/significant_results.tsv")
 if (!file.exists(file_path)) {
    formula_used_int <- '~ Intervention * time + (1 | subject)'
    # Create output directory for this fxnomic level
    output_dir <- file.path(output_base_dir, "interaction")
    dir.create(output_dir, showWarnings = FALSE)
    fit <- run_maaslin3(metacyc_tse, formula_used_int, output_dir)
}
```

```{r print_table}
#| label: table-of-significant-results
file_path <- file.path("../../output/daa/daa_mtc/interaction/all_results.tsv2")
total_features <- nrow(metacyc_tse)

daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
    filter(!is.na(feature)) %>%
    select(EnzymeName, metadata, value, coef, qval_individual, qval_joint, model)

table_caption <- paste("Table of features for metacyc- Total features analyzed:", total_features)
  
# Split the results by model
daa_tab_split <- split(daa_tab, daa_tab$model)
# Filter for qval_joint between 0 and 0.25
daa_tab_split <- lapply(daa_tab_split, function(df) {
  df <- df[df$qval_joint >= 0 & df$qval_joint <= 0.25, ]
  df
})
daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
    rename(metadata = "variable", value = "term") %>%
      select(-model)

# Format output for HTML or LaTeX
datatable(
  daa_tab_split$abundance,
  options = list(pageLength = 20, dom = 'Bfrtip'),
  caption = table_caption,
  rownames = FALSE
) %>%
 formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)

```

# Post-Hoc Analysis

## Within-Group Comparisons
```{r run_maaslin_o}
tse_4 <- metacyc_tse[, metacyc_tse$Treatment == "4"]
tse_B <- metacyc_tse[, metacyc_tse$Treatment == "B"]
tse_C <- metacyc_tse[, metacyc_tse$Treatment == "C"]

formula_used_int <- '~ time + (1 | subject)'
file_path <- file.path("../../output/daa/daa_mtc/4/significant_results.tsv")
tse_list = list(tse_4, tse_B, tse_C)
if (!file.exists(file_path)) {
  lapply(tse_list, function(tse){
    diet_dir <- file.path(output_base_dir, unique(tse$Treatment))
    dir.create(diet_dir, showWarnings = FALSE, recursive = TRUE)
    run_maaslin3(tse, formula_used_int, diet_dir)
  }
    )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_mtc", unique(tse$Treatment), "all_results.tsv2")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(EnzymeName, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of features for path coverage within group", unique(tse$Intervention),
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    # Filter for qval_joint between 0 and 0.25
    daa_tab_split <- lapply(daa_tab_split, function(df) {
      df <- df[df$qval_joint >= 0 & df$qval_joint <= 0.25, ]
      df
    })
    daa_tab_split$abundance <-  daa_tab_split$abundance %>%
      select(c(-model, -metadata, -value))
    
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 20, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

### Intervention Control
```{r}
tabs[[1]]
```

### Intervention HWP
```{r}
tabs[[2]]
```

### Intervention HWP + LGG
```{r}
tabs[[3]]
```

## Across-Group Comparisons
```{r}
tse_after <- metacyc_tse[, metacyc_tse$timepoint == "after"]
tse_before <- metacyc_tse[, metacyc_tse$timepoint == "before"]

formula_used_int <- '~ Intervention'
file_path <- file.path("../../output/daa/daa_mtc", unique(tse_before$timepoint), "all_results.tsv")
tse_list = list(tse_after, tse_before)
if (!file.exists(file_path)) {
lapply(tse_list, function(tse){
  timepoint_dir <- file.path(output_base_dir, unique(tse$timepoint))
  dir.create(timepoint_dir, showWarnings = FALSE, recursive = TRUE)
  
  run_maaslin3(tse, formula_used_int, timepoint_dir)
}
  )
}
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_mtc", unique(tse$timepoint), "all_results.tsv2")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(EnzymeName, metadata, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Table of features for metacyc at", unique(tse$duration),
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    # Filter for qval_joint between 0 and 0.25
    daa_tab_split <- lapply(daa_tab_split, function(df) {
      df <- df[df$qval_joint >= 0 & df$qval_joint <= 0.25, ]
      df
    })
    daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
      rename(value = "term") %>%
      select(c(-model, -metadata))
  
      datatable(
        daa_tab_split$abundance,
        options = list(pageLength = 20, dom = 'Bfrtip'),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )
```

### Before Intervention - Week 0
```{r}
tabs[[1]]
```

### Post Intervention - Week 8
```{r}
tabs[[2]]
```

```{r setup_m}

prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Intervention" 
formula_used <- "~ Intervention * time + (1 | id)"
correction <- "Benjamini-Hochberg FDR adjustment"

```

## Maaslin3 significant hits

This section highlights significant taxa from maaslin3.

Analysis information:

-   Method: maaslin3
-   Group variable: `r variable`
-   Formula used: **`r formula_used`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

```{r tsesub_m}
file_path1 <- file.path("../../output/daa/daa_mtc/interaction/significant_results.tsv")
file_path2 <- file.path("../../output/daa/daa_mtc/interaction/all_results.tsv")
daa_tab1 <- read_tsv(file_path1, col_names = TRUE, show_col_types = FALSE)
daa_tab1 <- na.omit(daa_tab1)
significant_fxns1 <- unique(daa_tab1$feature)
significant_fxns <- sub("^X", "", significant_fxns1)
enzymenames <- read_tsv("../../output/daa/daa_mtc/enzymenames.tsv")

daa_tab <- read_tsv(file_path2, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      filter(model == "abundance") %>%
      filter(feature %in% significant_fxns1) %>%
      select(feature,metadata, value, qval_joint) %>%
      mutate(qval_joint = signif(qval_joint, digits = 3)) 
```

```{r before-after_m}
df <- as.data.frame(colData(metacyc_tse))

plot_fxn <- function(fxn) {
  rownames(metacyc_tse) <- rowData(metacyc_tse)$Feature
  df$signal <- as.vector(assay(metacyc_tse, "relabundance")[fxn,])
  df$signal <- unlist(df$signal)
  
  enzyme_name <- enzymenames$EnzymeName[enzymenames$feature == fxn]
    
    # daa_tab_tax <- daa_tab %>%
    #   filter(feature == fxn)
    # 
    # anot_tax <- daa_tab_tax %>% 
    #     filter(metadata == "intervention")
    # 
    # anot_tax$pval_color <- ifelse(anot_tax$qval_joint <= 0.25, "red", "blue")
    # Subset for before and after
    df_before <- df %>% filter(timepoint == "before")
    df_after <- df %>% filter(timepoint == "after")
    
    # Prepare data for plotting
        vars <- c("subject", "timepoint", "signal", "time")
        dfm <- melt(df[, vars])
        colnames(dfm) <- vars
        
        # Filter and mutate for paired design
        df <- df %>%
          filter(paired == "yes") %>%
          group_by(subject) %>%
          mutate(
            change = if_else(timepoint == "after", signal - signal[timepoint == "before"], 0),
            change = if_else(timepoint == "before", change[timepoint == "after"], change)
          ) %>%
          ungroup() %>%
          as.data.frame(check.names = FALSE)
        
          min_change <- min(df$change, na.rm = TRUE)
          max_change <- max(df$change, na.rm = TRUE)
          median_change <- median(df$change, na.rm = TRUE)
    
    p12 <- ggplot(df, aes(x = Intervention, y = signal)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point() + 
      scale_y_log10(labels = scales::percent) + 
      theme(text = element_text(size = 15), legend.position = "none") +
      labs(title = enzyme_name, y = "Abundance (%)", x = "Intervention") +
      facet_wrap(~ time, nrow = 1)
    
    p3 <- ggplot(df, aes(x = time, y = signal)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point() + 
      geom_line(aes(group = subject, color = change)) +
      scale_color_gradient2(midpoint = median_change, low = "blue", mid = "white", high = "red", space = "Lab", limits = c(min_change, max_change)) +
      facet_wrap(~ Intervention) +
      scale_y_log10(labels = scales::percent) + 
      theme(text = element_text(size = 15), legend.position = "right") +
      labs(x = "Time", y = "Abundance (%)")
    
    # Combine the before and after plots
    return(wrap_plots(p12, p3, axes = "collect")
  )
  }

```

```{r}
plots_list <- list()
name_vec <- enzymenames$EnzymeName[match(significant_fxns, enzymenames$feature)]
plots_list <- setNames(lapply(significant_fxns, plot_fxn),
                      name_vec)
```

```{r hits_m}
#| layout-ncol: 1
#| fig-width: 18 
#| fig-height: 5 

ordered_plots <- plots_list[order(names(plots_list))]

for (plot_name in names(ordered_plots)) {
  print(ordered_plots[[plot_name]])
}

```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::
