---
title: "DAA"
author: ""
---

# Differential abundance analysis

```{r setup}
library(maaslin3)

# Source group-assignment
source("../funct.R")

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")
taxa_name <- "species"

prevalence.th <- 10/100
detection.th <- 0.1/100

variable <- "Intervention" 
formula_used_accross_gp <- '~ Intervention'
formula_used_within_gp <- '~ time + (1 | subject)'
formula_used_int <- '~ Intervention * time + (1 | subject)'
correction <- "Benjamini-Hochberg FDR adjustment"
```

Analysis information:

-   Method: maaslin3
-   Taxonomic level: **`r taxa_name`**
-   Group variable: `r variable`
-   Formula used for interaction: **`r formula_used_int`**
-   Formula used across groups: **`r formula_used_accross_gp`**
-   Formula used within groups: **`r formula_used_within_gp`**
-   Correction: **`r correction`**
-   Detection threshold: `r 100 * detection.th`%
-   Prevalence threshold: `r 100 * prevalence.th`%

Research Questions:

1.  How does each treatment influence gut microbiota composition over time?

2.  Are there significant differences between treatments?

We’ll use a mixed-effects model with:

-   Fixed Effects:

    -   Treatment: To test differences between treatments.

    -   Timepoint: Tests for changes over time across all treatments.

    -   Treatment:Timepoint: To test whether the effect of time depends on the treatment.

-   Random Effects:

    -   (1\|subject): To account for repeated measures within subjects.

Interpret the Results

-   Treatment:Timepoint Interaction:

A significant interaction indicates that the effect of time depends on the treatment group.

This answers Research Question 1 (how each treatment influences microbiota over time).

-   Treatment Effects:

The coefficients for Treatment indicate differences between treatment groups at baseline.

This answers Research Question 2 (differences between treatments).

```{r data_prep}
# Create output directory for results
output_base_dir <- "../../output/daa/daa_taxa"
dir.create(output_base_dir, showWarnings = FALSE)
```

```{r maaslin all}
# Run Maaslin3
run_maaslin3 <- function(taxa_tse, taxa_name, formula_used, dir) {
  # Keep only paired cases (individuals with both timepoint points)
  # paired.cases <- names(which(table(taxa_tse$subject) == 2))
  # taxa_tse <- taxa_tse[, taxa_tse$subject %in% paired.cases]
  # taxa_tse$treat <- taxa_tse$group

  # Run MaAsLin3
  fit <- maaslin3(
    taxa_tse, 
    output = dir,
    formula = formula_used,
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,
    standardize = TRUE,
    max_significance = 0.25,
    plot_associations = FALSE,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "WARN"
  )
  
  return(fit)
}
```

```{r run_maaslin}
taxa_tse <- altExp(tse, paste0(taxa_name, "_prevalent"))
formula_used_int <- '~ Intervention * time + (1 | subject)'
# Create output directory for this taxonomic level
output_dir <- file.path(output_base_dir, taxa_name)
dir.create(output_dir, showWarnings = FALSE)
file_path <- file.path("../../output/daa/daa_taxa", taxa_name, "significant_results.tsv")
if (!file.exists(file_path)) {
fit <- run_maaslin3(taxa_tse, taxa_name, formula_used_int, output_dir)
}
```

## Interaction Model

```{r run_maaslin_m}
#| label: tbl-table-interaction-species
  total_features <- nrow(taxa_tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, metadata, value, coef, qval_individual, qval_joint, model)
  taxa_of_interest <- unique(daa_tab$feature)
  
  table_caption <- paste("Table of significant features for", taxa_name, 
                           "- Total features analyzed:", total_features)
    
  # Split the results by model
  daa_tab_split <- split(daa_tab, daa_tab$model)
  daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
    rename(metadata = "variable", value = "term") %>%
      select(-model)
  
  # Format output for HTML or LaTeX
  datatable(
    daa_tab_split$abundance,
    extensions = c("Buttons"),
    options = list(
      pageLength = 6,
      dom = 'Bfrtip',
      buttons = c("copy", "csv", "excel")
    ),
    caption = table_caption,
    rownames = FALSE
  ) %>%
  formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)

```

## Post-Hoc Analysis

### Within-Group Comparisons
```{r run_maaslin_o}

tse_4 <- taxa_tse[, taxa_tse$Treatment == "4"]
tse_B <- taxa_tse[, taxa_tse$Treatment == "B"]
tse_C <- taxa_tse[, taxa_tse$Treatment == "C"]

formula_used_int <- '~ time + (1 | subject)'

tse_list = list(tse_4, tse_B, tse_C)
lapply(tse_list, function(tse){
  treatment_dir <- file.path(output_base_dir, unique(tse$Treatment), taxa_name)
  target_file <- file.path(treatment_dir, "all_results.tsv")
  if (!file.exists(target_file)) {
    dir.create(treatment_dir, showWarnings = FALSE, recursive = TRUE)
    run_maaslin3(tse, taxa_name, formula_used_int, treatment_dir)
  }
})
```

```{r}
file_path_w <- file.path("../../data/wilcoxon_tables/wilcox_species.csv")
daa_tab_w <- read_csv(file_path_w, col_names = TRUE, show_col_types = FALSE)
daa_tab_w <- daa_tab_w %>% rename(Taxon = "feature")
```


```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_taxa", unique(tse$Treatment), taxa_name, "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Comparison between timepoints within group", unique(tse$Treatment),"." , taxa_name, "level, relative abundance."," Method: Maaslin3, FDR: Wilcoxon", "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    daa_tab_w <- daa_tab_w[daa_tab_w$Treatment == unique(tse$Treatment), ]  %>%
      select(-Treatment)
    daa_tab_split$abundance <- merge(daa_tab_split$abundance, daa_tab_w, by = "feature") %>%
      select(-model)
    some_order <- c("feature", "mean.rel1",	"mean.rel2", "logFC.clr",	"coef",	"qval_individual",	"qval_joint",	"FDR")
      daa_tab_split$abundance <-  daa_tab_split$abundance  %>% 
        select(all_of(some_order))
      
      plot1 <- ggplot(daa_tab_split$abundance, aes(x = coef, y = logFC.clr)) +
        geom_point() +
        labs(title = "a) Internal validation") +
        geom_smooth(method = "lm", se = FALSE, color = "red")
      
      volcano_plot <- ggplot(daa_tab_split$abundance, aes(x = coef, y = qval_joint, size = mean.rel2)) +
        geom_point(aes(color = qval_joint < 0.25), alpha = 0.7) +  
        labs(
          title = "b) Volcano Plot: significance vs effect size",
          x = "Estimates (coef)",
          y = "qval_joint(<0.25)"
        ) +
        scale_color_manual(values = c("red", "blue")) +  
        theme_minimal() +
        theme(
          legend.position = "none",
          plot.title = element_text(hjust = 0.5)
        )
      volcano_plot <- volcano_plot + 
        geom_text_repel(data = subset(daa_tab_split$abundance, qval_joint < 0.25),
                        aes(label = feature), size = 3, box.padding = 0.3)
      
      plot <- wrap_plots(plot1, volcano_plot)
    
      table <- datatable(
        daa_tab_split$abundance,
        extensions = c("Buttons"),
        options = list(
          pageLength = 6,
          dom = 'Bfrtip',
          buttons = c("copy", "csv", "excel")
        ),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
      
       list(plot = plot, table = table)
}
  )
```

```{r}
#| label: fig-figure-within-genus_all
#| fig-width: 8
#| fig-height: 10

tab_heat <- lapply(tse_list, function(tse){
    file_path <- file.path("../../output/daa/daa_taxa", unique(tse$Treatment), taxa_name, "all_results.tsv")
    df <- read_tsv(file_path, show_col_types = FALSE)
    df$Intervention <- unique(tse$Intervention) 
    df$Treatment <- unique(tse$Treatment) 
    df_split <- split(df, df$model)
    return(df_split$abundance)
}
  )
tab_heat <- bind_rows(tab_heat)

tab_heat <- merge(tab_heat, daa_tab_w, by = c("feature", "Treatment"))

tab_heat <- tab_heat %>%
  mutate(
    signif_label = case_when(
      qval_joint < 0.05 ~ "*",
      qval_joint < 0.10 ~ "†",
      TRUE ~ ""
    ),
    label = ifelse(!is.na(logFC.clr), paste0(round(logFC.clr, 4), signif_label), "")
  )

tab_heat <- tab_heat[tab_heat$qval_joint < 0.25, ]

# Plot
ggplot(tab_heat, aes(x = Intervention, y = feature, fill = logFC.clr)) +
  geom_tile(color = "white", size = 0.3) +
  geom_text(aes(label = label), size = 5) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, na.value = "grey50") +
  theme_minimal(base_size = 14) +
  theme(text = element_text(family = "Roboto"), 
        axis.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()
  ) +
  labs(
    x = NULL, y = NULL, fill = "Effect Size"
  ) + labs(caption = "p < 0.05 = *   p <0.10 = †")
```

#### Group 4 - Estimates the change in relative abundance between timepoint 1 and timepoint 2
```{r}
#| label: tbl-table-within-species4
tabs[[1]]$table
```

```{r}
#| label: fig-figure-within-species4
#| fig-cap: "Group 4: (a)  Internal validation: checking that alternative tests give correlated results, (b) volcano plot showing the relation between effect size and significance"
#| fig-width: 10
#| fig-height: 5
tabs[[1]]$plot
```

#### Group B - Estimates the change in relative abundance between timepoint 1 and timepoint 2
```{r}
#| label: tbl-table-within-speciesB
tabs[[2]]$table
```

```{r}
#| label: fig-figure-within-speciesB
#| fig-cap: "Group B: (a)  Internal validation: checking that alternative tests give correlated results, (b) volcano plot showing the relation between effect size and significance"
#| fig-width: 10
#| fig-height: 5
tabs[[2]]$plot
```


#### Group C - Estimates the change in relative abundance between timepoint 1 and timepoint 2
```{r}
#| label: tbl-table-within-speciesC
tabs[[3]]$table
```

```{r}
#| label: fig-figure-within-speciesC
#| fig-cap: "Group C: (a)  Internal validation: checking that alternative tests give correlated results, (b) volcano plot showing the relation between effect size and significance"
#| fig-width: 10
#| fig-height: 5
tabs[[3]]$plot
```

### Across-Group Comparisons
```{r}
tse_after <- taxa_tse[, taxa_tse$timepoint == "after"]
tse_before <- taxa_tse[, taxa_tse$timepoint == "before"]

formula_used_int <- '~ Intervention'

tse_list = list(tse_before, tse_after)
lapply(tse_list, function(tse){
  timepoint_dir <- file.path(output_base_dir, unique(tse$timepoint), taxa_name)
  target_file <- file.path(timepoint_dir, "all_results.tsv")
  if (!file.exists(target_file)) {
    dir.create(timepoint_dir, showWarnings = FALSE, recursive = TRUE)
    run_maaslin3(tse, taxa_name, formula_used_int, timepoint_dir)
  }
})
```

```{r}
tabs <- lapply(tse_list, function(tse){
file_path <- file.path("../../output/daa/daa_taxa", unique(tse$timepoint), taxa_name, "all_results.tsv")
  total_features <- nrow(tse)
  daa_tab <- read_tsv(file_path, col_names = TRUE, show_col_types = FALSE) %>%
      filter(!is.na(feature)) %>%
      select(feature, value, coef, qval_individual, qval_joint, model)
  
  table_caption <- paste("Comparison between between treatments(reference = 4) at timepoint", unique(tse$timepoint), "." , taxa_name, "level, relative abundance."," Method: Maaslin3",
                           "- Total features analyzed:", total_features)
    
    # Split the results by model
    daa_tab_split <- split(daa_tab, daa_tab$model)
    daa_tab_split$abundance <-  daa_tab_split$abundance %>% 
    rename( value = "term") %>%
      select(-model)
  
      datatable(
        daa_tab_split$abundance,
        extensions = c("Buttons"),
        options = list(
          pageLength = 6,
          dom = 'Bfrtip',
          buttons = c("copy", "csv", "excel")
        ),
        caption = table_caption,
        rownames = FALSE
      ) %>%
      formatSignif(columns = c("qval_joint", "qval_individual", "coef"), digits = 3)
}
  )

```

#### Before Intervention - Estimates the difference in relative abundance between treatments at timepoint 1
```{r}
#| label: tbl-table-before-species
tabs[[1]]
```

#### Post Intervention - Estimates the difference in relative abundance between treatments at timepoint 2
```{r}
#| label: tbl-table-after-species
tabs[[2]]
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::

