---
title: "DAA Methods Comparison"
author: ""
---

In this report, we compare the findings of our three DAA methods (**Wilcoxon Paired**, **MaAsLin3**, and **LimROTS**) for longitudinal within-group comparisons (Week 8 vs. Baseline) within each of the three intervention arms (**Control**, **HWP**, and **HWP + LGG**).

Comparing multiple DAA methods helps validate findings: taxa identified by multiple methods represent highly robust biological signals.

```{r setup, message=FALSE, warning=FALSE}
# Source helper functions and load packages
source("../funct.R")
library(mia)
library(dplyr)
library(readr)
library(ggplot2)
library(ggVennDiagram)
library(DT)
library(knitr)
library(patchwork)
library(tidyr)

# Load the TreeSummarizedExperiment object
tse <- readRDS("../../data/tse.Rds")

# Mapping of Intervention groups to Treatment codes in MaAsLin3 output folder names
treatment_map <- c("Control" = "4", "HWP" = "B", "HWP + LGG" = "C")

# Helper to run Wilcoxon paired analysis on CLR abundances
run_wilcoxon_paired <- function(tse_obj, assay_name, group_name) {
  # Subset to group
  tse_g <- tse_obj[, tse_obj$Intervention == group_name]
  tse1 <- tse_g[, tse_g$timepoint == "before"]
  tse2 <- tse_g[, tse_g$timepoint == "after"]
  
  # Ensure pairing is aligned
  tse1 <- tse1[, order(tse1$subject)]
  tse2 <- tse2[, order(tse2$subject)]
  
  pvs <- sapply(rownames(tse_obj), function(tax) {
    wilcox.test(assay(tse1, "clr")[tax, ], assay(tse2, "clr")[tax, ], paired = TRUE)$p.value
  })
  
  log2fc <- sapply(rownames(tse_obj), function(tax) {
    mean(assay(tse2, "clr")[tax, ] - assay(tse1, "clr")[tax, ], na.rm = TRUE)
  })
  
  padj <- p.adjust(pvs, "BH")
  
  data.frame(
    feature = rownames(tse_obj),
    logFC = log2fc,
    p.value = pvs,
    FDR = padj
  )
}

# Helper to load MaAsLin3 longitudinal results
load_maaslin3_results <- function(group_code, level) {
  file_path <- file.path("../../output/daa/daa_taxa", group_code, level, "all_results.tsv")
  if (!file.exists(file_path)) {
    return(data.frame(feature = character(), coef = numeric(), FDR = numeric()))
  }
  read_tsv(file_path, show_col_types = FALSE) %>%
    filter(!is.na(feature), model == "abundance") %>%
    select(feature, coef, FDR = qval_joint)
}

# Helper to run LimROTS longitudinal analysis
run_limrots_long <- function(tse_obj, assay_name, group_name) {
  temp_tse <- tse_obj[, tse_obj$Intervention == group_name]
  res <- run_limrots_analysis(
    tse = temp_tse, 
    assay_name = assay_name, 
    group_col = "time", 
    formula_str = "~ 0 + time",
    niter = 200 # For fast rendering while preserving statistical validity
  )
  res %>% select(feature, logFC, p.value, FDR)
}

# Helper to plot Venn Diagram of overlaps
plot_venn_overlap <- function(sig_w, sig_m, sig_l, group_name) {
  all_sig <- unique(c(sig_w, sig_m, sig_l))
  if (length(all_sig) > 0) {
    x <- list(Wilcoxon = sig_w, MaAsLin3 = sig_m, LimROTS = sig_l)
    ggVennDiagram(x, label_alpha = 0.5) +
      scale_fill_distiller(palette = "Blues", direction = 1) +
      labs(title = paste("Significant Taxa Overlap (FDR < 0.25) -", group_name)) +
      theme(plot.title = element_text(size = 14, face = "bold"))
  } else {
    ggplot() + 
      annotate("text", x = 0.5, y = 0.5, label = "No significant features (FDR < 0.25) found by any method.") +
      theme_void() +
      labs(title = paste("Significant Taxa Overlap (FDR < 0.25) -", group_name))
  }
}

# Helper to plot effect size correlation scatter plots
plot_correlations <- function(df, group_name) {
  if (nrow(df) == 0) return(ggplot() + theme_void())
  
  # Correlation helper
  make_scatter <- function(df_sub, x_col, y_col, x_label, y_label) {
    corr_val <- cor(df_sub[[x_col]], df_sub[[y_col]], method = "spearman", use = "complete.obs")
    ggplot(df_sub, aes_string(x = x_col, y = y_col)) +
      geom_point(alpha = 0.7, color = "#4A5568", size = 2) +
      geom_smooth(method = "lm", color = "red", linetype = "dashed", se = FALSE) +
      annotate("text", x = min(df_sub[[x_col]], na.rm = TRUE), y = max(df_sub[[y_col]], na.rm = TRUE),
               label = paste("Spearman r =", round(corr_val, 3)), hjust = 0, vjust = 1, size = 4, fontface = "bold") +
      labs(x = x_label, y = y_label) +
      theme_minimal(base_size = 12)
  }
  
  p1 <- make_scatter(df, "logFC_wilcox", "logFC_limrots", "Wilcoxon log2FC", "LimROTS logFC")
  p2 <- make_scatter(df, "logFC_wilcox", "coef_maaslin", "Wilcoxon log2FC", "MaAsLin3 Coefficient")
  p3 <- make_scatter(df, "logFC_limrots", "coef_maaslin", "LimROTS logFC", "MaAsLin3 Coefficient")
  
  (p1 | p2 | p3) + plot_annotation(title = paste("Effect Size Correlations -", group_name))
}

# Helper to plot heatmap of effect sizes with significance asterisks
plot_heatmap_comparison <- function(df, group_name) {
  sig_df <- df %>%
    filter(FDR_wilcox < 0.25 | FDR_maaslin < 0.25 | FDR_limrots < 0.25)
  
  if (nrow(sig_df) == 0) {
    return(ggplot() + 
             annotate("text", x = 0.5, y = 0.5, label = "No significant features (FDR < 0.25) for heatmap.") +
             theme_void() +
             labs(title = paste("Comparison Heatmap -", group_name)))
  }
  
  # Reshape to long format for ggplot tiles
  df_long <- sig_df %>%
    select(feature, Wilcoxon = logFC_wilcox, MaAsLin3 = coef_maaslin, LimROTS = logFC_limrots) %>%
    pivot_longer(cols = c(Wilcoxon, MaAsLin3, LimROTS), names_to = "Method", values_to = "EffectSize")
  
  df_anno <- sig_df %>%
    select(feature, Wilcoxon = FDR_wilcox, MaAsLin3 = FDR_maaslin, LimROTS = FDR_limrots) %>%
    pivot_longer(cols = c(Wilcoxon, MaAsLin3, LimROTS), names_to = "Method", values_to = "FDR") %>%
    mutate(signif = ifelse(FDR < 0.25, "*", ""))
  
  df_plot <- df_long %>%
    dplyr::inner_join(df_anno, by = c("feature", "Method"))
  
  ggplot(df_plot, aes(x = Method, y = feature, fill = EffectSize)) +
    geom_tile(color = "white", size = 0.5) +
    scale_fill_gradient2(low = "#3182bd", mid = "white", high = "#de2d26", midpoint = 0, name = "Effect Size") +
    geom_text(aes(label = signif), color = "black", size = 6, vjust = 0.7) +
    labs(x = "Method", y = "Taxon", title = paste("Effect Sizes & Significance (* FDR < 0.25) -", group_name)) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(size = 10, face = "italic"),
      plot.title = element_text(size = 14, face = "bold")
    )
}
```

---

## Species-Level Comparison

We load the species altExp and run/retrieve results for each method.

```{r species-run, message=FALSE, warning=FALSE}
# Extract species level tse and compute CLR assay
tse_species <- mia::transformAssay(
  altExp(tse, "species_prevalent"), 
  assay.type = "relabundance",
  method = "clr",
  pseudocount = TRUE
)

species_results <- list()
for (group_name in names(treatment_map)) {
  # 1. Wilcoxon paired
  wilcox_res <- run_wilcoxon_paired(tse_species, "species_prevalent", group_name)
  # 2. MaAsLin3
  maaslin_res <- load_maaslin3_results(treatment_map[[group_name]], "species")
  # 3. LimROTS
  limrots_res <- run_limrots_long(tse, "species_prevalent", group_name)
  
  # Merge by feature
  merged_df <- wilcox_res %>%
    select(feature, logFC_wilcox = logFC, pval_wilcox = p.value, FDR_wilcox = FDR) %>%
    dplyr::inner_join(maaslin_res %>% select(feature, coef_maaslin = coef, FDR_maaslin = FDR), by = "feature") %>%
    dplyr::inner_join(limrots_res %>% select(feature, logFC_limrots = logFC, pval_limrots = p.value, FDR_limrots = FDR), by = "feature")
  
  species_results[[group_name]] <- merged_df
}
```

### Control Group (Species)

```{r species-control, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
df_ctrl <- species_results[["Control"]]
sig_w <- df_ctrl %>% filter(FDR_wilcox < 0.25) %>% pull(feature)
sig_m <- df_ctrl %>% filter(FDR_maaslin < 0.25) %>% pull(feature)
sig_l <- df_ctrl %>% filter(FDR_limrots < 0.25) %>% pull(feature)

p_venn <- plot_venn_overlap(sig_w, sig_m, sig_l, "Control (Species)")
p_cor  <- plot_correlations(df_ctrl, "Control (Species)")
p_heat <- plot_heatmap_comparison(df_ctrl, "Control (Species)")

p_venn
p_cor
p_heat
```

#### Species Concordance Table (Control)

```{r species-control-table}
#| label: tbl-species-concordance-control
datatable(
  df_ctrl %>%
    select(feature, Wilcoxon_logFC = logFC_wilcox, Wilcoxon_FDR = FDR_wilcox,
           Maaslin_coef = coef_maaslin, Maaslin_FDR = FDR_maaslin,
           Limrots_logFC = logFC_limrots, Limrots_FDR = FDR_limrots) %>%
    mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
  extensions = c("Buttons"),
  options = list(
    pageLength = 10,
    dom = 'Bfrtip',
    buttons = c("copy", "csv", "excel")
  ),
  caption = "Master Concordance Table for Control Arm (Species Level)",
  rownames = FALSE
)
```

---

### HWP Group (Species)

```{r species-hwp, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
df_hwp <- species_results[["HWP"]]
sig_w <- df_hwp %>% filter(FDR_wilcox < 0.25) %>% pull(feature)
sig_m <- df_hwp %>% filter(FDR_maaslin < 0.25) %>% pull(feature)
sig_l <- df_hwp %>% filter(FDR_limrots < 0.25) %>% pull(feature)

p_venn <- plot_venn_overlap(sig_w, sig_m, sig_l, "HWP (Species)")
p_cor  <- plot_correlations(df_hwp, "HWP (Species)")
p_heat <- plot_heatmap_comparison(df_hwp, "HWP (Species)")

p_venn
p_cor
p_heat
```

#### Species Concordance Table (HWP)

```{r species-hwp-table}
#| label: tbl-species-concordance-hwp
datatable(
  df_hwp %>%
    select(feature, Wilcoxon_logFC = logFC_wilcox, Wilcoxon_FDR = FDR_wilcox,
           Maaslin_coef = coef_maaslin, Maaslin_FDR = FDR_maaslin,
           Limrots_logFC = logFC_limrots, Limrots_FDR = FDR_limrots) %>%
    mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
  extensions = c("Buttons"),
  options = list(
    pageLength = 10,
    dom = 'Bfrtip',
    buttons = c("copy", "csv", "excel")
  ),
  caption = "Master Concordance Table for HWP Arm (Species Level)",
  rownames = FALSE
)
```

---

### HWP + LGG Group (Species)

```{r species-hwplgg, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
df_hwplgg <- species_results[["HWP + LGG"]]
sig_w <- df_hwplgg %>% filter(FDR_wilcox < 0.25) %>% pull(feature)
sig_m <- df_hwplgg %>% filter(FDR_maaslin < 0.25) %>% pull(feature)
sig_l <- df_hwplgg %>% filter(FDR_limrots < 0.25) %>% pull(feature)

p_venn <- plot_venn_overlap(sig_w, sig_m, sig_l, "HWP + LGG (Species)")
p_cor  <- plot_correlations(df_hwplgg, "HWP + LGG (Species)")
p_heat <- plot_heatmap_comparison(df_hwplgg, "HWP + LGG (Species)")

p_venn
p_cor
p_heat
```

#### Species Concordance Table (HWP + LGG)

```{r species-hwplgg-table}
#| label: tbl-species-concordance-hwplgg
datatable(
  df_hwplgg %>%
    select(feature, Wilcoxon_logFC = logFC_wilcox, Wilcoxon_FDR = FDR_wilcox,
           Maaslin_coef = coef_maaslin, Maaslin_FDR = FDR_maaslin,
           Limrots_logFC = logFC_limrots, Limrots_FDR = FDR_limrots) %>%
    mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
  extensions = c("Buttons"),
  options = list(
    pageLength = 10,
    dom = 'Bfrtip',
    buttons = c("copy", "csv", "excel")
  ),
  caption = "Master Concordance Table for HWP + LGG Arm (Species Level)",
  rownames = FALSE
)
```

---

## Genus-Level Comparison

We load the genus altExp and run/retrieve results for each method.

```{r genus-run, message=FALSE, warning=FALSE}
# Extract genus level tse and compute CLR assay
tse_genus <- mia::transformAssay(
  altExp(tse, "genus_prevalent"), 
  assay.type = "relabundance",
  method = "clr",
  pseudocount = TRUE
)

genus_results <- list()
for (group_name in names(treatment_map)) {
  # 1. Wilcoxon paired
  wilcox_res <- run_wilcoxon_paired(tse_genus, "genus_prevalent", group_name)
  # 2. MaAsLin3
  maaslin_res <- load_maaslin3_results(treatment_map[[group_name]], "genus")
  # 3. LimROTS
  limrots_res <- run_limrots_long(tse, "genus_prevalent", group_name)
  
  # Merge by feature
  merged_df <- wilcox_res %>%
    select(feature, logFC_wilcox = logFC, pval_wilcox = p.value, FDR_wilcox = FDR) %>%
    dplyr::inner_join(maaslin_res %>% select(feature, coef_maaslin = coef, FDR_maaslin = FDR), by = "feature") %>%
    dplyr::inner_join(limrots_res %>% select(feature, logFC_limrots = logFC, pval_limrots = p.value, FDR_limrots = FDR), by = "feature")
  
  genus_results[[group_name]] <- merged_df
}
```

### Control Group (Genus)

```{r genus-control, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
df_ctrl_g <- genus_results[["Control"]]
sig_w <- df_ctrl_g %>% filter(FDR_wilcox < 0.25) %>% pull(feature)
sig_m <- df_ctrl_g %>% filter(FDR_maaslin < 0.25) %>% pull(feature)
sig_l <- df_ctrl_g %>% filter(FDR_limrots < 0.25) %>% pull(feature)

p_venn <- plot_venn_overlap(sig_w, sig_m, sig_l, "Control (Genus)")
p_cor  <- plot_correlations(df_ctrl_g, "Control (Genus)")
p_heat <- plot_heatmap_comparison(df_ctrl_g, "Control (Genus)")

p_venn
p_cor
p_heat
```

#### Genus Concordance Table (Control)

```{r genus-control-table}
#| label: tbl-genus-concordance-control
datatable(
  df_ctrl_g %>%
    select(feature, Wilcoxon_logFC = logFC_wilcox, Wilcoxon_FDR = FDR_wilcox,
           Maaslin_coef = coef_maaslin, Maaslin_FDR = FDR_maaslin,
           Limrots_logFC = logFC_limrots, Limrots_FDR = FDR_limrots) %>%
    mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
  extensions = c("Buttons"),
  options = list(
    pageLength = 10,
    dom = 'Bfrtip',
    buttons = c("copy", "csv", "excel")
  ),
  caption = "Master Concordance Table for Control Arm (Genus Level)",
  rownames = FALSE
)
```

---

### HWP Group (Genus)

```{r genus-hwp, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
df_hwp_g <- genus_results[["HWP"]]
sig_w <- df_hwp_g %>% filter(FDR_wilcox < 0.25) %>% pull(feature)
sig_m <- df_hwp_g %>% filter(FDR_maaslin < 0.25) %>% pull(feature)
sig_l <- df_hwp_g %>% filter(FDR_limrots < 0.25) %>% pull(feature)

p_venn <- plot_venn_overlap(sig_w, sig_m, sig_l, "HWP (Genus)")
p_cor  <- plot_correlations(df_hwp_g, "HWP (Genus)")
p_heat <- plot_heatmap_comparison(df_hwp_g, "HWP (Genus)")

p_venn
p_cor
p_heat
```

#### Genus Concordance Table (HWP)

```{r genus-hwp-table}
#| label: tbl-genus-concordance-hwp
datatable(
  df_hwp_g %>%
    select(feature, Wilcoxon_logFC = logFC_wilcox, Wilcoxon_FDR = FDR_wilcox,
           Maaslin_coef = coef_maaslin, Maaslin_FDR = FDR_maaslin,
           Limrots_logFC = logFC_limrots, Limrots_FDR = FDR_limrots) %>%
    mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
  extensions = c("Buttons"),
  options = list(
    pageLength = 10,
    dom = 'Bfrtip',
    buttons = c("copy", "csv", "excel")
  ),
  caption = "Master Concordance Table for HWP Arm (Genus Level)",
  rownames = FALSE
)
```

---

### HWP + LGG Group (Genus)

```{r genus-hwplgg, fig.width=12, fig.height=6, warning=FALSE, message=FALSE}
df_hwplgg_g <- genus_results[["HWP + LGG"]]
sig_w <- df_hwplgg_g %>% filter(FDR_wilcox < 0.25) %>% pull(feature)
sig_m <- df_hwplgg_g %>% filter(FDR_maaslin < 0.25) %>% pull(feature)
sig_l <- df_hwplgg_g %>% filter(FDR_limrots < 0.25) %>% pull(feature)

p_venn <- plot_venn_overlap(sig_w, sig_m, sig_l, "HWP + LGG (Genus)")
p_cor  <- plot_correlations(df_hwplgg_g, "HWP + LGG (Genus)")
p_heat <- plot_heatmap_comparison(df_hwplgg_g, "HWP + LGG (Genus)")

p_venn
p_cor
p_heat
```

#### Genus Concordance Table (HWP + LGG)

```{r genus-hwplgg-table}
#| label: tbl-genus-concordance-hwplgg
datatable(
  df_hwplgg_g %>%
    select(feature, Wilcoxon_logFC = logFC_wilcox, Wilcoxon_FDR = FDR_wilcox,
           Maaslin_coef = coef_maaslin, Maaslin_FDR = FDR_maaslin,
           Limrots_logFC = logFC_limrots, Limrots_FDR = FDR_limrots) %>%
    mutate(across(where(is.numeric), ~ signif(.x, digits = 3))),
  extensions = c("Buttons"),
  options = list(
    pageLength = 10,
    dom = 'Bfrtip',
    buttons = c("copy", "csv", "excel")
  ),
  caption = "Master Concordance Table for HWP + LGG Arm (Genus Level)",
  rownames = FALSE
)
```

---

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r session-info}
#| label: session-info
sessionInfo()
```
:::
