---
title: "Analysis plan"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
load_result_object <- function() {
  path <- find_first(c(
    "results/metabolomics_results.rds", "../results/metabolomics_results.rds", "../../results/metabolomics_results.rds",
    "output/results/targeted_v2/results.rds", "../output/results/targeted_v2/results.rds", "../../output/results/targeted_v2/results.rds"
  ))
  if (is.na(path)) return(NULL)
  readRDS(path)
}
empty_state <- function(text) cat(text)
plan <- load_plan()
result_obj <- load_result_object()
model_spec <- read_optional_table(c("config/model_specification.csv", "../config/model_specification.csv", "../../config/model_specification.csv"))
model_status <- if (!is.null(result_obj) && "model_status" %in% names(result_obj)) result_obj$model_status else read_optional_table(c("results/model_status.tsv", "../results/model_status.tsv", "../../results/model_status.tsv"))
```


## Objective

Define the primary objective and the scientific estimand.

## Scope

Define included samples, visits, feature panels, exposure families, outcomes, and exclusions.

## Data processing steps

Describe raw data import, identifier alignment, metabolite transformation, scaling, filtering, feature mapping, and derived object creation.

## Covariate strategy

Describe core covariates, measurement-condition covariates, and sensitivity covariates.

## Model plan

```{r}
if (is.null(model_spec)) empty_state("Model specification is pending.") else model_spec
```

## Multiplicity

State the FDR method, correction family, and reporting threshold.

## Deliverables

List expected result objects, model-status logs, tables, figures, supplementary outputs, and rendered report outputs.

## Tools

Record relevant R packages, helper scripts, render command, and session information policy.
