---
title: "Clinical characteristics"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
load_result_object <- function() {
  path <- find_first(c(
    "results/metabolomics_results.rds", "../results/metabolomics_results.rds", "../../results/metabolomics_results.rds",
    "output/results/targeted_v2/results.rds", "../output/results/targeted_v2/results.rds", "../../output/results/targeted_v2/results.rds"
  ))
  if (is.na(path)) return(NULL)
  readRDS(path)
}
empty_state <- function(text) cat(text)
plan <- load_plan()
result_obj <- load_result_object()
model_spec <- read_optional_table(c("config/model_specification.csv", "../config/model_specification.csv", "../../config/model_specification.csv"))
model_status <- if (!is.null(result_obj) && "model_status" %in% names(result_obj)) result_obj$model_status else read_optional_table(c("results/model_status.tsv", "../results/model_status.tsv", "../../results/model_status.tsv"))
```


## Objective-specific context

State which population, visits, groups, exposures, and outcomes define these summaries.

## Sample counts by visit and group

```{r}
summary_tbl <- read_optional_table(c("results/sample_summary.tsv", "../results/sample_summary.tsv", "../../results/sample_summary.tsv", "results/data_summary.tsv", "../results/data_summary.tsv", "../../results/data_summary.tsv"))
if (is.null(summary_tbl)) empty_state("Sample summary is pending.") else summary_tbl
```

## Continuous variables by visit

Summarize continuous clinical, exposure, outcome, and covariate variables by analytic panel.

## Categorical variables by visit

Summarize categorical clinical, exposure, outcome, and covariate variables by analytic panel.

## Baseline characteristics

Include baseline maternal/child or study-entry characteristics when relevant to the analysis plan.

## Requested table of characteristics

Load or generate publication-style characteristic tables when requested.
