---
title: "Covariate and model specification"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
load_result_object <- function() {
  path <- find_first(c(
    "results/metabolomics_results.rds", "../results/metabolomics_results.rds", "../../results/metabolomics_results.rds",
    "output/results/targeted_v2/results.rds", "../output/results/targeted_v2/results.rds", "../../output/results/targeted_v2/results.rds"
  ))
  if (is.na(path)) return(NULL)
  readRDS(path)
}
empty_state <- function(text) cat(text)
plan <- load_plan()
result_obj <- load_result_object()
model_spec <- read_optional_table(c("config/model_specification.csv", "../config/model_specification.csv", "../../config/model_specification.csv"))
model_status <- if (!is.null(result_obj) && "model_status" %in% names(result_obj)) result_obj$model_status else read_optional_table(c("results/model_status.tsv", "../results/model_status.tsv", "../../results/model_status.tsv"))
```


## Analysis question

Restate the exposure, outcome, timing, and target model families.

## Covariate specification

### Covariate selection framework

Document the scientific role of core, measurement-condition, and sensitivity covariates.

### Balance tests or distribution checks

Use this section when intervention arms, groups, or strata require balance or distribution summaries.

### Causal roles

Define exposure, outcome, confounders, mediators, colliders, and precision variables when a causal rationale is used.

### DAG or rationale

Add a DAG or equivalent written rationale only when it is part of the analysis plan.

### Covariate blocks

```{r}
if (is.null(plan) || is.null(plan$variables$covariates)) {
  empty_state("Covariate blocks are pending in the config.")
} else {
  data.frame(block = names(plan$variables$covariates), variables = vapply(plan$variables$covariates, paste, collapse = "; ", FUN.VALUE = character(1)))
}
```

### Collinearity

Report pairwise and model-matrix collinearity checks where available. Preserve thresholds such as VIF limits in source and captions.

### Exposure-covariate screening

Summarize empirical support for sensitivity covariates without using p-values alone to override pre-specified core covariates.

### Covariate-outcome screening

Summarize covariate-outcome screening only as decision support unless the analysis plan assigns it a formal selection role.

### Combined covariate screening summary

Document retained, reviewed, dropped, and sensitivity-only covariates.

### Covariate selection traceability

Every material covariate decision should point to config, model spec, diagnostics, or decision log.

### Missingness

Report missingness and complete-case implications by visit/panel/model family.

## Model specification

### Proposed models

List formulas by model family, exposure timing, outcome panel, and covariate block.

### Mathematical model forms

State the symbolic form of linear, logistic, mixed, mutual-timing, or sensitivity models as applicable.

### Data snapshot

Report sample, feature, and row counts for planned model families.

### Variable key

Define user-facing labels and source variable names.
