---
title: "Primary results"
---

```{r}
#| label: report-context
#| include: false
find_first <- function(paths) {
  hits <- paths[file.exists(paths)]
  if (length(hits)) hits[[1]] else NA_character_
}
load_plan <- function() {
  path <- find_first(c("config/analysis_plan.R", "../config/analysis_plan.R", "../../config/analysis_plan.R"))
  if (is.na(path)) return(NULL)
  env <- new.env(parent = baseenv())
  sys.source(path, envir = env)
  if (exists("analysis_plan", envir = env, inherits = FALSE)) get("analysis_plan", envir = env) else NULL
}
read_optional_table <- function(paths) {
  path <- find_first(paths)
  if (is.na(path)) return(NULL)
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("tsv", "txt")) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE) else read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}
load_result_object <- function() {
  path <- find_first(c(
    "results/metabolomics_results.rds", "../results/metabolomics_results.rds", "../../results/metabolomics_results.rds",
    "output/results/targeted_v2/results.rds", "../output/results/targeted_v2/results.rds", "../../output/results/targeted_v2/results.rds"
  ))
  if (is.na(path)) return(NULL)
  readRDS(path)
}
empty_state <- function(text) cat(text)
plan <- load_plan()
result_obj <- load_result_object()
model_spec <- read_optional_table(c("config/model_specification.csv", "../config/model_specification.csv", "../../config/model_specification.csv"))
model_status <- if (!is.null(result_obj) && "model_status" %in% names(result_obj)) result_obj$model_status else read_optional_table(c("results/model_status.tsv", "../results/model_status.tsv", "../../results/model_status.tsv"))
```


## Signal summary

```{r}
primary_tbl <- read_optional_table(c("results/primary_results.tsv", "../results/primary_results.tsv", "../../results/primary_results.tsv"))
if (is.null(primary_tbl)) empty_state("Primary results are pending.") else head(primary_tbl, 50)
```

## Question-specific result sections

Create one section per model family in the model specification. Preserve exposure-family subsections when the analysis separates nutrients, continuous scores, and categorical classes.

## Concurrent or age-specific associations

Report feature-wise models by visit or exposure window when present.

## Prospective associations

Report earlier-exposure/later-panel models when present.

## Mutual-timing or repeated-measures associations

Report paired timing terms, main effects, and interaction terms separately when they answer different questions.

## Top associations

Show top-ranked associations only after clarifying whether ranking is primary, descriptive, or post hoc.

## Sensitivity summary

Compare primary and sensitivity estimates using stable keys. Report estimate shifts, sign changes, and FDR-threshold reclassification.

## Execution status

```{r}
if (is.null(model_status)) empty_state("Model-status log is pending.") else model_status
```
