---
title: "Study overview"
---

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")

analysis_data <- extract_analysis_data("../data/MAE_original.rds")
mae <- analysis_data$mae
df_long <- analysis_data$data
metabolite_features_count <- length(analysis_data$metabolite_cols)

visit_label <- function(v) {
  dplyr::case_when(
    v == 4 ~ "6 Months",
    v == 5 ~ "1 Year",
    v == 6 ~ "2 Years",
    v == 7 ~ "5-6 Years (Follow-up)",
    TRUE ~ paste0("Visit ", v)
  )
}
```

## Research aim

The objective of this study is to investigate the associations between early-life circulating metabolomic profiles and neurocognitive function in childhood. 

Specifically, the study aims to:

1. Examine if circulating serum metabolites at **6 months of age** (`visit_4`) predict neurocognitive development at **2 years of age**.
2. Examine if circulating serum metabolites at **1 year of age** (`visit_5`) predict neurocognitive development at **2 years of age**.
3. Evaluate concurrent associations between serum metabolites and neurocognitive development at **2 years of age** (`visit_6`).

```{r}
#| label: research-questions-table

question_tbl <- tibble::tibble(
  Question = c("Q1", "Q2", "Q3"),
  Analysis = c("Predictive 6-month metabolites", "Predictive 1-year metabolites", "Concurrent 2-year metabolites"),
  `Exposure panel` = c("visit_4 serum NMR metabolites", "visit_5 serum NMR metabolites", "visit_6 serum NMR metabolites"),
  `Outcome panel` = rep("2-year Bayley-III and HINE outcomes", 3),
  Role = c("Primary", "Primary", "Secondary")
)

make_table(question_tbl, caption = "Planned analysis questions", table_mode = "compact")
```

---

## Visit mapping and analytic panels

Visit mapping and matched metabolomics sample counts are pulled from the `MultiAssayExperiment` dataset.

```{r}
#| label: analytic-panels-table

panel_tbl <- df_long %>%
  distinct(StudyID, visit_num) %>%
  count(visit_num, name = "n_children") %>%
  mutate(
    mae_visit = paste0("visit_", visit_num),
    child_age = visit_label(visit_num),
    data_collected = dplyr::case_when(
      visit_num %in% c(4L, 5L) ~ "Serum NMR metabolites",
      visit_num == 6L ~ "Serum NMR metabolites; Bayley-III; HINE",
      visit_num == 7L ~ "Later follow-up panel; excluded from primary analysis",
      TRUE ~ "Serum NMR metabolites"
    ),
    metabolite_features = metabolite_features_count
  ) %>%
  transmute(
    Visit = mae_visit,
    Age = child_age,
    `Data collected` = data_collected,
    `Children, n` = n_children,
    `Metabolite features, n` = metabolite_features
  )

make_table(panel_tbl, caption = "Analytic child-age panels from MAE", table_mode = "compact")
```

---

## Clinical outcomes availability

Available neurocognitive outcomes are summarized below for both the full unmatched clinical cohort and the matched analysis samples (Visits 4, 5, and 6).

### Unmatched clinical cohort

The table below summarizes the total available neurocognitive measurements in the raw clinical database ($N = 439$), regardless of metabolomics sample matching.

```{r}
#| label: raw-clinical-outcomes-summary

raw_clinical <- haven::read_sav("../data/FOPP_clinical_variables_child_cognition_20260617.sav")

# Derive HINE_optimal following the study protocol (preterm excluded)
prem_raw <- as.numeric(raw_clinical$Premature)
ham_raw <- as.numeric(raw_clinical$Hammersmith6)
raw_is_premature <- !is.na(prem_raw) & prem_raw == 2
raw_clinical$HINE_optimal <- ifelse(raw_is_premature, NA_character_,
                                    ifelse(ham_raw >= 74, "Optimal", "Suboptimal"))

outcome_vars <- c(
  CCognition_indexpoints6 = "Bayley composite cognitive",
  CLanguage_indexpoints6 = "Bayley composite language",
  CExpressivelanguage_standardpoints6 = "Bayley expressive language",
  CReceptivelanguage_standardpoints6 = "Bayley receptive language",
  CMotor_indexpoints6 = "Bayley composite motor",
  CFinemotor_standardpoints6 = "Bayley fine motor",
  CGrossmotor_standardpoints6 = "Bayley gross motor",
  Hammersmith6 = "HINE global score",
  HINE_optimal = "HINE optimal/suboptimal, preterm excluded"
)

raw_outcomes_tbl <- tibble(
  Variable = names(outcome_vars),
  Outcome = unname(outcome_vars),
  Family = ifelse(Variable == "HINE_optimal", "Binary", "Continuous"),
  `Available, n` = vapply(Variable, function(v) sum(!is.na(raw_clinical[[v]])), integer(1)),
  `Missing, n` = vapply(Variable, function(v) sum(is.na(raw_clinical[[v]])), integer(1))
)

make_table(raw_outcomes_tbl, caption = "Availability of 2-year neurocognitive outcome measurements in raw clinical database (unmatched)", table_mode = "compact")
```

### Matched analysis samples (Visits 4, 5, and 6)

The table below summarizes the available measurements for the subsets of children who have both clinical outcomes and matched metabolomics data at Visit 4 (6 months, $N = 258$), Visit 5 (1 year, $N = 257$), or Visit 6 (2 years, $N = 239$).

```{r}
#| label: clinical-outcomes-summary-all-visits

v4_tse <- experiments(mae)[["visit_4"]]
v4_clinical <- as.data.frame(colData(v4_tse))

v5_tse <- experiments(mae)[["visit_5"]]
v5_clinical <- as.data.frame(colData(v5_tse))

v6_tse <- experiments(mae)[["visit_6"]]
v6_clinical <- as.data.frame(colData(v6_tse))

outcomes_tbl_all <- tibble(
  Variable = names(outcome_vars),
  Outcome = unname(outcome_vars),
  Family = ifelse(Variable == "HINE_optimal", "Binary", "Continuous"),
  `Visit 4 Available (6mo), n` = vapply(Variable, function(v) sum(!is.na(v4_clinical[[v]])), integer(1)),
  `Visit 5 Available (1y), n` = vapply(Variable, function(v) sum(!is.na(v5_clinical[[v]])), integer(1)),
  `Visit 6 Available (2y), n` = vapply(Variable, function(v) sum(!is.na(v6_clinical[[v]])), integer(1))
)

make_table(outcomes_tbl_all, caption = "Availability of 2-year neurocognitive outcome measurements in matched metabolomics samples across child visits", table_mode = "compact")
```

## Session info

::: {.callout-note collapse="true" title="Show session information"}
```{r}
#| label: session-info
sessionInfo()
```
:::


