---
title: "Study overview"
---

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")
library(dplyr)
library(tidyr)

results <- readRDS("../output/results/targeted/prenatal_diet_study2_results.rds")
model_status <- results$model_status

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = paste0("visit_", 4:7)
)

raw_df <- extract_long_from_mae(
  analysis_obj$mae,
  visits = paste0("visit_", 4:7)
)
```

## Research aim

Maternal prenatal diet exposures measured in early and late pregnancy are evaluated against the child plasma metabolome across infancy and childhood.

Questions:

1. Q1: early-pregnancy diet associations across child ages
2. Q2: late-pregnancy diet associations across child ages
3. Q3: mutual timing models comparing early and late pregnancy in the same regression
4. Q4: longitudinal diet-by-age interaction models across childhood

## Analytic panels

```{r}
#| label: study-overview-01
panel_tbl <- raw_df %>%
  distinct(StudyID, visit_num) %>%
  count(visit_num, name = "n_children") %>%
  mutate(
    child_age = visit_label(visit_num),
    metabolite_features = length(analysis_obj$metabolite_cols)
  ) %>%
  select(child_age, n_children, metabolite_features)

make_table(panel_tbl, caption = "Analytic child-age panels", table_mode = "compact")
```

## Exposure timing and families

```{r}
#| label: study-overview-02
q4_fams <- unique(model_status$exposure[model_status$question == "Q4"])
q5_fams <- unique(model_status$exposure[model_status$question == "Q5"])

exposure_tbl <- tibble(
  component = c(
    "Q1: Early-Pregnancy Age-Specific Models",
    "Q2: Late-Pregnancy Age-Specific Models",
    "Q3: Mutual Timing Model",
    "Q4: Longitudinal Interaction Model"
  ),
  exposure_timing = c(
    "Early pregnancy",
    "Late pregnancy",
    "Early and late pregnancy entered together",
    "Early or late pregnancy exposure carried across repeated child ages"
  ),
  exposure_families = c(
    paste(pretty_var(q4_fams), collapse = "; "),
    paste(pretty_var(q5_fams), collapse = "; "),
    paste(pretty_var(unique(model_status$exposure[model_status$question == "Q6"])), collapse = "; "),
    paste(pretty_var(unique(model_status$exposure[model_status$question == "Q7"])), collapse = "; ")
  )
)

make_table(exposure_tbl, caption = "Exposure timing and families", table_mode = "wide")
```

## Covariate strategy

```{r}
#| label: study-overview-03
covar_tbl <- tibble(
  analysis_component = c(
    "Q1/Q2: Age-Specific Models",
    "Q3: Mutual Timing Model",
    "Q4: Longitudinal Interaction Model"
  ),
  primary_covariates = c(
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time",
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time",
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time"
  ),
  sensitivity_additions = c(
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight",
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight",
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight"
  )
)

make_table(covar_tbl, caption = "Primary and sensitivity covariate strategy", table_mode = "wide")
```

## Model families

```{r}
#| label: study-overview-04
model_tbl <- tibble(
  component = c(
    "Q1: Early-Pregnancy Age-Specific Models",
    "Q2: Late-Pregnancy Age-Specific Models",
    "Q3: Mutual Timing Model",
    "Q4: Longitudinal Interaction Model"
  ),
  model_family = c(
    "Feature-wise linear model at each child age",
    "Feature-wise linear model at each child age",
    "Feature-wise linear model with early and late pregnancy terms together",
    "Feature-wise mixed model with diet-by-age interaction"
  ),
  outcome = c(
    "Child metabolite at 6 months, 12 months, 24 months, and 5-6 years",
    "Child metabolite at 6 months, 12 months, 24 months, and 5-6 years",
    "Child metabolite at 6 months, 12 months, 24 months, and 5-6 years",
    "Repeated child metabolite across all available ages"
  )
)

make_table(model_tbl, caption = "Primary model families", table_mode = "wide")
```

## Multiple testing

False-discovery control is based on Benjamini-Hochberg adjustment across metabolites, but the adjustment scope differs by model family.

- Age-specific outputs apply BH within each visit-specific exposure model.
- Mutual timing outputs apply BH separately to the early-term and late-term coefficients within each visit-specific exposure pair.
- Longitudinal outputs apply BH separately within each exposure and coefficient family.

## Result inventory

```{r}
#| label: study-overview-05
inventory_tbl <- build_result_inventory_prenatal_diet(results)

make_table(inventory_tbl, caption = "Returned result inventory", table_mode = "compact")
```

## Execution status

```{r}
#| label: study-overview-06
status_tbl <- model_status %>%
  mutate(
    question = dplyr::case_when(
      question == "Q4" ~ "Q1: Early-Pregnancy Age-Specific Models",
      question == "Q5" ~ "Q2: Late-Pregnancy Age-Specific Models",
      question == "Q6" ~ "Q3: Mutual Timing Model",
      question == "Q7" ~ "Q4: Longitudinal Interaction Model",
      TRUE ~ question
    ),
    child_age = ifelse(is.na(visit), NA_character_, visit_label(as.integer(visit))),
    exposure = pretty_var(exposure),
    n_complete = ifelse(
      is.na(n_complete_min) | !is.finite(n_complete_min),
      NA_character_,
      ifelse(n_complete_min == n_complete_max, as.character(n_complete_min), paste0(n_complete_min, "-", n_complete_max))
    )
  ) %>%
  select(question, analysis, child_age, timing, exposure, status, reason, n_complete, n_features_tested, term_family)

make_table(status_tbl, caption = "Saved execution-status inventory", table_mode = "wide")
```

## Interpretation notes

1. The age-specific and mutual-timing models carry the clearest inferential weight in the current Study 2 package.
2. Prenatal exposures are baseline pregnancy-window variables; child-age rows represent linked analytic panels, not repeated prenatal measurements.
3. The longitudinal interaction model is part of the target question set and now returns coefficient-level output across all attempted exposures, but it should still be interpreted more cautiously than the age-specific models because the fits generated substantial warnings.
4. Detailed coefficient tables, sensitivity comparisons, and supporting figures are reported elsewhere in the site; this page is the study-level map.
