---
title: "Supplementary tables"
---

```{r}
#| label: setup
#| include: false
source("analysis_data_utils.R")
source("report_utils.R")
library(dplyr)
library(tidyr)

results <- readRDS("../output/results/targeted/prenatal_diet_study2_results.rds")
model_status <- results$model_status

analysis_obj <- prepare_analysis_data(
  mae_path = "../data/MAE_original.rds",
  visits = paste0("visit_", 4:7)
)

raw_df <- extract_long_from_mae(
  analysis_obj$mae,
  visits = paste0("visit_", 4:7)
)

age_res <- results$age_specific
long_res <- results$longitudinal
age_res_sens <- results$age_specific_sensitivity
long_res_sens <- results$longitudinal_sensitivity

has_any_nonmissing <- function(df, cols) {
  cols <- intersect(cols, names(df))
  if (length(cols) == 0) return(rep(FALSE, nrow(df)))
  rowSums(!is.na(df[, cols, drop = FALSE])) > 0
}

early_all <- flatten_age_specific(age_res, "early")
late_all <- flatten_age_specific(age_res, "late")
mutual_early_all <- flatten_mutual(age_res, "early")
mutual_late_all <- flatten_mutual(age_res, "late")

early_all_sens <- if (!is.null(age_res_sens)) flatten_age_specific(age_res_sens, "early") else data.frame()
late_all_sens <- if (!is.null(age_res_sens)) flatten_age_specific(age_res_sens, "late") else data.frame()
mutual_early_all_sens <- if (!is.null(age_res_sens)) flatten_mutual(age_res_sens, "early") else data.frame()
mutual_late_all_sens <- if (!is.null(age_res_sens)) flatten_mutual(age_res_sens, "late") else data.frame()

q4_expos <- sort(unique(model_status$exposure[model_status$question == "Q4"]))
q5_expos <- sort(unique(model_status$exposure[model_status$question == "Q5"]))

raw_df$has_early_exposure <- has_any_nonmissing(raw_df, q4_expos)
raw_df$has_late_exposure <- has_any_nonmissing(raw_df, q5_expos)

child_panel_counts <- raw_df %>%
  distinct(StudyID, visit_num) %>%
  count(StudyID, name = "n_panels")
```

## Supplementary Table S1. Analytic sample derivation

```{r}
#| label: supplementary-tables-01
sample_tbl <- tibble(
  analytic_step = c(
    "Children with at least one linked metabolomics panel",
    "Children with a 6-month metabolomics panel",
    "Children with a 12-month metabolomics panel",
    "Children with a 24-month metabolomics panel",
    "Children with a 5-6-year metabolomics panel",
    "Children with any linked early-pregnancy exposure",
    "Children with any linked late-pregnancy exposure",
    "Children with both early- and late-pregnancy exposures in linked panels",
    "Children contributing at least two child-age panels"
  ),
  n_children = c(
    n_distinct(raw_df$StudyID),
    n_distinct(raw_df$StudyID[raw_df$visit_num == 4]),
    n_distinct(raw_df$StudyID[raw_df$visit_num == 5]),
    n_distinct(raw_df$StudyID[raw_df$visit_num == 6]),
    n_distinct(raw_df$StudyID[raw_df$visit_num == 7]),
    n_distinct(raw_df$StudyID[raw_df$has_early_exposure]),
    n_distinct(raw_df$StudyID[raw_df$has_late_exposure]),
    n_distinct(raw_df$StudyID[raw_df$has_early_exposure & raw_df$has_late_exposure]),
    sum(child_panel_counts$n_panels >= 2, na.rm = TRUE)
  )
)

make_table(sample_tbl, caption = "Supplementary Table S1. Analytic sample derivation")
```

## Supplementary Table S2. Exposure definitions

```{r}
#| label: supplementary-tables-02
exposure_tbl <- bind_rows(
  tibble(
    component = "Q1: Early-Pregnancy Age-Specific Models",
    exposure_timing = "Early pregnancy",
    exposure_code = q4_expos
  ),
  tibble(
    component = "Q2: Late-Pregnancy Age-Specific Models",
    exposure_timing = "Late pregnancy",
    exposure_code = q5_expos
  )
) %>%
  mutate(
    exposure_label = pretty_var(exposure_code),
    scale = case_when(
      grepl("DietaryPatterns", exposure_code) ~ "Diet pattern score",
      grepl("DII", exposure_code) ~ "Diet inflammatory index score",
      grepl("IDQ", exposure_code) ~ "Diet quality score",
      TRUE ~ "Continuous percentage or intake metric"
    ),
    linked_use = case_when(
      grepl("1$", exposure_code) ~ "Used in early-pregnancy age-specific, mutual-timing, and early longitudinal attempts",
      grepl("2$", exposure_code) ~ "Used in late-pregnancy age-specific, mutual-timing, and late longitudinal attempts",
      TRUE ~ "Used in prenatal diet modeling"
    )
  ) %>%
  relocate(exposure_label, .after = exposure_code)

make_table(exposure_tbl, caption = "Supplementary Table S2. Exposure definitions and coding")
```

## Supplementary Table S3. Primary model specifications

```{r}
#| label: supplementary-tables-03
model_tbl <- tibble(
  analysis_component = c(
    "Q1: Early-Pregnancy Age-Specific Models",
    "Q2: Late-Pregnancy Age-Specific Models",
    "Q3: Mutual Timing Model",
    "Q4: Longitudinal Interaction Model"
  ),
  outcome = c(
    "Child metabolite at each age",
    "Child metabolite at each age",
    "Child metabolite at each age with early and late pregnancy terms together",
    "Repeated child metabolite across all available ages"
  ),
  model_family = c(
    "Feature-wise linear model",
    "Feature-wise linear model",
    "Feature-wise linear model",
    "Feature-wise mixed model"
  ),
  primary_covariates = c(
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time",
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time",
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time",
    "Intervention; child sex/gender; maternal ethnicity; maternal disease; visit-specific fasting time"
  ),
  sensitivity_additions = c(
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight",
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight",
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight",
    "Pre-pregnancy BMI; maternal age; prior GDM; breastfeeding duration; maternal smoking; university education; child birth weight"
  ),
  multiple_testing = c(
    "Benjamini-Hochberg across metabolites within each visit-specific exposure model",
    "Benjamini-Hochberg across metabolites within each visit-specific exposure model",
    "Benjamini-Hochberg across metabolites applied separately to the early-term and late-term coefficients within each visit-specific exposure pair",
    "Benjamini-Hochberg across metabolites applied separately within each longitudinal exposure and coefficient family"
  )
)

make_table(model_tbl, caption = "Supplementary Table S3. Primary model specifications")
```

## Supplementary Table S4. Returned result inventory

```{r}
#| label: supplementary-tables-04
inventory_tbl <- build_result_inventory_prenatal_diet(results)

make_table(inventory_tbl, caption = "Supplementary Table S4. Returned result inventory")
```

## Supplementary Table S5. Candidate associations with q < 0.10

```{r}
#| label: supplementary-tables-05
candidate_age_specific <- bind_rows(
  early_all %>% mutate(component = paste0("Q1 early pregnancy | ", visit_label(visit))),
  late_all %>% mutate(component = paste0("Q2 late pregnancy | ", visit_label(visit)))
) %>%
  filter(q.value < 0.10) %>%
  transmute(
    component,
    exposure = pretty_var(exposure),
    feature,
    n,
    beta = estimate,
    conf_low = conf.low,
    conf_high = conf.high,
    p_value = p.value,
    q_value = q.value
  )

candidate_mutual <- bind_rows(
  mutual_early_all %>% mutate(component = paste0("Q3 mutual timing early term | ", visit_label(visit))),
  mutual_late_all %>% mutate(component = paste0("Q3 mutual timing late term | ", visit_label(visit)))
) %>%
  filter(q.value < 0.10) %>%
  transmute(
    component,
    exposure = pretty_var(exposure_pair),
    feature,
    n,
    beta = estimate,
    conf_low = conf.low,
    conf_high = conf.high,
    p_value = p.value,
    q_value = q.value
  )

candidate_tbl <- bind_rows(candidate_age_specific, candidate_mutual) %>%
  arrange(q_value, p_value)

make_table(candidate_tbl, caption = "Supplementary Table S5. Candidate associations meeting q < 0.10")
```

## Supplementary Table S6. Primary-vs-sensitivity stability summary

```{r}
#| label: supplementary-tables-06
long_coef_primary <- long_res %>% filter("feature" %in% names(long_res))
long_coef_sens <- long_res_sens %>% filter("feature" %in% names(long_res_sens))

stability_tbl <- bind_rows(
  compare_primary_sensitivity(early_all, early_all_sens, c("exposure", "feature", "visit")) %>% mutate(component = "Q1: Early-Pregnancy Age-Specific Models"),
  compare_primary_sensitivity(late_all, late_all_sens, c("exposure", "feature", "visit")) %>% mutate(component = "Q2: Late-Pregnancy Age-Specific Models"),
  compare_primary_sensitivity(mutual_early_all, mutual_early_all_sens, c("exposure_pair", "feature", "visit")) %>% mutate(component = "Q3: Mutual Timing: Early-Pregnancy Term"),
  compare_primary_sensitivity(mutual_late_all, mutual_late_all_sens, c("exposure_pair", "feature", "visit")) %>% mutate(component = "Q3: Mutual Timing: Late-Pregnancy Term"),
  compare_primary_sensitivity(long_coef_primary, long_coef_sens, c("exposure", "feature", "term")) %>% mutate(component = "Q4: Longitudinal Interaction Model")
) %>%
  group_by(component) %>%
  summarise(
    matched_rows = n(),
    median_abs_beta_change = median(abs_delta, na.rm = TRUE),
    p90_abs_beta_change = quantile(abs_delta, 0.90, na.rm = TRUE),
    direction_flip_fraction = mean(direction_flip, na.rm = TRUE),
    q_reclassification_fraction = mean(q_reclassified, na.rm = TRUE),
    .groups = "drop"
  )

make_table(stability_tbl, caption = "Supplementary Table S6. Primary-vs-sensitivity stability summary")
```

## Supplementary Table S7. Longitudinal execution status inventory

```{r}
#| label: supplementary-tables-07
status_tbl <- model_status %>%
  mutate(
    component = case_when(
      question == "Q4" ~ "Q1: Early-Pregnancy Age-Specific Models",
      question == "Q5" ~ "Q2: Late-Pregnancy Age-Specific Models",
      question == "Q6" ~ "Q3: Mutual Timing Model",
      question == "Q7" ~ "Q4: Longitudinal Interaction Model",
      TRUE ~ question
    ),
    child_age = ifelse(is.na(visit), NA_character_, visit_label(as.integer(visit))),
    exposure = pretty_var(exposure),
    complete_case_n = ifelse(
      is.na(n_complete_min) | !is.finite(n_complete_min),
      NA_character_,
      ifelse(n_complete_min == n_complete_max, as.character(n_complete_min), paste0(n_complete_min, "-", n_complete_max))
    )
  ) %>%
  select(component, analysis, child_age, timing, exposure, term_family, status, reason, complete_case_n, n_features_tested)

make_table(status_tbl, caption = "Supplementary Table S7. Saved execution-status inventory")
```
